---
title: "R Basics"
---

# R as a Calculator
```{r}
1 + 1
3 * (4 + 5)
```

# Objects & Assignment
```{r}
x <- 10
y <- 3.5
x + y
```

# Object Naming Rules

- R variable names can contain letters, numbers, periods, and underscores.  However, they cannot start with a number or underscore.  R is case-sensitive, so `age`, `Age`, and `AGE` would be considered three different variables.
- R variable names should be descriptive and meaningful.  Avoid using reserved words or function names as variable names.
- A variable can have a short name (like x and y) or a more descriptive name (age, carname, total_volume). Rules for R variables are: 
- A variable name must start with a letter and can be a combination of letters, digits, period(.) and underscore(_). If it starts with period(.), it cannot be followed by a digit.
- A variable name cannot start with a number or underscore (_). Variable names are case-sensitive (age, Age and AGE are three different variables). Reserved words cannot be used as variables (TRUE, FALSE, NULL, etc.).
- Variable names should not contain spaces. Use underscore (_) or period (.) to separate words in a variable name.
- Variable names should be meaningful and descriptive. Avoid using single-letter variable names except for temporary variables in loops or functions.

# Basic Operations in R

R supports various basic operations, including: 

- **Arithmetic Operations**: Addition (`+`), subtraction (`-`), multiplication (`*`), division (`/`), and exponentiation (`^`). Example: 

```{r}
a <- 10
b <- 5
my_sum <- a + b
diff <- a - b
prod <- a * b
quot <- a / b
power <- a ^ b
my_sum; diff; prod; quot; power
```

- **Comparison Operations**: Equal to (`==`), not equal to (`!=`), greater than (`>`), less than (`<`), greater than or equal to (`>=`), and less than or equal to (`<=`). Example:

```{r}
x <- 10
y <- 5
eq <- x == y
neq <- x != y
gt <- x > y
lt <- x < y
gte <- x >= y
lte <- x <= y
eq; neq; gt; lt; gte; lte
```

- **Logical Operations**: AND (`&`), OR (`|`), and NOT (`!`). Example:

```{r}
p <- TRUE
q <- FALSE
and_result <- p & q
or_result <- p | q
not_result <- !p
and_result; or_result; not_result
```

# Comments in R

Comments in R are created using the `#` symbol. Anything following the `#` on the same line is considered a comment and is ignored by R during execution. Example:

```{r}
# This is a comment
x <- 10  # Assigning value to x
y <- 5   # Assigning value to y
total <- x + y  # Calculating the sum of x and y
total  # Output the sum
```

# Getting Help in R

R provides several ways to get help and documentation for functions and packages:

- `?function_name`: Displays the documentation for a specific function. Example:

```{r}
?mean
```

- `help(function_name)`: Another way to access the documentation for a function. Example:

```{r}
help(mean)
```

- `help.search("keyword")`: Searches for help topics related to a specific keyword. Example:

```{r}
help.search("regression")
```

- `example(function_name)`: Shows examples of how to use a specific function.  Example:

```{r}
example(mean)
```

- `vignette("package_name")`: Opens the vignette (detailed documentation) for a specific package.  Example:

```{r}
vignette("dplyr")
```

- `??keyword`: Searches for help topics related to a specific keyword (similar to `help.search`). Example:

```{r}
??regression
```

# Installing and Loading Packages in R

R has a vast ecosystem of packages that extend its functionality. To use a package, you need to install it first and then load it into your R session. 

- **Installing a Package**: Use the `install.packages("package_name")` function to install a package from CRAN.  Example:

```{r eval=FALSE}
install.packages("ggplot2")
```

- **Loading a Package**: Use the `library(package_name)` function to load an installed package into your R session. Example:

```{r}
library(ggplot2)

# Now you can use functions from the ggplot2 package
```

# Saving and Loading Workspaces in R

You can save your R workspace (all objects in memory) to a file and load it later: 

- **Saving Workspace**:  Use the `save.image("file_name.RData")` function to save the entire workspace to a file. Example:

```{r eval=FALSE}
save.image("my_workspace.RData")
```

- **Loading Workspace**: Use the `load("file_name.RData")` function to load a saved workspace from a file. Example:

```{r eval=FALSE}
load("my_workspace. RData")
```

# Working Directory
```{r}
getwd()
# setwd("/path/you/want") # avoid in reproducible code; prefer here::here() for projects
```

# Vectors (Atomic)
```{r}
nums <- c(1, 2, 3, 4)
chars <- c("a", "b", "c")
logical <- c(TRUE, FALSE, TRUE)
typeof(nums); typeof(chars); typeof(logical)
```

# Exercises
1. Create an object `z` that stores `(5^2 + 7)/3`.
2. Use `?seq` and create a sequence from 0 to 1 by 0.1.
3. Inspect `typeof()` for a few objects you create.