---
title: "Base R Functions & Apply Family"
---

# Introduction

Base R provides a rich set of built-in functions for data manipulation, statistical calculations, and programming. The apply family of functions is particularly powerful for applying operations across different dimensions of data structures.  This chapter covers essential base R functions and the apply family in detail.

# Common Utility Functions

Base R includes numerous utility functions for statistical calculations, data generation, and manipulation. 

## Statistical Functions

```{r}
# Create sample data
x <- 1:10
print(x)

# Basic statistical functions
sum(x)        # Sum of all elements
mean(x)       # Arithmetic mean
median(x)     # Median value
sd(x)         # Standard deviation
var(x)        # Variance
min(x)        # Minimum value
max(x)        # Maximum value
range(x)      # Returns min and max
length(x)     # Number of elements
```

## Quantile Function

The `quantile()` function calculates sample quantiles corresponding to given probabilities: 

```{r}
x <- 1:10
quantile(x)                    # Default quartiles (0%, 25%, 50%, 75%, 100%)
quantile(x, probs = 0.9)       # 90th percentile
quantile(x, probs = c(0.1, 0.5, 0.9))  # Custom percentiles
```

## Sequence and Repetition Functions

### seq() - Generate Sequences

The `seq()` function creates sequences of numbers:

```{r}
# Different ways to create sequences
seq(0, 1, by = 0.1)           # From 0 to 1, increment by 0.1
seq(0, 1, length.out = 11)    # From 0 to 1, 11 equally spaced values
seq(from = 5, to = 50, by = 5) # From 5 to 50, increment by 5
seq_along(letters[1:5])        # Sequence along the length of an object
```

### rep() - Repeat Values

The `rep()` function repeats values:

```{r}
# Different repetition patterns
rep(5, times = 3)              # Repeat 5 three times
rep(c(1, 2), times = 3)        # Repeat vector three times
rep(c(1, 2), each = 3)         # Repeat each element three times
rep(1:3, length.out = 10)      # Repeat to reach specified length
```

# The Apply Family of Functions

The apply family provides efficient ways to apply functions across different dimensions of data structures. These functions are alternatives to writing explicit loops.

## apply() - For Matrices and Arrays

The `apply()` function applies a function over the margins of an array or matrix. 

**Syntax**:  `apply(X, MARGIN, FUN, ... )`
- `X`: Array or matrix
- `MARGIN`: 1 = rows, 2 = columns, c(1,2) = both
- `FUN`: Function to apply
- `...`: Additional arguments to FUN

```{r}
# Create a sample matrix
m <- matrix(1:12, nrow = 3, ncol = 4)
print(m)

# Apply functions across rows (MARGIN = 1)
apply(m, 1, sum)      # Row sums
apply(m, 1, mean)     # Row means
apply(m, 1, max)      # Row maxima

# Apply functions across columns (MARGIN = 2)
apply(m, 2, sum)      # Column sums
apply(m, 2, mean)     # Column means
apply(m, 2, sd)       # Column standard deviations

# Using built-in optimized functions (when available)
rowSums(m)            # Equivalent to apply(m, 1, sum) but faster
colSums(m)            # Equivalent to apply(m, 2, sum) but faster
rowMeans(m)           # Equivalent to apply(m, 1, mean) but faster
colMeans(m)           # Equivalent to apply(m, 2, mean) but faster
```

## lapply() - For Lists and Vectors

The `lapply()` function applies a function to each element of a list or vector and returns a list. 

**Syntax**: `lapply(X, FUN, ...)`

```{r}
# Create sample data
lst <- list(a = 1:5, b = 6:10, c = 11:15)
print(lst)

# Apply function to each list element
lapply(lst, sum)      # Sum of each element
lapply(lst, mean)     # Mean of each element
lapply(lst, length)   # Length of each element

# Using lapply with custom functions
lapply(lst, function(x) x^2)              # Square each element
lapply(lst, function(x) x[x > 8])         # Filter elements > 8
```

## sapply() - Simplified lapply()

The `sapply()` function is similar to `lapply()` but attempts to simplify the result to a vector or matrix when possible.

```{r}
# Same data as above
lst <- list(a = 1:5, b = 6:10, c = 11:15)

# sapply returns simplified results
sapply(lst, sum)      # Returns named vector instead of list
sapply(lst, mean)     # Returns named vector
sapply(lst, range)    # Returns matrix when each result has same length

# Compare lapply vs sapply
result_lapply <- lapply(lst, mean)
result_sapply <- sapply(lst, mean)
class(result_lapply)  # List
class(result_sapply)  # Named numeric vector
```

## vapply() - Type-Safe sapply()

The `vapply()` function is like `sapply()` but with explicit specification of the return type, making it safer and faster.

**Syntax**: `vapply(X, FUN, FUN.VALUE, ...)`

```{r}
# vapply with explicit return type specification
lst <- list(a = 1:5, b = 6:10, c = 11:15)

# Specify that each function call returns a single numeric value
vapply(lst, mean, FUN.VALUE = numeric(1))

# Specify that each function call returns two numeric values
vapply(lst, range, FUN.VALUE = numeric(2))

# This would throw an error if the function doesn't return the expected type
# vapply(lst, mean, FUN.VALUE = character(1))  # Error!
```

## mapply() - Multivariate Apply

The `mapply()` function applies a function to multiple arguments element-wise.

**Syntax**:  `mapply(FUN, ..., MoreArgs = NULL, SIMPLIFY = TRUE)`

```{r}
# Apply function to multiple vectors element-wise
vec1 <- 1:3
vec2 <- 10:12
vec3 <- 100:102

# Add corresponding elements from two vectors
mapply(sum, vec1, vec2)
# Equivalent to:  c(sum(1, 10), sum(2, 11), sum(3, 12))

# Add corresponding elements from three vectors
mapply(sum, vec1, vec2, vec3)

# Using mapply with custom functions
mapply(function(x, y) x^y, vec1, c(2, 3, 4))

# mapply with lists
list1 <- list(a = 1:3, b = 4:6)
list2 <- list(x = 7:9, y = 10:12)
mapply(function(a, b) sum(a) + sum(b), list1, list2)
```

## tapply() - Apply by Groups

The `tapply()` function applies a function to subsets of a vector based on grouping factors.

**Syntax**:  `tapply(X, INDEX, FUN, ...)`

```{r}
# Sample data with grouping
values <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
groups <- c("A", "B", "A", "B", "A", "B", "A", "B", "A", "B")

# Apply function by groups
tapply(values, groups, mean)    # Mean by group
tapply(values, groups, sum)     # Sum by group
tapply(values, groups, length)  # Count by group

# Using tapply with built-in datasets
tapply(mtcars$mpg, mtcars$cyl, mean)    # Mean mpg by cylinder count
tapply(mtcars$hp, mtcars$gear, median)  # Median horsepower by gear count
```

# Data Subsetting and Indexing

R provides powerful subsetting capabilities for extracting specific elements from data structures.

## Vector Subsetting

```{r}
# Create sample vector
x <- c(10, 20, 30, 40, 50)
names(x) <- letters[1:5]

# Subsetting by position
x[1]              # First element
x[c(1, 3, 5)]     # Elements 1, 3, and 5
x[-2]             # All except second element
x[1:3]            # First three elements

# Subsetting by name
x["a"]            # Element named 'a'
x[c("a", "c")]    # Elements named 'a' and 'c'

# Logical subsetting
x[x > 25]         # Elements greater than 25
x[x %in% c(20, 40)]  # Elements equal to 20 or 40
```

## Data Frame Subsetting

```{r}
# Create sample data frame
df <- data.frame(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  age = c(25, 30, 35, 28, 32),
  salary = c(50000, 60000, 70000, 55000, 65000)
)
print(df)

# Subsetting columns
df$name           # Single column by name (returns vector)
df["name"]        # Single column by name (returns data frame)
df[c("name", "age")]  # Multiple columns

# Subsetting rows
df[1, ]           # First row
df[c(1, 3), ]     # Rows 1 and 3
df[df$age > 30, ] # Rows where age > 30

# Subsetting specific elements
df[1, "name"]     # Specific cell
df[2, 3]          # Row 2, column 3
df[df$salary > 60000, c("name", "salary")]  # Conditional row and column selection
```

## Matrix Subsetting

```{r}
# Create sample matrix
m <- matrix(1:20, nrow = 4, ncol = 5)
rownames(m) <- paste0("row", 1:4)
colnames(m) <- paste0("col", 1:5)
print(m)

# Subsetting by position
m[1, 2]           # Element in row 1, column 2
m[1, ]            # Entire first row
m[, 3]            # Entire third column
m[1:2, 3:4]       # Submatrix:  rows 1-2, columns 3-4

# Subsetting by name
m["row1", "col3"] # Element by row and column names
m[c("row1", "row3"), ] # Specific rows by name
```

# Advanced Function Applications

## Custom Functions with Apply

```{r}
# Define custom functions for use with apply family
# Function to calculate coefficient of variation
cv <- function(x) {
  sd(x) / mean(x) * 100
}

# Function to calculate range
my_range <- function(x) {
  max(x) - min(x)
}

# Apply custom functions
sample_data <- matrix(rnorm(20), nrow = 4)
apply(sample_data, 2, cv)        # CV for each column
apply(sample_data, 1, my_range)  # Range for each row
```

## Nested Apply Operations

```{r}
# Create nested list structure
nested_list <- list(
  group1 = list(a = 1:5, b = 6:10),
  group2 = list(c = 11:15, d = 16:20)
)

# Apply function to nested structure
lapply(nested_list, function(group) {
  sapply(group, mean)
})

# Alternative using nested apply
lapply(nested_list, function(group) {
  lapply(group, summary)
})
```

# Performance Considerations

## Vectorization vs Apply vs Loops

```{r}
# Create large dataset for performance comparison
n <- 10000
x <- rnorm(n)
m <- matrix(rnorm(n * 100), nrow = n)

# Timing different approaches (conceptual - actual timing may vary)
# Vectorized (fastest)
result1 <- colMeans(m)

# Apply family (fast)
result2 <- apply(m, 2, mean)

# Explicit loop (slowest)
result3 <- numeric(ncol(m))
for(i in 1:ncol(m)) {
  result3[i] <- mean(m[, i])
}

# All results should be equivalent
all.equal(result1, result2)
all.equal(result1, result3)
```

# Practical Examples

## Data Analysis with Base R Functions

```{r}
# Using mtcars dataset for practical examples
data(mtcars)
head(mtcars)

# Summary statistics by group using tapply
tapply(mtcars$mpg, mtcars$cyl, summary)

# Multiple statistics using sapply
stats_list <- list(
  mean_mpg = mean(mtcars$mpg),
  median_hp = median(mtcars$hp),
  sd_wt = sd(mtcars$wt)
)
sapply(stats_list, round, digits = 2)

# Apply functions to multiple columns
numeric_cols <- sapply(mtcars, is.numeric)
sapply(mtcars[numeric_cols], function(x) c(mean = mean(x), sd = sd(x)))
```

# Exercises

## Exercise 1: Matrix Operations
Create a 4x6 matrix of random numbers and use `apply()` to:
1. Calculate the maximum value in each row
2. Calculate the standard deviation of each column
3. Find the median of each row

```{r}
# Solution template: 
# set.seed(123)
# m <- matrix(rnorm(24), nrow = 4)
# apply(m, 1, max)      # Row maxima
# apply(m, 2, sd)       # Column standard deviations  
# apply(m, 1, median)   # Row medians
```

## Exercise 2: IQR Calculation
Write a base R snippet to compute the Interquartile Range (IQR) for each numeric column of the `mtcars` dataset.

```{r}
# Solution template: 
# sapply(mtcars, IQR)
# # or
# apply(mtcars, 2, IQR)
```

## Exercise 3: lapply vs sapply Comparison
Create a list with mixed data types and compare the behavior of `lapply()` vs `sapply()`.

```{r}
# Solution template: 
# mixed_list <- list(
#   numbers = 1:5,
#   characters = letters[1:3],
#   logicals = c(TRUE, FALSE, TRUE),
#   matrix = matrix(1:4, nrow = 2)
# )
# 
# lapply(mixed_list, class)  # Returns list
# sapply(mixed_list, class)  # Returns character vector
# 
# lapply(mixed_list, length) # Returns list
# sapply(mixed_list, length) # Returns named numeric vector
```

## Exercise 4: Custom Function with mapply
Use `mapply()` to calculate the area of rectangles given vectors of lengths and widths.

```{r}
# Solution template:
# lengths <- c(2, 4, 6, 8)
# widths <- c(3, 5, 7, 9)
# mapply(function(l, w) l * w, lengths, widths)
# # or simply: 
# mapply(`*`, lengths, widths)
```

## Exercise 5: Data Filtering and Aggregation
Using the built-in `iris` dataset:
1. Filter rows where Sepal.Length > 6
2. Calculate mean values for each numeric column by Species
3. Find the species with the highest average Petal.Width

```{r}
# Solution template:
# # 1. Filter
# filtered_iris <- iris[iris$Sepal.Length > 6, ]
# 
# # 2. Mean by species
# numeric_vars <- c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal. Width")
# species_means <- sapply(numeric_vars, function(var) {
#   tapply(iris[[var]], iris$Species, mean)
# })
# 
# # 3. Species with highest average Petal.Width
# petal_means <- tapply(iris$Petal. Width, iris$Species, mean)
# names(petal_means)[which.max(petal_means)]
```

# Summary

Base R functions and the apply family provide powerful tools for data manipulation and analysis: 

- **Utility functions** like `sum()`, `mean()`, `seq()`, and `rep()` form the foundation of data analysis
- **apply()** works on matrices and arrays across specified dimensions
- **lapply()** applies functions to lists, returning lists
- **sapply()** simplifies lapply results when possible
- **vapply()** provides type-safe apply operations
- **mapply()** handles multiple arguments element-wise
- **tapply()** applies functions by groups
- **Subsetting** allows precise data extraction using various indexing methods

Understanding these functions is crucial for efficient R programming and forms the basis for more advanced data manipulation techniques. 