---
title: "Data Types & Data Structures"
---

# Introduction

This comprehensive guide covers R's data types and structures with detailed explanations suitable for all skill levels. We'll explore not just *how* to use each structure, but *why* and *when* to use them effectively.

## Learning Objectives

By the end of this chapter, you will be able to: 

-   **Beginner**:  Understand basic data types and create simple data structures
-  **Intermediate**: Master complex data manipulation and understand memory concepts
-   **Advanced**: Optimize performance and implement advanced object-oriented concepts

## R's Memory Model:  Foundation Concepts

R stores all data as objects in memory.  Understanding this is crucial for efficient programming: 

: :: {.callout-important}
## Key Memory Concepts

- **Copy-on-modify**: R creates copies when objects are modified
- **Reference counting**:  R tracks how many variables point to an object
- **Garbage collection**:  Automatic memory cleanup of unused objects
- **Object attributes**: Metadata attached to R objects
::: 

#   Atomic Vectors:  The Building Blocks

## Understanding Atomic Vectors

Atomic vectors are the fundamental building blocks of R.  Every other data structure is built upon them.

### The Six Atomic Types

```{r}
#| label: atomic-types

# 1. Logical (boolean)
logical_vec <- c(TRUE, FALSE, TRUE, NA)
cat("Logical:", typeof(logical_vec), "\n")

# 2. Integer (whole numbers with L suffix)
integer_vec <- c(1L, 2L, 3L, NA_integer_)
cat("Integer:", typeof(integer_vec), "\n")

# 3. Double (real numbers)
double_vec <- c(1.5, 2.7, 3.14159, NA_real_)
cat("Double:", typeof(double_vec), "\n")

# 4. Character (strings)
character_vec <- c("hello", "world", "R", NA_character_)
cat("Character:", typeof(character_vec), "\n")

# 5. Complex (complex numbers)
complex_vec <- c(1+2i, 3+4i, 5+0i, NA_complex_)
cat("Complex:", typeof(complex_vec), "\n")

# 6. Raw (bytes - rarely used in typical analysis)
raw_vec <- charToRaw("Hello")
cat("Raw:", typeof(raw_vec), "\n")
```

###   Vector Properties and Inspection

```{r}
#| label: vector-properties

# Create a sample vector for analysis
sample_vec <- c(10, 20, 30, 40, 50)

# Essential properties every R user should know
length(sample_vec)    # Number of elements
typeof(sample_vec)    # Internal storage type
class(sample_vec)     # Object class (what R thinks it is)
mode(sample_vec)      # Storage mode (legacy, but still used)

# Check what kind of object we have
is.vector(sample_vec)    # Is it a vector?
is.atomic(sample_vec)    # Is it atomic (not a list)?
is.numeric(sample_vec)   # Is it numeric?
is.double(sample_vec)    # Specifically double precision?
```

**Theory**: Vectors are homogeneous - all elements must be the same type. When you mix types, R applies coercion rules.

###  Type Coercion Hierarchy

```{r}
#| label: coercion-hierarchy

# R's coercion hierarchy:  logical < integer < double < character
# Mixing types forces coercion to the "highest" type

# Logical to integer
logical_to_int <- c(TRUE, FALSE, 1L)
cat("Result type:", typeof(logical_to_int), "\n")
print(logical_to_int)

# Integer to double  
int_to_double <- c(1L, 2.5)
cat("Result type:", typeof(int_to_double), "\n")
print(int_to_double)

# Double to character (everything becomes character)
mixed_all <- c(TRUE, 1L, 2.5, "hello")
cat("Result type:", typeof(mixed_all), "\n")
print(mixed_all)
```

::: {.callout-warning}
## Coercion Warning

Automatic coercion can lead to unexpected results. Always check your data types when debugging! 
:::

###  Special Values in R

```{r}
#| label: special-values

# R has several special values that beginners often find confusing
special_vals <- c(
  normal = 42,
  missing = NA,           # Not Available (missing value)
  not_number = NaN,       # Not a Number (0/0)
  positive_inf = Inf,     # Positive infinity (1/0)
  negative_inf = -Inf,    # Negative infinity (-1/0)
  null_length = NULL      # NULL has length 0
)

# Functions to test for special values
test_vec <- c(1, NA, NaN, Inf, -Inf)
cat("Original vector:\n")
print(test_vec)

cat("\nTesting for special values:\n")
cat("is.na():", is.na(test_vec), "\n")         # TRUE for NA and NaN
cat("is.nan():", is.nan(test_vec), "\n")       # TRUE only for NaN
cat("is. infinite():", is.infinite(test_vec), "\n") # TRUE for ±Inf
cat("is. finite():", is.finite(test_vec), "\n")     # FALSE for NA, NaN, ±Inf
```

##   Vector Creation Methods

### Basic Creation Functions

```{r}
#| label: vector-creation-basic

# Method 1: c() function (combine/concatenate)
basic_vec <- c(1, 2, 3, 4, 5)

# Method 2: Colon operator (integer sequences)
seq_colon <- 1:10
reverse_seq <- 10:1

# Method 3: seq() function (more flexible sequences)
seq_by_step <- seq(from = 0, to = 100, by = 10)
seq_by_length <- seq(from = 0, to = 1, length.out = 11)
seq_along_other <- seq_along(c("a", "b", "c", "d"))

cat("Basic vector:", basic_vec, "\n")
cat("Colon sequence:", seq_colon, "\n") 
cat("Sequence by step:", seq_by_step, "\n")
cat("Sequence by length:", seq_by_length, "\n")
```

###  Advanced Creation Patterns

```{r}
#| label: vector-creation-advanced

# rep() function - repetition patterns
rep_times <- rep(c(1, 2, 3), times = 3)        # Repeat entire vector
rep_each <- rep(c(1, 2, 3), each = 3)          # Repeat each element
rep_length_out <- rep(c(1, 2), length.out = 7) # Repeat to specific length

cat("rep times:", rep_times, "\n")
cat("rep each:", rep_each, "\n")
cat("rep length. out:", rep_length_out, "\n")

# Creating named vectors
named_vec <- c(first = 1, second = 2, third = 3)
print(named_vec)

# Alternative way to add names
numbers <- 1:5
names(numbers) <- c("one", "two", "three", "four", "five")
print(numbers)

# Using paste() for systematic naming
systematic_names <- paste0("item_", 1:5)
cat("Systematic names:", systematic_names, "\n")
```

##  Vector Indexing and Subsetting

Vector indexing in R is extremely powerful and flexible.

### Positive Integer Indexing

```{r}
#| label: positive-indexing

# Create a sample vector with names
fruits <- c("apple", "banana", "cherry", "date", "elderberry", "fig")
names(fruits) <- paste0("fruit_", 1:6)
print(fruits)

# Single element access
cat("First fruit:", fruits[1], "\n")
cat("Third fruit:", fruits[3], "\n")

# Multiple specific elements
selected_fruits <- fruits[c(1, 3, 5)]
cat("Selected fruits:", selected_fruits, "\n")

# Range selection (slicing)
fruit_range <- fruits[2:4]
cat("Fruit range (2-4):", fruit_range, "\n")

# Using sequences for complex patterns
odd_positions <- fruits[seq(1, length(fruits), by = 2)]
cat("Odd positions:", odd_positions, "\n")
```

### Negative Integer Indexing (Exclusion)

```{r}
#| label: negative-indexing

# Negative indexing excludes elements
all_but_first <- fruits[-1]
cat("All except first:", all_but_first, "\n")

# Exclude multiple elements
exclude_multiple <- fruits[-c(2, 4, 6)]
cat("Excluding positions 2,4,6:", exclude_multiple, "\n")

# Exclude ranges
exclude_range <- fruits[-(2:4)]
cat("Excluding range 2-4:", exclude_range, "\n")
```

###  Logical Indexing (Conditional Selection)

```{r}
#| label: logical-indexing

# Create data for logical indexing examples
prices <- c(1.50, 3.20, 0.80, 4.50, 2.10, 6.00)
names(prices) <- fruits

# Simple logical conditions
expensive_items <- prices > 3.00
cat("Expensive items (>3.00):", prices[expensive_items], "\n")

# Multiple conditions with logical operators
moderate_prices <- prices >= 1.00 & prices <= 4.00
cat("Moderate prices (1-4):", prices[moderate_prices], "\n")

# Using %in% operator for membership testing
target_fruits <- c("apple", "cherry", "fig")
selected_by_name <- prices[names(prices) %in% target_fruits]
cat("Selected by name:", selected_by_name, "\n")

# Complex logical expressions
complex_condition <- (prices > 2.00) | (nchar(names(prices)) > 10)
cat("Complex condition:", prices[complex_condition], "\n")
```

### Named Indexing

```{r}
#| label: named-indexing

# Access by single name
apple_price <- prices["fruit_1"]
cat("Apple price:", apple_price, "\n")

# Access by multiple names
fruit_selection <- prices[c("fruit_1", "fruit_3", "fruit_6")]
cat("Selected fruit prices:", fruit_selection, "\n")

# Dynamic name creation
dynamic_names <- paste0("fruit_", c(1, 3, 5))
dynamic_selection <- prices[dynamic_names]
cat("Dynamic selection:", dynamic_selection, "\n")
```

##   Vector Operations and Arithmetic

### Scalar Operations (Broadcasting)

```{r}
#| label: scalar-operations

# Create sample data
values <- c(10, 20, 30, 40, 50)
scalar <- 5

# Arithmetic operations with scalars
cat("Original values:", values, "\n")
cat("Add 5:", values + scalar, "\n")
cat("Multiply by 5:", values * scalar, "\n")
cat("Divide by 5:", values / scalar, "\n")
cat("Power of 2:", values^2, "\n")
cat("Modulus 3:", values %% 3, "\n")
cat("Integer division by 3:", values %/% 3, "\n")
```

### Vector-to-Vector Operations

```{r}
#| label: vector-operations

# Element-wise operations between vectors of same length
vec1 <- c(1, 2, 3, 4, 5)
vec2 <- c(10, 20, 30, 40, 50)

cat("Vector 1:", vec1, "\n")
cat("Vector 2:", vec2, "\n")
cat("Addition:", vec1 + vec2, "\n")
cat("Multiplication:", vec1 * vec2, "\n")
cat("Division:", vec2 / vec1, "\n")

# Vector recycling with different lengths
short_vec <- c(1, 2)
long_vec <- c(10, 20, 30, 40, 50, 60)

cat("\nShort vector:", short_vec, "\n")
cat("Long vector:", long_vec, "\n")
cat("Addition (with recycling):", short_vec + long_vec, "\n")
```

: :: {.callout-note}
## Vector Recycling

When operating on vectors of different lengths, R recycles the shorter vector.  This is powerful but can lead to unexpected results if not understood properly. 
:::

###  Mathematical Functions

```{r}
#| label: math-functions

# Mathematical functions work element-wise on vectors
x <- c(1, 4, 9, 16, 25)

cat("Original values:", x, "\n")
cat("Square root:", sqrt(x), "\n")
cat("Natural log:", log(x), "\n")
cat("Log base 10:", log10(x), "\n")
cat("Exponential:", exp(c(0, 1, 2)), "\n")

# Trigonometric functions
angles <- c(0, pi/4, pi/2, pi)
cat("Angles (radians):", round(angles, 3), "\n")
cat("Sine:", round(sin(angles), 3), "\n")
cat("Cosine:", round(cos(angles), 3), "\n")

# Rounding functions
messy_numbers <- c(3.14159, 2.71828, 1.41421)
cat("Original:", messy_numbers, "\n")
cat("Round to 2 places:", round(messy_numbers, 2), "\n")
cat("Ceiling:", ceiling(messy_numbers), "\n")
cat("Floor:", floor(messy_numbers), "\n")
cat("Truncate:", trunc(messy_numbers), "\n")
```

##  Set Operations and Comparisons

### Set Operations

```{r}
#| label: set-operations

# Create sets for demonstration
set_a <- c("apple", "banana", "cherry", "date")
set_b <- c("cherry", "date", "elderberry", "fig")

cat("Set A:", set_a, "\n")
cat("Set B:", set_b, "\n")

# Core set operations
union_result <- union(set_a, set_b)
cat("Union (A ∪ B):", union_result, "\n")

intersection_result <- intersect(set_a, set_b)
cat("Intersection (A ∩ B):", intersection_result, "\n")

difference_result <- setdiff(set_a, set_b)
cat("Difference (A - B):", difference_result, "\n")

difference_reverse <- setdiff(set_b, set_a)
cat("Difference (B - A):", difference_reverse, "\n")

# Set equality and membership
cat("Are sets equal?", setequal(set_a, set_b), "\n")
cat("Is 'apple' in set A?", "apple" %in% set_a, "\n")
cat("Is 'grape' in set A?", "grape" %in% set_a, "\n")
```

###  Advanced Comparison Operations

```{r}
#| label: advanced-comparisons

# Comparison vectors
scores_1 <- c(85, 92, 78, 96, 88)
scores_2 <- c(88, 90, 78, 94, 91)

cat("Scores 1:", scores_1, "\n")
cat("Scores 2:", scores_2, "\n")

# Element-wise comparisons
cat("Equal elements:", scores_1 == scores_2, "\n")
cat("Score 1 greater:", scores_1 > scores_2, "\n")
cat("Score difference >= 5:", abs(scores_1 - scores_2) >= 5, "\n")

# Aggregated logical operations
cat("Any score1 > score2? ", any(scores_1 > scores_2), "\n")
cat("All scores > 75?", all(scores_1 > 75), "\n")
cat("How many scores1 > 85?", sum(scores_1 > 85), "\n")

# Finding positions where conditions are true
high_scores <- which(scores_1 > 90)
cat("Positions with scores > 90:", high_scores, "\n")

# Finding maximum and minimum positions
max_position <- which.max(scores_1)
min_position <- which.min(scores_1)
cat("Highest score position:", max_position, "value:", scores_1[max_position], "\n")
cat("Lowest score position:", min_position, "value:", scores_1[min_position], "\n")
```

#   Factors:  Categorical Data Excellence

## Understanding Factors

Factors are R's sophisticated way of handling categorical data. They're built on integers with associated labels, making them memory-efficient and statistically meaningful.

###   Basic Factor Creation

```{r}
#| label: basic-factors

# Create a simple factor from character data
colors <- c("red", "blue", "green", "red", "blue", "green", "red")
color_factor <- factor(colors)

print(color_factor)
cat("Levels:", levels(color_factor), "\n")
cat("Number of levels:", nlevels(color_factor), "\n")

# Internal representation
cat("Internal structure (integer codes):", as.integer(color_factor), "\n")
cat("Storage type:", typeof(color_factor), "\n")
cat("Memory usage comparison:\n")
cat("  Character:", object.size(colors), "bytes\n")
cat("  Factor:", object.size(color_factor), "bytes\n")
```

###  Controlling Factor Levels

```{r}
#| label: factor-levels

# Explicitly set levels and their order
sizes <- c("small", "large", "medium", "small", "medium", "large")

# Without specifying levels (alphabetical order)
size_factor_auto <- factor(sizes)
cat("Automatic levels:", levels(size_factor_auto), "\n")

# With explicitly ordered levels
size_factor_ordered <- factor(sizes, 
                             levels = c("small", "medium", "large"))
cat("Custom levels:", levels(size_factor_ordered), "\n")

# Ordered factors (with ranking)
size_factor_ranked <- factor(sizes, 
                           levels = c("small", "medium", "large"),
                           ordered = TRUE)
cat("Ordered factor:", size_factor_ranked, "\n")
cat("Is ordered? ", is.ordered(size_factor_ranked), "\n")

# Comparisons work with ordered factors
cat("small < medium? ", size_factor_ranked[1] < size_factor_ranked[3], "\n")
```

###  Factor Level Manipulation

```{r}
#| label: factor-manipulation

# Create a factor for manipulation
ratings <- factor(c("good", "bad", "excellent", "good", "fair", "bad"))
print(ratings)

# View current levels
cat("Original levels:", levels(ratings), "\n")

# Reorder levels logically
levels(ratings) <- c("bad", "fair", "good", "excellent")
cat("Reordered levels:", levels(ratings), "\n")

# Add new levels (useful before adding new data)
levels(ratings) <- c(levels(ratings), "outstanding")
cat("After adding level:", levels(ratings), "\n")

# Remove unused levels
ratings_subset <- ratings[ratings != "bad"]
cat("Before droplevels:", levels(ratings_subset), "\n")
ratings_clean <- droplevels(ratings_subset)
cat("After droplevels:", levels(ratings_clean), "\n")
```

###   Advanced Factor Operations with forcats

```{r}
#| label: advanced-factors
#| message: false

library(forcats)

# Create sample survey data
responses <- factor(c("Agree", "Disagree", "Strongly Agree", "Neutral", 
                     "Agree", "Strongly Disagree", "Agree", "Neutral",
                     "Disagree", "Strongly Agree"))

# Reorder by frequency
freq_ordered <- fct_infreq(responses)
cat("By frequency:\n")
print(table(freq_ordered))

# Reorder by another variable
scores <- c(4, 2, 5, 3, 4, 1, 4, 3, 2, 5)
score_ordered <- fct_reorder(responses, scores, mean)
cat("Ordered by mean score:\n")
print(tapply(scores, score_ordered, mean))

# Collapse rare levels
collapsed <- fct_lump_min(responses, min = 2)
cat("After collapsing rare levels:\n")
print(table(collapsed))

# Recode levels
recoded <- fct_recode(responses,
  "Positive" = "Agree",
  "Very Positive" = "Strongly Agree", 
  "Negative" = "Disagree",
  "Very Negative" = "Strongly Disagree"
)
print(table(recoded))
```

#   Lists: The Swiss Army Knife

## Understanding Lists

Lists are recursive data structures - they can contain any R object, including other lists. This makes them incredibly flexible for complex data structures.

###   Basic List Creation and Structure

```{r}
#| label: basic-lists

# Create a heterogeneous list
student_record <- list(
  name = "Alice Johnson",
  student_id = 12345,
  grades = c(85, 92, 78, 96, 88),
  courses = c("Math", "Science", "English", "History", "Art"),
  is_honors = TRUE,
  graduation_date = as.Date("2024-06-15")
)

print(student_record)

# Examine list structure
cat("\nList structure:\n")
str(student_record)

# List properties
cat("Length (number of elements):", length(student_record), "\n")
cat("Names of elements:", names(student_record), "\n")
cat("Is it a list?", is.list(student_record), "\n")
```

###  List Indexing Methods

Understanding the difference between `[`, `[[`, and `$` is crucial for list manipulation.

```{r}
#| label:  list-indexing

# Method 1: Single bracket [ ] - returns a list
subset_list <- student_record[1]  # Returns list containing first element
cat("Using [1] - class:", class(subset_list), "\n")
print(subset_list)

# Multiple elements with single bracket
multiple_elements <- student_record[c("name", "student_id")]
print(multiple_elements)

# Method 2: Double bracket [[ ]] - returns the actual element
actual_name <- student_record[[1]]  # Returns the character vector
cat("Using [[1]] - class:", class(actual_name), "\n")
print(actual_name)

# Access by name with double bracket
grades <- student_record[["grades"]]
cat("Grades class:", class(grades), "\n")
print(grades)

# Method 3: Dollar sign $ - convenient for named elements
student_name <- student_record$name
cat("Using $ - class:", class(student_name), "\n")
print(student_name)
```

::: {.callout-tip}
## List Indexing Memory Aid

- `my_list[1]` → Returns a list (like a box containing the item)
- `my_list[[1]]` → Returns the actual item (takes item out of box)
- `my_list$name` → Convenient shortcut for `my_list[["name"]]`
:::

###  List Modification and Manipulation

```{r}
#| label: list-modification

student_record <- list(
  name = "Alice Johnson",
  student_id = 12345,
  grades = c(85, 92, 78, 96, 88),
  courses = c("Math", "Science", "English", "History", "Art"),
  is_honors = TRUE,
  graduation_date = as.Date("2024-06-15")
)
# Start with a copy for modification
student_copy <- student_record

# Add new elements
student_copy$gpa <- mean(student_copy$grades) / 100 * 4  # Simple GPA calculation
student_copy$advisor <- "Dr. Smith"
student_copy$extracurricular <- c("Drama Club", "Science Olympics")

# Modify existing elements
student_copy$grades[1] <- 90  # Change first grade
student_copy$is_honors <- FALSE

# Remove elements (set to NULL)
student_copy$advisor <- NULL

cat("Modified list structure:\n")
str(student_copy)

# Adding nested structure
student_copy$contact_info <- list(
  email = "alice. johnson@school.edu",
  phone = "555-0123",
  address = list(
    street = "123 Main St",
    city = "Academic City",
    zip = "12345"
  )
)

cat("After adding nested structure:\n")
str(student_copy, max.level = 2)  # Limit depth for readability
```

###  List Apply Functions

The apply family of functions is essential for working with lists efficiently.

```{r}
#| label: list-apply

# Create a list of numeric vectors for demonstration
data_list <- list(
  dataset_1 = c(23, 45, 67, 89, 12),
  dataset_2 = c(34, 56, 78, 90, 23, 45),
  dataset_3 = c(12, 34, 56, 78, 90, 12, 34)
)

# lapply - returns a list
means_list <- lapply(data_list, mean)
cat("lapply result (list):\n")
print(means_list)

# sapply - simplifies to vector when possible
means_vector <- sapply(data_list, mean)
cat("sapply result (vector):", means_vector, "\n")

# mapply - multivariate apply
weights <- c(0.3, 0.5, 0.2)
weighted_means <- mapply(function(x, w) sum(x * w), data_list, weights)
cat("Weighted means:", weighted_means, "\n")

# vapply - like sapply but with type checking (safer)
lengths_safe <- vapply(data_list, length, integer(1))
cat("Lengths (vapply):", lengths_safe, "\n")
```

###   Advanced List Operations with purrr

```{r}
#| label: advanced-list-purrr
#| message: false

library(purrr)

# Modern functional programming with purrr
# map family - consistent and predictable
data_list <- list(
  dataset_1 = c(23, 45, 67, 89, 12),
  dataset_2 = c(34, 56, 78, 90, 23, 45),
  dataset_3 = c(12, 34, 56, 78, 90, 12, 34)
)
# map returns a list
summary_stats <- map(data_list, summary)
print(summary_stats[[1]])

# map_dbl returns numeric vector
means_purrr <- map_dbl(data_list, mean)
cat("Means with map_dbl:", means_purrr, "\n")

# map_chr returns character vector  
length_descriptions <- map_chr(data_list, ~ paste("Length:", length(.x)))
cat("Descriptions:", length_descriptions, "\n")


# map2 for working with two lists simultaneously
multipliers <- c(2, 3, 4)
scaled_data <- map2(data_list, multipliers, ~ .x * .y)
cat("First scaled dataset:", scaled_data[[1]], "\n")

# Complex transformations with map
complex_stats <- data_list |>
  map(~ list(
    mean = mean(.x),
    sd = sd(.x),
    min = min(.x),
    max = max(.x),
    n = length(.x)
  ))

print(complex_stats$dataset_1)
```

#   Matrices: Linear Algebra Powerhouse

## Understanding Matrices

Matrices are 2-dimensional, homogeneous data structures.  They're essentially vectors with a dimension attribute, making them perfect for mathematical operations.

###   Matrix Creation Methods

```{r}
#| label: matrix-creation

# Method 1: matrix() function
basic_matrix <- matrix(1:12, nrow = 3, ncol = 4)
cat("Basic matrix (filled by columns):\n")
print(basic_matrix)

# Fill by rows instead
by_rows <- matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
cat("Matrix filled by rows:\n")
print(by_rows)

# Method 2: Combining vectors
row_1 <- c(1, 2, 3, 4)
row_2 <- c(5, 6, 7, 8)
row_3 <- c(9, 10, 11, 12)

# Row binding
matrix_rbind <- rbind(row_1, row_2, row_3)
cat("Matrix from row binding:\n")
print(matrix_rbind)

# Column binding
col_matrix <- cbind(c(1, 5, 9), c(2, 6, 10), c(3, 7, 11), c(4, 8, 12))
cat("Matrix from column binding:\n")
print(col_matrix)

# Method 3: Adding dimensions to a vector
vector_to_matrix <- 1:12
dim(vector_to_matrix) <- c(3, 4)
cat("Vector converted to matrix:\n")
print(vector_to_matrix)
```

###  Matrix Properties and Attributes

```{r}
#| label: matrix-properties

# Create a named matrix for demonstration
sales_matrix <- matrix(
  c(100, 150, 120, 200, 
    110, 160, 130, 210,
    105, 155, 125, 195),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    quarters = c("Q1", "Q2", "Q3"),
    products = c("Product_A", "Product_B", "Product_C", "Product_D")
  )
)

print(sales_matrix)

# Matrix properties
cat("Dimensions:", dim(sales_matrix), "\n")
cat("Number of rows:", nrow(sales_matrix), "\n")
cat("Number of columns:", ncol(sales_matrix), "\n")
cat("Total elements:", length(sales_matrix), "\n")

# Dimension names
cat("Row names:", rownames(sales_matrix), "\n")
cat("Column names:", colnames(sales_matrix), "\n")

# Matrix type information
cat("Class:", class(sales_matrix), "\n")
cat("Type:", typeof(sales_matrix), "\n")
cat("Is matrix?", is.matrix(sales_matrix), "\n")
cat("Is array?", is.array(sales_matrix), "\n")  # Matrices are also arrays
```

###  Matrix Indexing and Subsetting

```{r}
#| label: matrix-indexing

# Single element access
cat("Element [2,3]:", sales_matrix[2, 3], "\n")
cat("Q2 Product_C:", sales_matrix["Q2", "Product_C"], "\n")

# Row access (returns vector)
q1_sales <- sales_matrix[1, ]
cat("Q1 sales:", q1_sales, "\n")
cat("Class of row:", class(q1_sales), "\n")  # Note: becomes vector

# Column access
product_a_sales <- sales_matrix[, 1]
cat("Product A sales:", product_a_sales, "\n")

# Multiple rows/columns (returns matrix)
multi_quarters <- sales_matrix[1:2, ]
cat("Q1-Q2 sales:\n")
print(multi_quarters)

multi_products <- sales_matrix[, 2:4]
cat("Products B-D:\n")
print(multi_products)

# Submatrix
submatrix <- sales_matrix[1:2, 2:3]
cat("Submatrix Q1-Q2, Products B-C:\n")
print(submatrix)

# Logical indexing
high_sales <- sales_matrix > 150
cat("High sales (>150):\n")
print(sales_matrix[high_sales])  # Returns as vector
```

###   Matrix Arithmetic Operations

```{r}
#| label: matrix-arithmetic

# Create matrices for operations
A <- matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2)
B <- matrix(c(5, 6, 7, 8), nrow = 2, ncol = 2)

cat("Matrix A:\n"); print(A)
cat("Matrix B:\n"); print(B)

# Element-wise operations
cat("A + B (element-wise addition):\n")
print(A + B)

cat("A * B (element-wise multiplication):\n")
print(A * B)

# Matrix multiplication (linear algebra)
cat("A %*% B (matrix multiplication):\n")
print(A %*% B)

# Scalar operations
cat("A * 3 (scalar multiplication):\n")
print(A * 3)

# Matrix powers
cat("A^2 (element-wise squaring):\n")
print(A^2)
```

###  Advanced Matrix Operations

```{r}
#| label: advanced-matrix

# Create a square matrix for advanced operations
square_matrix <- matrix(c(4, 7, 2, 6), nrow = 2, ncol = 2)
cat("Square matrix:\n"); print(square_matrix)

# Linear algebra operations
cat("Transpose:\n"); print(t(square_matrix))
cat("Determinant:", det(square_matrix), "\n")

# Matrix inverse
inverse_matrix <- solve(square_matrix)
cat("Inverse matrix:\n"); print(inverse_matrix)

# Verify inverse:  A * A^(-1) = I
identity_check <- square_matrix %*% inverse_matrix
cat("A * A^(-1) (should be identity):\n")
print(round(identity_check, 10))  # Round to handle floating point errors

# Eigenvalues and eigenvectors
eigen_result <- eigen(square_matrix)
cat("Eigenvalues:", eigen_result$values, "\n")
cat("Eigenvectors:\n"); print(eigen_result$vectors)

# Matrix decomposition
qr_decomp <- qr(square_matrix)
cat("QR decomposition Q:\n"); print(qr.Q(qr_decomp))
cat("QR decomposition R:\n"); print(qr.R(qr_decomp))
```

###   Matrix Apply Operations

```{r}
#| label: matrix-apply

# Create a larger matrix for demonstration
test_matrix <- matrix(rnorm(20, mean = 10, sd = 3), nrow = 4, ncol = 5)
rownames(test_matrix) <- paste("Row", 1:4)
colnames(test_matrix) <- paste("Col", 1:5)

cat("Test matrix:\n")
print(round(test_matrix, 2))

# Apply functions across margins
# MARGIN = 1 for rows, MARGIN = 2 for columns

# Row operations
row_sums <- apply(test_matrix, 1, sum)
row_means <- apply(test_matrix, 1, mean)
row_ranges <- apply(test_matrix, 1, function(x) max(x) - min(x))

cat("Row sums:", round(row_sums, 2), "\n")
cat("Row means:", round(row_means, 2), "\n")
cat("Row ranges:", round(row_ranges, 2), "\n")

# Column operations  
col_sums <- apply(test_matrix, 2, sum)
col_sds <- apply(test_matrix, 2, sd)

cat("Column sums:", round(col_sums, 2), "\n")
cat("Column standard deviations:", round(col_sds, 2), "\n")

# Built-in functions (faster than apply for simple operations)
cat("rowSums() result:", round(rowSums(test_matrix), 2), "\n")
cat("colMeans() result:", round(colMeans(test_matrix), 2), "\n")
```

#   Data Frames: The Data Analysis Workhorse

## Understanding Data Frames

Data frames are the most important data structure for data analysis.  They're like matrices but allow different data types in different columns, making them perfect for real-world datasets.

###   Data Frame Creation

```{r}
#| label: dataframe-creation

# Basic data frame creation
employee_df <- data.frame(
  employee_id = 1:6,
  first_name = c("Alice", "Bob", "Charlie", "Diana", "Eve", "Frank"),
  last_name = c("Johnson", "Smith", "Brown", "Davis", "Wilson", "Miller"),
  department = c("IT", "Sales", "IT", "HR", "Sales", "IT"),
  salary = c(75000, 65000, 80000, 70000, 68000, 82000),
  start_date = as.Date(c("2020-01-15", "2019-06-01", "2021-03-10", 
                        "2020-08-20", "2019-11-05", "2022-02-01")),
  is_manager = c(FALSE, TRUE, FALSE, TRUE, FALSE, TRUE),
  performance_rating = factor(c("Excellent", "Good", "Excellent", 
                               "Good", "Fair", "Excellent"),
                             levels = c("Poor", "Fair", "Good", "Excellent"),
                             ordered = TRUE),
  stringsAsFactors = FALSE  # Keep strings as characters
)

print(employee_df)
```

###  Data Frame Properties and Inspection

```{r}
#| label: dataframe-properties

# Basic properties
cat("Dimensions:", dim(employee_df), "\n")
cat("Number of rows:", nrow(employee_df), "\n")
cat("Number of columns:", ncol(employee_df), "\n")
cat("Column names:", names(employee_df), "\n")

# Data structure overview
cat("Data frame structure:\n")
str(employee_df)

# Summary statistics
cat("Summary statistics:\n")
summary(employee_df)

# Data types of each column
column_types <- sapply(employee_df, class)
cat("Column types:\n")
print(column_types)

# Check for missing values
missing_values <- colSums(is.na(employee_df))
cat("Missing values per column:\n")
print(missing_values)
```

###   Data Frame Indexing and Access

```{r}
#| label: dataframe-indexing

# Column access methods
cat("First names (using $):", employee_df$first_name, "\n")
cat("First names (using [[]]):", employee_df[["first_name"]], "\n")
cat("Class of extracted column:", class(employee_df$salary), "\n")

# Single bracket returns data frame
first_name_df <- employee_df["first_name"]
cat("Class of [column]:", class(first_name_df), "\n")

# Multiple column selection
name_salary <- employee_df[c("first_name", "last_name", "salary")]
cat("Multiple columns:\n")
print(head(name_salary, 3))

# Row access
first_employee <- employee_df[1, ]
cat("First employee:\n")
print(first_employee)

# Specific cell access
alice_salary <- employee_df[1, "salary"]
cat("Alice's salary:", alice_salary, "\n")

# Range selection
first_three <- employee_df[1:3, c("first_name", "department", "salary")]
cat("First three employees (selected columns):\n")
print(first_three)
```

###  Data Frame Filtering and Subsetting

```{r}
#| label: dataframe-filtering
employee_df <- data.frame(
  employee_id = 1:6,
  first_name = c("Alice", "Bob", "Charlie", "Diana", "Eve", "Frank"),
  last_name = c("Johnson", "Smith", "Brown", "Davis", "Wilson", "Miller"),
  department = c("IT", "Sales", "IT", "HR", "Sales", "IT"),
  salary = c(75000, 65000, 80000, 70000, 68000, 82000),
  start_date = as.Date(c("2020-01-15", "2019-06-01", "2021-03-10", 
                        "2020-08-20", "2019-11-05", "2022-02-01")),
  is_manager = c(FALSE, TRUE, FALSE, TRUE, FALSE, TRUE),
  performance_rating = factor(c("Excellent", "Good", "Excellent", 
                               "Good", "Fair", "Excellent"),
                             levels = c("Poor", "Fair", "Good", "Excellent"),
                             ordered = TRUE),
  stringsAsFactors = FALSE
)
# Simple filtering
it_employees <- employee_df[employee_df$department == "IT", ]
cat("IT employees:\n")
print(it_employees[, c("first_name", "last_name", "salary")])

# Multiple conditions with logical operators
high_earners_it <- employee_df[employee_df$department == "IT" & 
                              employee_df$salary > 75000, ]
cat("High-earning IT employees:\n")
print(high_earners_it[, c("first_name", "salary")])

# Using subset() function (more readable)
managers_subset <- subset(employee_df, 
                         is_manager == TRUE,
                         select = c("first_name", "last_name", "department", "salary"))
cat("Managers (using subset):\n")
print(managers_subset)

# Complex filtering with %in%
target_departments <- c("IT", "Sales")
it_or_sales <- employee_df[employee_df$department %in% target_departments, ]
cat("IT or Sales employees:\n")
print(it_or_sales[, c("first_name", "department", "salary")])

# Filtering with date conditions
recent_hires <- employee_df[employee_df$start_date > as.Date("2020-01-01"), ]
cat("Recent hires (after 2020-01-01):\n")
print(recent_hires[, c("first_name", "start_date")])
```

###  Data Frame Modification

```{r}
#| label: dataframe-modification

# Create a working copy
df_work <- employee_df

# Add new columns
df_work$full_name <- paste(df_work$first_name, df_work$last_name)
df_work$years_employed <- as.numeric(Sys.Date() - df_work$start_date) / 365.25
df_work$salary_category <- ifelse(df_work$salary > 75000, "High", "Standard")

# Conditional column creation
df_work$bonus_eligible <- with(df_work, 
  (years_employed > 2) & (performance_rating %in% c("Good", "Excellent"))
)

# Calculate bonus amount based on conditions
df_work$bonus <- ifelse(df_work$bonus_eligible,
                       df_work$salary * 0.1,  # 10% bonus
                       0)

cat("Modified data frame with new columns:\n")
print(df_work[, c("full_name", "years_employed", "salary_category", "bonus")])

# Modify existing values
df_work$salary[df_work$first_name == "Alice"] <- 78000  # Give Alice a raise
df_work$is_manager[df_work$first_name == "Charlie"] <- TRUE  # Promote Charlie

# Add new rows
new_employee <- data.frame(
  employee_id = 7,
  first_name = "Grace",
  last_name = "Lee",
  department = "HR",
  salary = 72000,
  start_date = as.Date("2023-01-15"),
  is_manager = FALSE,
  performance_rating = factor("Good", levels = levels(df_work$performance_rating)),
  full_name = "Grace Lee",
  years_employed = 1.0,
  salary_category = "Standard",
  bonus_eligible = FALSE,
  bonus = 0
)

df_work <- rbind(df_work, new_employee)
cat("After adding new employee:", nrow(df_work), "total employees\n")
```

###   Advanced Data Frame Operations

```{r}
#| label: advanced-dataframe
#| message: false

# Using dplyr for modern data manipulation
library(dplyr)

# Modern data manipulation with pipes
summary_stats <- employee_df |>
  mutate(
    full_name = paste(first_name, last_name),
    years_employed = as.numeric(Sys.Date() - start_date) / 365.25,
    salary_k = salary / 1000  # Convert to thousands
  ) |>
  filter(department %in% c("IT", "Sales")) |>
  group_by(department, is_manager) |>
  summarise(
    count = n(),
    avg_salary = mean(salary),
    avg_years = mean(years_employed),
    .groups = "drop"
  ) |>
  arrange(department, desc(is_manager))

cat("Summary statistics by department and management status:\n")
print(summary_stats)

# Advanced filtering with case_when
employee_df_enhanced <- employee_df |>
  mutate(
    experience_level = case_when(
      as.numeric(Sys.Date() - start_date) / 365.25 < 2 ~ "Junior",
      as.numeric(Sys.Date() - start_date) / 365.25 < 4 ~ "Mid-level", 
      TRUE ~ "Senior"
    ),
    compensation_tier = case_when(
      salary < 70000 ~ "Tier 1",
      salary < 80000 ~ "Tier 2",
      TRUE ~ "Tier 3"
    )
  )

cat("Enhanced employee data with calculated fields:\n")
print(employee_df_enhanced |> 
       select(first_name, department, experience_level, compensation_tier))
```

###  Data Frame Aggregation and Grouping

```{r}
#| label: dataframe-aggregation

# Base R aggregation
dept_summary <- aggregate(salary ~ department, data = employee_df, 
                         FUN = function(x) c(mean = mean(x), 
                                           median = median(x), 
                                           count = length(x)))
cat("Department salary summary:\n")
print(dept_summary)

# Multiple grouping variables
dept_mgr_summary <- aggregate(salary ~ department + is_manager, 
                             data = employee_df, 
                             FUN = mean)
cat("Average salary by department and management status:\n")
print(dept_mgr_summary)

# Using by() function for more complex operations
performance_by_dept <- by(employee_df$performance_rating, 
                         employee_df$department, 
                         table)
cat("Performance ratings by department:\n")
print(performance_by_dept)

# Table operations for cross-tabulation
dept_performance_table <- table(employee_df$department, 
                               employee_df$performance_rating)
cat("Department vs Performance cross-table:\n")
print(dept_performance_table)

# Proportional tables
dept_performance_prop <- prop.table(dept_performance_table, margin = 1)
cat("Performance distribution within each department:\n")
print(round(dept_performance_prop, 3))
```

#  Arrays: Multi-Dimensional Data

## Understanding Arrays

Arrays extend matrices to more than two dimensions.  They're useful for storing multi-dimensional scientific or business data.

###  Array Creation and Structure

```{r}
#| label: array-creation

# Create a 3-dimensional array (think:  multiple matrices stacked)
# Dimensions: 4 quarters × 3 products × 2 years
sales_array <- array(
  data = c(100, 120, 110, 130,    # Q1-Q4 for Product A, Year 1
           150, 160, 140, 170,    # Q1-Q4 for Product B, Year 1  
           200, 210, 190, 220,    # Q1-Q4 for Product C, Year 1
           110, 130, 120, 140,    # Q1-Q4 for Product A, Year 2
           160, 170, 150, 180,    # Q1-Q4 for Product B, Year 2
           210, 220, 200, 230),   # Q1-Q4 for Product C, Year 2
  dim = c(4, 3, 2),  # 4 quarters, 3 products, 2 years
  dimnames = list(
    quarters = paste0("Q", 1:4),
    products = c("Product_A", "Product_B", "Product_C"), 
    years = c("2022", "2023")
  )
)

print(sales_array)

# Array properties
cat("Array dimensions:", dim(sales_array), "\n")
cat("Dimension names:\n")
print(dimnames(sales_array))
cat("Total elements:", length(sales_array), "\n")
```

###  Array Indexing and Slicing

```{r}
#| label:  array-indexing

# Single element access
q2_productb_2022 <- sales_array["Q2", "Product_B", "2022"]
cat("Q2 Product B 2022 sales:", q2_productb_2022, "\n")

# Slice operations
# All quarters and products for 2022
sales_2022 <- sales_array[, , "2022"]
cat("All 2022 sales:\n"); print(sales_2022)

# Product A sales across all quarters and years
product_a_sales <- sales_array[, "Product_A", ]
cat("Product A sales across time:\n"); print(product_a_sales)

# Q1 sales for all products and years
q1_sales <- sales_array["Q1", , ]
cat("Q1 sales across products and years:\n"); print(q1_sales)

# Multiple quarters for specific product and year
q1_q2_productc_2023 <- sales_array[c("Q1", "Q2"), "Product_C", "2023"]
cat("Q1-Q2 Product C 2023:", q1_q2_productc_2023, "\n")
```

###   Array Operations with Apply

```{r}
#| label: array-apply

# Apply operations across different dimensions
# MARGIN = 1: across quarters
# MARGIN = 2: across products  
# MARGIN = 3: across years

# Total sales by quarter (sum across products and years)
quarterly_totals <- apply(sales_array, 1, sum)
cat("Quarterly totals:", quarterly_totals, "\n")

# Total sales by product (sum across quarters and years)
product_totals <- apply(sales_array, 2, sum)
cat("Product totals:", product_totals, "\n")

# Total sales by year (sum across quarters and products)
yearly_totals <- apply(sales_array, 3, sum)
cat("Yearly totals:", yearly_totals, "\n")

# Average sales by quarter and product (across years)
avg_by_quarter_product <- apply(sales_array, c(1, 2), mean)
cat("Average sales by quarter and product (across years):\n")
print(avg_by_quarter_product)

# Growth rates between years for each quarter-product combination
growth_rates <- apply(sales_array, c(1, 2), function(x) (x[2] - x[1]) / x[1] * 100)
cat("Growth rates (2022 to 2023) by quarter and product:\n")
print(round(growth_rates, 1))
```

#   Advanced Data Structures and Concepts

## Environments:  R's Scoping System

```{r}
#| label:  environments

# Environments store object-name bindings
current_env <- environment()
global_env <- .GlobalEnv

# Create a new environment
my_env <- new.env()
my_env$x <- 10
my_env$y <- 20
my_env$calculate <- function(a, b) a + b

# List objects in environment
cat("Objects in my_env:", ls(my_env), "\n")

# Access environment objects
cat("x from my_env:", my_env$x, "\n")
cat("Calculation result:", my_env$calculate(my_env$x, my_env$y), "\n")

# Environment hierarchy
cat("Current environment:", environmentName(current_env), "\n")
cat("Parent environment:", environmentName(parent.env(current_env)), "\n")

# Search path for R objects
search_path <- search()
cat("R search path:\n")
print(search_path)
```

## Object-Oriented Programming:  S3 Classes

```{r}
#| label: s3-classes

# Create an S3 object (simple object-oriented programming)
create_bank_account <- function(owner, balance = 0) {
  account <- list(
    owner = owner,
    balance = balance,
    transactions = data.frame(
      date = as.Date(character(0)),
      type = character(0),
      amount = numeric(0),
      stringsAsFactors = FALSE
    )
  )
  class(account) <- "bank_account"
  return(account)
}

# Create methods for our class
print.bank_account <- function(x) {
  cat("Bank Account\n")
  cat("Owner:", x$owner, "\n")
  cat("Balance:  $", x$balance, "\n")
  cat("Transactions:", nrow(x$transactions), "\n")
}

deposit.bank_account <- function(account, amount) {
  if (amount <= 0) {
    stop("Deposit amount must be positive")
  }
  account$balance <- account$balance + amount
  new_transaction <- data.frame(
    date = Sys.Date(),
    type = "deposit",
    amount = amount
  )
  account$transactions <- rbind(account$transactions, new_transaction)
  return(account)
}

withdraw.bank_account <- function(account, amount) {
  if (amount <= 0) {
    stop("Withdrawal amount must be positive")
  }
  if (amount > account$balance) {
    stop("Insufficient funds")
  }
  account$balance <- account$balance - amount
  new_transaction <- data.frame(
    date = Sys.Date(),
    type = "withdrawal", 
    amount = -amount
  )
  account$transactions <- rbind(account$transactions, new_transaction)
  return(account)
}

# Use our S3 class
my_account <- create_bank_account("John Doe", 1000)
print(my_account)

my_account <- deposit.bank_account(my_account, 500)
my_account <- withdraw.bank_account(my_account, 200)
print(my_account)
```

#   Performance Optimization and Best Practices

## Memory Efficiency and Performance

### Pre-allocation vs Growing Objects

```{r}
#| label: performance-preallocation

# Demonstrate the performance difference
n <- 10000

# BAD: Growing objects (very slow)
system.time({
  bad_vector <- numeric(0)
  for (i in 1:n) {
    bad_vector <- c(bad_vector, i^2)
  }
})

# GOOD: Pre-allocation (much faster)
system.time({
  good_vector <- numeric(n)
  for (i in 1:n) {
    good_vector[i] <- i^2
  }
})

# BEST: Vectorized operations (fastest)
system.time({
  best_vector <- (1:n)^2
})

# Verify all methods give same result
cat("All methods equivalent:", 
    identical(bad_vector, good_vector) && identical(good_vector, best_vector), "\n")
```

### Data Structure Selection for Performance

```{r}
#| label: performance-structures
#| message: false

library(microbenchmark)

# Create test data
n_rows <- 1000
test_matrix <- matrix(rnorm(n_rows * 10), nrow = n_rows)
test_df <- as.data.frame(test_matrix)

# Compare performance for column operations
performance_test <- microbenchmark(
  matrix_colsums = colSums(test_matrix),
  df_apply = apply(test_df, 2, sum),
  df_sapply = sapply(test_df, sum),
  times = 100
)

print(performance_test)

# Memory usage comparison
cat("Matrix memory:", as.numeric(object.size(test_matrix)), "bytes\n")
cat("Data frame memory:", as.numeric(object.size(test_df)), "bytes\n")
```

### Factors vs Characters for Memory

```{r}
#| label: memory-factors

# Compare memory usage:  factors vs characters for categorical data
n_observations <- 100000
categories <- c("Category_A", "Category_B", "Category_C", "Category_D")

# Create character vector
char_vector <- sample(categories, n_observations, replace = TRUE)

# Create factor
factor_vector <- factor(char_vector)

# Memory comparison
char_size <- as.numeric(object.size(char_vector))
factor_size <- as.numeric(object.size(factor_vector))

cat("Character vector memory:", char_size, "bytes\n")
cat("Factor memory:", factor_size, "bytes\n")
cat("Factor saves:", round((char_size - factor_size) / char_size * 100, 1), "% memory\n")

# Performance comparison for operations
cat("\nPerformance comparison for table operations:\n")
performance_categorical <- microbenchmark(
  character_table = table(char_vector),
  factor_table = table(factor_vector),
  times = 50
)
print(performance_categorical)
```

## Best Practices Summary

### 1. Data Structure Selection Guide

```{r}
#| label: selection-guide

# Function to recommend data structure
recommend_structure <- function(data_description) {
  cat("Data Structure Recommendation Guide\n")
  cat("===================================\n\n")
  
  recommendations <- list(
    "Homogeneous numeric data, mathematical operations" = "Matrix or Numeric Vector",
    "Heterogeneous tabular data" = "Data Frame or Tibble", 
    "Categorical data with known levels" = "Factor",
    "Hierarchical or nested data" = "List",
    "Mixed-type collection of objects" = "List",
    "Multi-dimensional scientific data" = "Array",
    "Simple sequence of same-type values" = "Atomic Vector"
  )
  
  for (i in seq_along(recommendations)) {
    cat("•", names(recommendations)[i], "\n")
    cat("  →", recommendations[[i]], "\n\n")
  }
}

recommend_structure()
```

### 2. Common Pitfalls and Solutions

```{r}
#| label: pitfalls

# Common pitfall demonstrations and solutions

cat("PITFALL 1: Automatic type coercion\n")
mixed_data <- c(1, 2, 3, "four")
cat("Result type:", typeof(mixed_data), "\n")
cat("Solution: Use lists for mixed types\n\n")

cat("PITFALL 2: Factor level surprises\n")
original_factor <- factor(c("low", "medium", "high"))
cat("Original levels:", levels(original_factor), "\n")
# Adding new level without updating factor
# new_factor <- c(original_factor, "very_high")  # This would cause issues
cat("Solution: Update levels before adding new data\n\n")

cat("PITFALL 3: Ignoring missing values\n")
data_with_na <- c(1, 2, NA, 4, 5)
cat("Mean without na. rm:", mean(data_with_na), "\n")
cat("Mean with na.rm=TRUE:", mean(data_with_na, na.rm = TRUE), "\n\n")

cat("PITFALL 4: Matrix vs data frame confusion\n")
test_matrix <- matrix(1:6, nrow = 2)
test_df <- data.frame(x = 1:2, y = 3:4, z = 5:6)
cat("Matrix column access returns:", class(test_matrix[, 1]), "\n")
cat("Data frame column access returns:", class(test_df[, 1]), "\n")
cat("Solution: Use drop=FALSE to maintain structure\n")
cat("Matrix with drop=FALSE:", class(test_matrix[, 1, drop = FALSE]), "\n")
```

### 3. Debugging and Validation Functions

```{r}
#| label: debugging-functions

# Utility functions for data structure debugging
debug_structure <- function(obj, name = deparse(substitute(obj))) {
  cat("=== Debug Info for", name, "===\n")
  cat("Class:", class(obj), "\n")
  cat("Type:", typeof(obj), "\n")
  cat("Length/Dimensions:", 
      if(is.null(dim(obj))) length(obj) else paste(dim(obj), collapse = "×"), "\n")
  
  if (is.factor(obj)) {
    cat("Factor levels:", nlevels(obj), "\n")
    cat("Ordered:", is.ordered(obj), "\n")
  }
  
  if (any(is.na(obj))) {
    cat("Missing values:", sum(is.na(obj)), "\n")
  }
  
  cat("Memory usage:", as.numeric(object.size(obj)), "bytes\n")
  cat("========================\n\n")
}

# Validate data frame structure
validate_dataframe <- function(df, required_cols = NULL) {
  cat("Data Frame Validation\n")
  cat("====================\n")
  
  # Basic structure
  cat("Dimensions:", paste(dim(df), collapse = " × "), "\n")
  cat("Complete cases:", sum(complete.cases(df)), "/", nrow(df), "\n")
  
  # Check required columns
  if (!is.null(required_cols)) {
    missing_cols <- required_cols[!required_cols %in% names(df)]
    if (length(missing_cols) > 0) {
      cat("⚠️  Missing required columns:", paste(missing_cols, collapse = ", "), "\n")
    } else {
      cat("✅ All required columns present\n")
    }
  }
  
  # Data type summary
  cat("\nColumn types:\n")
  col_types <- sapply(df, function(x) paste(class(x), collapse = "/"))
  for (i in seq_along(col_types)) {
    cat("  ", names(col_types)[i], ":", col_types[i], "\n")
  }
  
  cat("====================\n\n")
}

# Example usage
debug_structure(employee_df)
validate_dataframe(employee_df, required_cols = c("employee_id", "first_name", "salary"))
```

# Summary and Quick Reference

## Data Structure Comparison Table

| Structure | Dimensions | Homogeneous | Best Use Cases | Memory Efficiency |
|-----------|------------|-------------|----------------|-------------------|
| **Vector** | 1D | ✅ Yes | Mathematical operations, sequences | ⭐⭐⭐⭐⭐ |
| **Factor** | 1D | ✅ Yes | Categorical data, statistical modeling | ⭐⭐⭐⭐⭐ |
| **Matrix** | 2D | ✅ Yes | Linear algebra, image data | ⭐⭐⭐⭐ |
| **Array** | ND | ✅ Yes | Multi-dimensional scientific data | ⭐⭐⭐⭐ |
| **List** | 1D | ❌ No | Complex nested structures | ⭐⭐⭐ |
| **Data Frame** | 2D | ❌ No | Data analysis, mixed-type tables | ⭐⭐⭐ |

## Key Takeaways

###   For Beginners
- **Start with vectors and data frames** - they cover 80% of use cases
- **Understand indexing** - mastering `$`, `[]`, and `[[]]` is crucial
- **Use factors for categorical data** - they save memory and improve performance
###  For Intermediate Users
- **Leverage lists for complex data** - they allow nesting and mixed types
- **Master apply functions** - they enable efficient operations across data structures
- **Choose the right structure** - consider data type, operations, and memory needs
###   For Advanced Users
- **Optimize performance** - pre-allocate objects and prefer vectorized operations
- **Utilize S3 classes** - for custom object-oriented programming
- **Debug and validate** - use custom functions to ensure data integrity and structure
