---
title: "String Functions"
---


# Learning Objectives

By the end of this chapter, you will be able to: 

- Master fundamental string manipulation functions in R
- Apply string functions to pharmaceutical data cleaning and processing
- Use regular expressions for complex pattern matching in drug names and codes
- Handle medication names, dosage forms, and clinical terminology
- Clean and standardize pharmaceutical datasets using string operations

# Introduction

String manipulation is crucial in pharmaceutical data analysis. You'll frequently work with: 

- Drug names and generic/brand name mapping
- Dosage forms and strengths
- Medical terminology and coding systems (ICD, MedDRA, ATC codes)
- Patient identifiers and study codes
- Adverse event descriptions
- Clinical trial data with text fields

# Setup

```{r setup}
# Load required packages
library(stringr)
library(dplyr)
library(purrr)
library(tools)

# Suppress warnings for cleaner output
options(warn = -1)
```

# Basic String Functions

## 1. String Length and Basic Information

```{r basic-info}
# Basic string functions with pharmaceutical examples
drug_names <- c("Aspirin", "Ibuprofen", "Acetaminophen", "Warfarin", "Metformin")

# Get string lengths
cat("String lengths:\n")
print(nchar(drug_names))

# Check if strings are empty
medical_terms <- c("Hypertension", "", "Diabetes", NA, "Pneumonia")
cat("\nEmpty string check:\n")
print(nzchar(medical_terms))  # Returns FALSE for empty strings

# Handle NA values properly
cat("\nNA value check:\n")
print(is.na(medical_terms))
```

## 2. Case Conversion

```{r case-conversion}
# Case conversion for drug names
brand_names <- c("Tylenol", "Advil", "Lipitor", "Nexium")

cat("Original names:", paste(brand_names, collapse = ", "), "\n")
cat("Uppercase:", paste(toupper(brand_names), collapse = ", "), "\n")
cat("Lowercase:", paste(tolower(brand_names), collapse = ", "), "\n")
cat("Title Case:", paste(toTitleCase(tolower(brand_names)), collapse = ", "), "\n")
```

## 3. String Concatenation

```{r concatenation}
# Combining drug information
drug_name <- c("Aspirin", "Ibuprofen", "Acetaminophen")
strength <- c("81mg", "200mg", "500mg")
form <- c("Tablet", "Capsule", "Tablet")

# Basic concatenation
cat("Basic concatenation:\n")
print(paste(drug_name, strength, form))

# Concatenation with custom separator
cat("\nWith custom separator:\n")
print(paste(drug_name, strength, form, sep = " | "))

# Concatenation without separator
cat("\nWithout separator:\n")
print(paste0(drug_name, "_", strength))
```

# Substring Operations

## 1. Extracting Substrings

```{r substrings}
# Working with NDC (National Drug Code) numbers
ndc_codes <- c("12345-678-90", "98765-432-10", "11111-222-33")

cat("NDC Codes:", paste(ndc_codes, collapse = ", "), "\n")

# Extract manufacturer code (first part)
cat("Manufacturer codes (base R):", paste(substr(ndc_codes, 1, 5), collapse = ", "), "\n")

# Extract product code (middle part)
cat("Product codes:", paste(substr(ndc_codes, 7, 9), collapse = ", "), "\n")

# Using stringr for more intuitive operations
cat("First 5 chars (stringr):", paste(str_sub(ndc_codes, 1, 5), collapse = ", "), "\n")
cat("Last 2 chars:", paste(str_sub(ndc_codes, -2, -1), collapse = ", "), "\n")
```

## 2. Finding and Replacing Text

```{r find-replace}
# Standardizing drug names
messy_drug_names <- c("aspirin 81 mg", "IBUPROFEN-200MG", "Acetaminophen/500mg")

cat("Original messy names:\n")
print(messy_drug_names)

# Replace hyphens and slashes with spaces
cleaned_names <- str_replace_all(messy_drug_names, "[-/]", " ")
cat("\nAfter replacing separators:\n")
print(cleaned_names)

# Remove extra whitespace
cat("\nAfter removing extra whitespace:\n")
print(str_squish(cleaned_names))

# Find positions of patterns
drug_text <- "Patient took Aspirin 81mg twice daily"
cat("\nPosition of 'Aspirin' in text:\n")
print(str_locate(drug_text, "Aspirin"))

cat("\nPositions of all numbers:\n")
print(str_locate_all(drug_text, "[0-9]+"))
```

# Pattern Matching and Regular Expressions

## 1. Basic Pattern Detection

```{r pattern-detection}
# Detect patterns in medication descriptions
medications <- c(
  "Aspirin 81mg tablet once daily",
  "Metformin 500mg twice daily",
  "Insulin injection as needed",
  "Lisinopril 10mg daily",
  "Warfarin 5mg as directed"
)

cat("Medications containing 'mg':\n")
print(medications[str_detect(medications, "mg")])

cat("\nMedications starting with 'Aspirin':\n")
print(medications[str_detect(medications, "^Aspirin")])

cat("\nMedications ending with 'daily':\n")
print(medications[str_detect(medications, "daily$")])
```

## 2. Advanced Regular Expressions

```{r advanced-regex}
# Extract dosage information
dosage_pattern <- "[0-9]+\\.?[0-9]*mg"
extracted_dosages <- str_extract(medications, dosage_pattern)

cat("Extracted dosages:\n")
for(i in seq_along(medications)) {
  cat(sprintf("%s -> %s\n", medications[i], 
              ifelse(is.na(extracted_dosages[i]), "No dosage found", extracted_dosages[i])))
}

# Complex pattern:  Extract drug name and dosage
drug_dosage_pattern <- "^([A-Za-z]+)\\s+([0-9]+\\.?[0-9]*mg)"
matches <- str_match(medications, drug_dosage_pattern)

cat("\nDrug name and dosage extraction:\n")
for(i in seq_along(medications)) {
  if(! is.na(matches[i, 1])) {
    cat(sprintf("Drug: %s, Dosage:  %s\n", matches[i, 2], matches[i, 3]))
  } else {
    cat(sprintf("No match for:  %s\n", medications[i]))
  }
}
```

# Working with coded terms

## 1. ICD-10 Code Processing

```{r icd-codes}
# Working with ICD-10 diagnostic codes
icd_codes <- c("E11.9", "I10", "J44.1", "N18.6", "M79.3")
icd_descriptions <- c(
  "Type 2 diabetes mellitus without complications",
  "Essential hypertension",
  "Chronic obstructive pulmonary disease with acute exacerbation",
  "End stage renal disease",
  "Panniculitis, unspecified"
)

# Create a lookup function
create_icd_lookup <- function(codes, descriptions) {
  lookup_table <- setNames(descriptions, codes)
  return(lookup_table)
}

icd_lookup <- create_icd_lookup(icd_codes, icd_descriptions)

# Function to get description from code
get_icd_description <- function(code, lookup_table) {
  description <- lookup_table[code]
  ifelse(is.na(description), "Code not found", description)
}

# Test the function
patient_codes <- c("E11.9", "I10", "Z99.9")  # Last one doesn't exist
cat("ICD-10 Code Lookup Results:\n")
for(i in seq_along(patient_codes)) {
  cat(sprintf("%s:  %s\n", patient_codes[i], get_icd_description(patient_codes[i], icd_lookup)))
}
```

## 2. ATC Code Analysis

```{r atc-codes}
# Anatomical Therapeutic Chemical (ATC) codes
atc_codes <- c("A02BC01", "C09AA02", "A10BA02", "B01AA03", "N02BA01")
drug_names_atc <- c("Omeprazole", "Enalapril", "Metformin", "Warfarin", "Aspirin")

# Extract anatomical group (first character)
anatomical_groups <- str_sub(atc_codes, 1, 1)

# Create descriptive labels
anatomical_labels <- c(
  "A" = "Alimentary tract and metabolism",
  "B" = "Blood and blood forming organs",
  "C" = "Cardiovascular system",
  "D" = "Dermatologicals",
  "G" = "Genito urinary system and sex hormones",
  "H" = "Systemic hormonal preparations",
  "J" = "Antiinfectives for systemic use",
  "L" = "Antineoplastic and immunomodulating agents",
  "M" = "Musculo-skeletal system",
  "N" = "Nervous system",
  "P" = "Antiparasitic products",
  "R" = "Respiratory system",
  "S" = "Sensory organs",
  "V" = "Various"
)

# Map codes to descriptions
atc_mapping <- data.frame(
  Drug = drug_names_atc,
  ATC_Code = atc_codes,
  Anatomical_Group = anatomical_labels[anatomical_groups],
  stringsAsFactors = FALSE
)

cat("ATC Code Analysis:\n")
print(atc_mapping)
```

# Data Cleaning Applications

## 1. Standardizing Drug Names

```{r drug-standardization}
# Real-world messy drug data
messy_drug_data <- c(
  "ASPIRIN 81 MG TABLET",
  "aspirin 81mg tab",
  "Aspirin, 81 mg, tablet",
  "ASA 81MG TABS",
  "aspirin_81_mg_tablet",
  "Aspirin (81mg) - Tablet"
)

# Function to standardize drug names
standardize_drug_name <- function(drug_names) {
  # Convert to lowercase
  clean_names <- tolower(drug_names)
  
  # Replace various separators with spaces
  clean_names <- str_replace_all(clean_names, "[_,\\-()]+", " ")
  
  # Standardize abbreviations
  clean_names <- str_replace_all(clean_names, "\\btab[s]?\\b", "tablet")
  clean_names <- str_replace_all(clean_names, "\\basa\\b", "aspirin")
  
  # Remove extra whitespace
  clean_names <- str_squish(clean_names)
  
  # Standardize mg format
  clean_names <- str_replace_all(clean_names, "([0-9]+)\\s*mg", "\\1mg")
  
  return(clean_names)
}

standardized_names <- standardize_drug_name(messy_drug_data)

cat("Drug Name Standardization Results:\n")
for(i in seq_along(messy_drug_data)) {
  cat(sprintf("Original: %s\n", messy_drug_data[i]))
  cat(sprintf("Standardized: %s\n\n", standardized_names[i]))
}
```

## 2. Parsing Dosage Information

```{r dosage-parsing}
# Extract structured information from prescription strings
prescriptions <- c(
  "Metformin 500mg twice daily with meals",
  "Lisinopril 10mg once daily in morning",
  "Warfarin 5mg daily as directed by INR",
  "Aspirin 81mg once daily for cardioprotection",
  "Insulin 10 units subcutaneous before meals"
)

# Function to parse prescription information
parse_prescription <- function(prescription_text) {
  # Extract drug name (assumes it's the first word)
  drug_name <- str_extract(prescription_text, "^[A-Za-z]+")
  
  # Extract dosage with units
  dosage <- str_extract(prescription_text, "[0-9]+\\.?[0-9]*\\s?(mg|units|mcg|g)")
  
  # Extract frequency information
  frequency <- case_when(
    str_detect(prescription_text, "twice|bid") ~ "Twice daily",
    str_detect(prescription_text, "once|daily") ~ "Once daily",
    str_detect(prescription_text, "three times|tid") ~ "Three times daily",
    str_detect(prescription_text, "four times|qid") ~ "Four times daily",
    TRUE ~ "As directed"
  )
  
  # Extract special instructions
  special_instructions <- case_when(
    str_detect(prescription_text, "with meals") ~ "With meals",
    str_detect(prescription_text, "before meals") ~ "Before meals",
    str_detect(prescription_text, "morning") ~ "In morning",
    str_detect(prescription_text, "bedtime") ~ "At bedtime",
    TRUE ~ "No special instructions"
  )
  
  return(data.frame(
    Drug = drug_name,
    Dosage = dosage,
    Frequency = frequency,
    Instructions = special_instructions,
    stringsAsFactors = FALSE
  ))
}

# Parse all prescriptions
prescription_data <- map_dfr(prescriptions, parse_prescription)

cat("Parsed Prescription Information:\n")
print(prescription_data)
```

# Advanced String Operations

## 1. Working with Patient Identifiers

```{r patient-ids}
# Function to generate patient IDs
generate_patient_id <- function(site_code, patient_number) {
  # Format: SITE001-P-0001
  formatted_number <- str_pad(patient_number, 4, pad = "0")
  patient_id <- paste0(site_code, "-P-", formatted_number)
  return(patient_id)
}

# Generate some patient IDs
site_codes <- c("NYC", "LAX", "CHI")
patient_numbers <- 1:3

# Create patient IDs for demonstration
sample_ids <- c()
for(site in site_codes) {
  for(num in patient_numbers) {
    sample_ids <- c(sample_ids, generate_patient_id(site, num))
  }
}

cat("Generated Patient IDs:\n")
print(sample_ids[1:6])  # Show first 6

# Function to validate patient ID format
validate_patient_id <- function(patient_id) {
  pattern <- "^[A-Z]{3}-P-[0-9]{4}$"
  return(str_detect(patient_id, pattern))
}

# Test validation
test_ids <- c("NYC-P-0001", "INVALID-ID", "LAX-P-99", "CHI-P-0123")
cat("\nValidation Results:\n")
for(i in seq_along(test_ids)) {
  cat(sprintf("%s: %s\n", test_ids[i], 
              ifelse(validate_patient_id(test_ids[i]), "Valid", "Invalid")))
}
```

## 2. Processing Adverse Event Reports

```{r adverse-events}
# Working with MedDRA terms
adverse_events <- c(
  "Nausea and vomiting",
  "Severe headache",
  "Skin rash and itching",
  "Diarrhea, mild",
  "Fatigue and weakness",
  "Allergic reaction - hives"
)

# Function to standardize adverse event terms
standardize_ae_terms <- function(ae_terms) {
  # Convert to title case
  clean_terms <- str_to_title(ae_terms)
  
  # Standardize common words
  clean_terms <- str_replace_all(clean_terms, "\\bAnd\\b", "and")
  clean_terms <- str_replace_all(clean_terms, "\\bOf\\b", "of")
  
  # Remove severity indicators for consistency
  severity_pattern <- "(\\b(mild|moderate|severe|serious)\\s*,? \\s*|\\s*-\\s*)"
  clean_terms <- str_remove_all(clean_terms, regex(severity_pattern, ignore_case = TRUE))
  
  # Remove extra whitespace
  clean_terms <- str_squish(clean_terms)
  
  return(clean_terms)
}

standardized_ae <- standardize_ae_terms(adverse_events)

cat("Adverse Event Standardization:\n")
for(i in seq_along(adverse_events)) {
  cat(sprintf("Original: %s\n", adverse_events[i]))
  cat(sprintf("Standardized: %s\n\n", standardized_ae[i]))
}
```

## 3. Clinical Trial Data Processing Example

```{r clinical-trial-example}
library(purrr)
# Comprehensive example:  Processing clinical trial medication data
clinical_trial_data <- data.frame(
  subject_id = c("001-001", "001-002", "001-003"),
  visit = c("Baseline", "Week 4", "Week 8"),
  concomitant_medications = c(
    "Metformin 500mg BID, Lisinopril 10mg QD",
    "Metformin 500mg BID, Lisinopril 10mg QD, Aspirin 81mg QD",
    "Metformin 1000mg BID, Lisinopril 20mg QD"
  ),
  stringsAsFactors = FALSE
)

# Function to parse medication strings into structured data
parse_medications <- function(med_string, subject_id, visit) {
  # Split by comma to get individual medications
  individual_meds <- str_split(med_string, ",\\s*")[[1]]
  
  # Extract information for each medication
  med_data <- map_dfr(individual_meds, function(med) {
    # Extract drug name (first word(s) before dosage)
    # Fixed:  Removed invalid lookahead syntax
    drug_name <- str_extract(med, "^[A-Za-z\\s]+(? =\\s+[0-9])")
    drug_name <- str_trim(drug_name)
    
    # Extract dosage
    dosage <- str_extract(med, "[0-9]+\\.?[0-9]*\\s*(mg|mcg|g|units)")
    
    # Extract frequency
    frequency <- case_when(
      str_detect(med, "BID|twice") ~ "Twice daily",
      str_detect(med, "QD|once|daily") ~ "Once daily",
      str_detect(med, "TID|three") ~ "Three times daily",
      str_detect(med, "QID|four") ~ "Four times daily",
      TRUE ~ "As directed"
    )
    
    return(data.frame(
      drug_name = drug_name,
      dosage = dosage,
      frequency = frequency,
      stringsAsFactors = FALSE
    ))
  })
  
  # Add subject and visit information
  med_data$subject_id <- subject_id
  med_data$visit <- visit
  
  return(med_data)
}

# Alternative function that's more robust (doesn't rely on lookahead)
parse_medications_robust <- function(med_string, subject_id, visit) {
  # Split by comma to get individual medications
  individual_meds <- str_split(med_string, ",\\s*")[[1]]
  
  # Extract information for each medication
  med_data <- map_dfr(individual_meds, function(med) {
    # Method 1: Use word boundaries to extract drug name
    # Extract everything before the first number
    drug_name <- str_extract(med, "^[^0-9]+")
    drug_name <- str_trim(drug_name)
    
    # Method 2: Alternative approach using str_match
    # match <- str_match(med, "^([A-Za-z\\s]+)\\s+([0-9]+\\.?[0-9]*\\s*(mg|mcg|g|units))")
    # drug_name <- str_trim(match[, 2])
    # dosage <- match[, 3]
    
    # Extract dosage
    dosage <- str_extract(med, "[0-9]+\\.?[0-9]*\\s*(mg|mcg|g|units)")
    
    # Extract frequency
    frequency <- case_when(
      str_detect(med, "BID|twice") ~ "Twice daily",
      str_detect(med, "QD|once|daily") ~ "Once daily",
      str_detect(med, "TID|three") ~ "Three times daily",
      str_detect(med, "QID|four") ~ "Four times daily",
      TRUE ~ "As directed"
    )
    
    return(data.frame(
      drug_name = drug_name,
      dosage = dosage,
      frequency = frequency,
      stringsAsFactors = FALSE
    ))
  })
  
  # Add subject and visit information
  med_data$subject_id <- subject_id
  med_data$visit <- visit
  
  return(med_data)
}

# Process all medication data using the robust function
processed_meds <- pmap_dfr(
  list(
    clinical_trial_data$concomitant_medications,
    clinical_trial_data$subject_id,
    clinical_trial_data$visit
  ),
  parse_medications_robust
)

# Display the structured data
cat("Processed Clinical Trial Medication Data:\n")
print(processed_meds[, c("subject_id", "visit", "drug_name", "dosage", "frequency")])
```

# Performance Tips and Best Practices

## 1. Efficient String Operations

```{r performance-tips}
# Create a moderately sized dataset for demonstration
large_drug_list <- rep(c("Aspirin", "Ibuprofen", "Acetaminophen"), 1000)

# Vectorized operations are more efficient
cat("Efficient approach - vectorized operation:\n")
start_time <- Sys.time()
drug_doses_vectorized <- paste(large_drug_list, "100mg")
end_time <- Sys.time()
cat(sprintf("Time taken: %s seconds\n", round(end_time - start_time, 4)))

# Avoid loops for simple operations
cat("\nLess efficient approach - loop-based:\n")
start_time <- Sys.time()
drug_doses_loop <- character(length(large_drug_list))
for (i in seq_along(large_drug_list)) {
  drug_doses_loop[i] <- paste(large_drug_list[i], "100mg")
}
end_time <- Sys.time()
cat(sprintf("Time taken: %s seconds\n", round(end_time - start_time, 4)))

cat("\nFirst 5 results are identical:")
cat(sprintf("Vectorized: %s\n", paste(drug_doses_vectorized[1:5], collapse = ", ")))
cat(sprintf("Loop-based: %s\n", paste(drug_doses_loop[1:5], collapse = ", ")))
```

## 2. Error Handling in String Operations

```{r error-handling}
# Robust function for extracting dosage information
extract_dosage_safe <- function(medication_string) {
  tryCatch({
    # Handle NULL or NA inputs
    if (is.null(medication_string) || is.na(medication_string)) {
      return(NA_character_)
    }
    
    # Convert to character if not already
    med_str <- as.character(medication_string)
    
    # Extract dosage pattern
    dosage_pattern <- "[0-9]+\\.?[0-9]*\\s*(mg|mcg|g|units)"
    dosage <- str_extract(med_str, regex(dosage_pattern, ignore_case = TRUE))
    
    # Return standardized format
    if (! is.na(dosage)) {
      return(str_to_lower(str_replace(dosage, "\\s+", "")))
    } else {
      return("No dosage found")
    }
    
  }, error = function(e) {
    warning(paste("Error processing medication string:", medication_string, "-", e$message))
    return("Error in processing")
  })
}

# Test with various inputs
test_medications <- c(
  "Aspirin 81mg",
  "Invalid format",
  NA,
  "Metformin 500 MG twice daily",
  ""
)

cat("Safe Dosage Extraction Results:\n")
for(i in seq_along(test_medications)) {
  result <- extract_dosage_safe(test_medications[i])
  cat(sprintf("Input: '%s' -> Output: '%s'\n", 
              ifelse(is.na(test_medications[i]), "NA", test_medications[i]), 
              result))
}
```

# Summary and Key Takeaways

## Essential String Functions for Pharmaceutical Analysis

The key functions you should master include:

1. **Basic Operations**:  `nchar()`, `toupper()`, `tolower()`, `paste()`, `paste0()`
2. **Substring Operations**: `substr()`, `str_sub()`, `str_extract()`
3. **Pattern Matching**: `str_detect()`, `str_locate()`, `str_match()`
4. **Text Replacement**: `str_replace()`, `str_replace_all()`, `str_remove()`
5. **Text Cleaning**: `str_squish()`, `str_trim()`

## Best Practices

```{r best-practices}
# Create a summary of best practices
best_practices <- data.frame(
  Practice = c(
    "Standardization",
    "Error Handling", 
    "Vectorization",
    "Regular Expressions",
    "Documentation"
  ),
  Description = c(
    "Always clean and standardize data before analysis",
    "Use tryCatch() for robust string processing",
    "Use vectorized operations for better performance", 
    "Learn regex patterns for complex text matching",
    "Document your string processing logic clearly"
  ),
  Example = c(
    "toupper(), str_squish()",
    "tryCatch(), is.na() checks",
    "str_replace_all() vs loops", 
    "str_extract() with patterns",
    "Comments and function names"
  ),
  stringsAsFactors = FALSE
)

cat("String Processing Best Practices:\n")
print(best_practices)
```

## Common Pharmaceutical Applications

- **Drug name standardization and mapping**:  Cleaning messy drug names from different sources
- **Dosage extraction and validation**: Parsing prescription information
- **Medical coding**: Working with ICD-10, ATC, MedDRA codes
- **Patient identifier processing**: Generating and validating study IDs
- **Adverse event text mining**: Standardizing AE terms and descriptions
- **Clinical trial data parsing**: Processing complex medication strings

## Next Steps

1. Practice these techniques with your own pharmaceutical datasets
2. Explore more advanced text mining packages (tm, quanteda, tidytext)
3. Learn about natural language processing for clinical notes
4. Study regulatory requirements for data standardization (CDISC standards)
5. Implement automated data quality checks using string validation functions

---

*This comprehensive guide provides the foundation for effective string manipulation in pharmaceutical R programming. Remember to adapt these techniques to your specific use cases and always validate your results with domain experts.*


# Comparing stringr and base R string functions

We'll begin with a lookup table between the most important stringr functions and their base R equivalents.

```{r}
library(stringr)
data_stringr_base_diff <- tibble::tribble(
  ~stringr,                                        ~base_r,
  "str_detect(string, pattern)",                   "grepl(pattern, x)",
  "str_dup(string, times)",                        "strrep(x, times)",
  "str_extract(string, pattern)",                  "regmatches(x, m = regexpr(pattern, text))",
  "str_extract_all(string, pattern)",              "regmatches(x, m = gregexpr(pattern, text))",
  "str_length(string)",                            "nchar(x)",
  "str_locate(string, pattern)",                   "regexpr(pattern, text)",
  "str_locate_all(string, pattern)",               "gregexpr(pattern, text)",
  "str_match(string, pattern)",                    "regmatches(x, m = regexec(pattern, text))",
  "str_order(string)",                             "order(...)",
  "str_replace(string, pattern, replacement)",     "sub(pattern, replacement, x)",
  "str_replace_all(string, pattern, replacement)", "gsub(pattern, replacement, x)",
  "str_sort(string)",                              "sort(x)",
  "str_split(string, pattern)",                    "strsplit(x, split)",
  "str_sub(string, start, end)",                   "substr(x, start, stop)",
  "str_subset(string, pattern)",                   "grep(pattern, x, value = TRUE)",
  "str_to_lower(string)",                          "tolower(x)",
  "str_to_title(string)",                          "tools::toTitleCase(text)",
  "str_to_upper(string)",                          "toupper(x)",
  "str_trim(string)",                              "trimws(x)",
  "str_which(string, pattern)",                    "grep(pattern, x)",
  "str_wrap(string)",                              "strwrap(x)"
)

# create MD table, arranged alphabetically by stringr fn name
data_stringr_base_diff |> 
  dplyr::mutate(dplyr::across(
      .cols = everything(),
      .fns = ~ paste0("`", .x, "`"))
  ) |> 
  dplyr::arrange(stringr) |> 
  dplyr::rename(`base R` = base_r) |> 
  gt::gt() |> 
  gt::fmt_markdown(columns = everything()) |> 
  gt::tab_options(column_labels.font.weight = "bold")
```

Overall the main differences between base R and stringr are:

1.  stringr functions start with `str_` prefix; base R string functions have no consistent naming scheme.

2.  The order of inputs is usually different between base R and stringr. In base R, the `pattern` to match usually comes first; in stringr, the `string` to manupulate always comes first. This makes stringr easier to use in pipes, and with `lapply()` or `purrr::map()`.

3.  Functions in stringr tend to do less, where many of the string processing functions in base R have multiple purposes.

4.  The output and input of stringr functions has been carefully designed. For example, the output of `str_locate()` can be fed directly into `str_sub()`; the same is not true of `regexpr()` and `substr()`.

5.  Base functions use arguments (like `perl`, `fixed`, and `ignore.case`) to control how the pattern is interpreted. To avoid dependence between arguments, stringr instead uses helper functions (like `fixed()`, `regex()`, and `coll()`).

Next we'll walk through each of the functions, noting the similarities and important differences. These examples are adapted from the stringr documentation and here they are contrasted with the analogous base R operations.


```{r}
library(dplyr)
ae <- haven::read_sas("./data/sdtm/ae.sas7bdat")
ae1 <- ae |> 
  select (USUBJID,AEDECOD,AEBODSYS)

#find function in sas, str_detect

ae2 <- ae1 |> 
  filter(str_detect(AEDECOD,"HER"))

ae2 <- ae1 |> 
  filter(str_detect(AEDECOD,"^HIATUS ")) # start of the string

ae2 <- ae1 |> 
  filter(str_detect(AEDECOD,"HERNIA$")) # end of the string

#substr in sas

ae2 <- ae1 |> 
  mutate(newvar1=str_sub(AEDECOD,1,6),
         newvar2=str_sub(AEDECOD,2,6),
         newvar3=str_sub(AEDECOD,-3),
         newvar4=str_length(AEDECOD))

#cat function in sas , Str_c in r

ae2 <- ae1 |> 
  mutate(newvar1=str_c(AEDECOD,AEBODSYS,sep="/"),
         newvar2=paste(AEDECOD,AEBODSYS,sep="/"))

# scan function in sas, word in r

ae2 <- ae1 |> 
  mutate(newvar1=word(AEDECOD,1),
         newvar2=word(AEDECOD,2))

# upper & lower case

ae2 <- ae1 |> 
  mutate(newvar1=str_to_lower(AEDECOD),
         newvar2=str_to_upper(AEDECOD),
         newvar3=str_to_title(AEDECOD),
         newvar4=str_to_sentence(AEDECOD))

#str_trim

a <- "  this is    my String   "
b <- str_trim(a)
c <- str_replace_all(a," "," ")
d <- str_squish(a)

```
