---
title: "date and time functions"
---

### Date and Time Base Functions in R

Date and time manipulation is a common task in data analysis. R provides several base functions to work with dates and times. Here are some of the most commonly used base R functions for handling date and time data:

```{r}
# Get the current date
current_date <- Sys.Date()
print(current_date)
# Get the current date and time
current_datetime <- Sys.time()
print(current_datetime)
```
### Creating Date and Time Objects
### Year Formats
| Code | Meaning                     | Example (`format(dt, ...)`) |
| ---- | --------------------------- | --------------------------- |
| `%Y` | 4-digit year                | `"2025"`                    |
| `%y` | 2-digit year (00–99)        | `"25"`                      |
| `%C` | Century (year / 100, 00–99) | `"20"` (for 2025; OS-dep.)  |

### Month Formats
| Code | Meaning                           | Example      |
| ---- | --------------------------------- | ------------ |
| `%m` | Month number (01–12)              | `"11"`       |
| `%b` | Abbrev. month name (locale)       | `"Nov"`      |
| `%B` | Full month name (locale)          | `"November"` |
| `%h` | Same as `%b` (abbrev. month name) | `"Nov"`      |

### Day Formats
| Code | Meaning                                | Example                             |
| ---- | -------------------------------------- | ----------------------------------- |
| `%d` | Day of month (01–31)                   | `"14"`                              |
| `%e` | Day of month (1–31, padded with space) | `"14"` (or `" 7"` for 7th; OS-dep.) |
| `%j` | Day of year (001–366)                  | `"318"`                             |

### Weekday Formats

| Code | Meaning                                            | Example        |
| ---- | -------------------------------------------------- | -------------- |
| `%a` | Abbrev. weekday name (locale)                      | `"Fri"`        |
| `%A` | Full weekday name (locale)                         | `"Friday"`     |
| `%w` | Weekday number (0–6, **0 = Sunday**)               | `"5"`          |
| `%u` | ISO weekday number (1–7, **1 = Monday**) (OS-dep.) | `"5"` (Friday) |

### Time Formats
| Code | Meaning                        | Example    |
| ---- | ------------------------------ | ---------- |
| `%H` | Hour (00–23)                   | `"14"`     |
| `%I` | Hour (01–12)                   | `"02"`     |
| `%M` | Minute (00–59)                 | `"30"`     |
| `%S` | Second (00–61, allows for leap) | `"05"`    |
| `%p` | AM/PM indicator (locale)       | `"PM"`     |

###Locale-dependent “full date/time” codes

###These depend on your system locale (language + region):

| Code | Meaning                     | Example                      |
| ---- | --------------------------- | ---------------------------- |
| `%c` | Date and time representation | `"Fri Nov 14 16:05:30 2025"` |
| `%x` | Date representation         | `"11/14/25"` (varies by locale) |
| `%X` | Time representation         | `"16:05:30"`                 |

```{r}
# Create a Date object
dt <- as.Date("2025-11-14")
# Create a POSIXct object (date-time)
dt <- as.POSIXct("2025-11-14 16:05:30")
# Format examples
format(dt, "%Y-%m-%d")  # e.g. "2025-11-14"
format(dt, "%m/%d/%y")  # e.g. "11/14/25"
format(dt, "%B %d, %Y") # e.g. "November 14, 2025"
format(dt, "%c")  # e.g. "Fri Nov 14 16:05:30 2025"
format(dt, "%x")  # e.g. "11/14/25" (depends on locale)
format(dt, "%X")  # e.g. "16:05:30"

```


| SAS informat  | R parsing                                |
| ------------- | ---------------------------------------- |
| `ddmmyy10.`   | `as.Date(x, "%d/%m/%Y")`                 |
| `ddmmyyd10.`  | `as.Date(x, "%d-%m-%Y")`                 |
| `ddmmyyp10.`  | `as.Date(x, "%d.%m.%Y")`                 |
| `mmddyy10.`   | `as.Date(x, "%m/%d/%Y")`                 |
| `date9.`      | `as.Date(x, "%d%b%Y")`                   |
| `julian7.`    | `as.Date(strptime(x, "%Y%j"))`           |
| `time8.`      | `as.POSIXct(x, "%H:%M:%S")`              |
| `time10.`     | `as.POSIXct(x, "%I:%M:%S%p")` (if am/pm) |
| `datetime18.` | `as.POSIXct(x, "%d%b%Y:%H:%M:%S")`       |
| `datetime20.` | `as.POSIXct(x, "%d%b%Y:%I:%M:%S%p")`     |

| SAS format    | Meaning                                    | R equivalent                             |
| ------------- | ------------------------------------------ | ---------------------------------------- |
| `worddate18.` | Month date, year (e.g., February 23, 2003) | `format(d, "%B %d, %Y")`                 |
| `weekdate24.` | Weekday, month date, year                  | `format(d, "%A, %B %d, %Y")`             |
| `year.`       | numeric year                               | `as.integer(format(d, "%Y"))`            |
| `month.`      | month number (1–12)                        | `as.integer(format(d, "%m"))`            |
| `day.`        | day of month                               | `as.integer(format(d, "%d"))`            |
| `weekday.`    | day of week (1–7)                          | `as.integer(format(d, "%u"))`  (Mon = 1) |



```{r}
# Convert a character string to a Date object
date_string <- "2023-10-01"
date_object <- as.Date(date_string)
print(date_object)
# Convert a character string to a POSIXct object (date-time)
datetime_string <- "2023-10-01 12:34:56"
datetime_object <- as.POSIXct(datetime_string)
print(datetime_object)
# Extract components from a Date object
year <- format(date_object, "%Y")
month <- format(date_object, "%m")
day <- format(date_object, "%d")
print(paste("Year:", year, "Month:", month, "Day:", day))
# Extract components from a POSIXct object
hour <- format(datetime_object, "%H")
minute <- format(datetime_object, "%M")
second <- format(datetime_object, "%S")
print(paste("Hour:", hour, "Minute:", minute, "Second:", second))
# Perform date arithmetic
tomorrow <- current_date + 1
yesterday <- current_date - 1
print(paste("Tomorrow:", tomorrow, "Yesterday:", yesterday))
# Calculate the difference between two dates
date_diff <- as.numeric(difftime(current_date, date_object, units = "days"))
print(paste("Difference in days:", date_diff))
# Format a Date object to a specific string format
formatted_date <- format(current_date, "%B %d, %Y")
print(formatted_date)
# Format a POSIXct object to a specific string format
formatted_datetime <- format(current_datetime, "%Y-%m-%d %H:%M:%S")
print(formatted_datetime)
```

These functions allow you to create, manipulate, and format date and time data in R. For more advanced date and time handling, you might consider using the `lubridate` package, which provides a more user-friendly interface for working with dates and times.

A practical, CDISC‑friendly guide to working with **dates/times** and implementing **partial date imputation** in R using **base R** and **lubridate**.

## 1. Date & Time Basics in R

### 1.1 Core classes

-   **Date**: calendar dates without time.
-   **POSIXct**: seconds since 1970‑01‑01 (compact, good for storage/arithmetic).
-   **POSIXlt**: list of components (year, mon, mday, hour, min, sec) — convenient to extract parts.


### 1.2 Converting numerics to Date (origins matter)

Different systems use different **origins** for numeric dates: - Excel: 1900‑01‑01 (beware 1900 leap‑year bug in older Excels) - SAS: 1960‑01‑01 - R: 1970‑01‑01

``` {r}
num_dates <- c(0, 159, 464, 15674)

as.Date(num_dates, origin = "1960-01-01")  # SAS origin
as.Date(num_dates, origin = "1970-01-01")  # R origin
as.Date(num_dates, origin = "1900-01-01")  # Excel origin
```
### 1.3 Custom date formats

``` r
x1 <- c("01/05/1965", "08/16/1975")
as.Date(x1, "%m/%d/%Y")

x2 <- c("05-01-1965", "16-08-1975")
as.Date(x2, "%d-%m-%Y")

x3 <- c("01MAY1965", "16AUG1975")
as.Date(x3, "%d%b%Y")

# Unknown format? Try multiple patterns
x4 <- c("01MAY1965MORNING", "16AUGUST1975")
as.Date(x4, tryFormats = c("%d%b%Y", "%d%B%Y"))
```

### 1.4 Datetime strings (ISO 8601 & time zones)

``` r
x <- c("2015-10-19T10:15", "2018-12-19T23:05")
# Keep the literal 'T' in the format string
as.POSIXct(x, format = "%Y-%m-%dT%H:%M", tz = "UTC")

# Extract parts
xt <- as.POSIXlt(x, format = "%Y-%m-%dT%H:%M")
format(xt, "%Y-%m-%d")  # date part
format(xt, "%H:%M")     # time part
```

## 2. Lubridate: Fast, Friendly Parsing & Arithmetic

Install & load: `install.packages("lubridate"); library(lubridate)`

### 2.1 Intuitive parsers

-   `ymd()`, `mdy()`, `dmy()` for dates
-   `ymd_hms()`, `ymd_hm()`, `ymd_h()` for date‑times

``` r
ymd("20210529")
dmy("29/05/2021")
ymd_hm("2019/01/19T17:15")  # delimiter agnostic
```

### 2.2 Accessors and rounding

``` r
x <- ymd_hms("2009-08-03 12:01:59")
year(x); month(x); mday(x); wday(x); hour(x); minute(x); second(x)

floor_date(x, "month")   # 2009-08-01
ceiling_date(x, "month") # 2009-09-01
round_date(x, "month")    # 2009-08-01
rollback(x)               # last day of previous month
```

### 2.3 Timespans

-   **Durations**: exact seconds, e.g., `ddays(2)`
-   **Periods**: human units, e.g., `months(1)`
-   **Intervals**: start–end pair, e.g., `interval(start, end)`

``` r
start <- ymd("2015-10-19")
end   <- ymd("2018-12-19")
int   <- interval(start, end)
as.duration(int)       # exact seconds
as.period(int)         # years/months/days
```


### 3 ADaM‑Style Partial Date Imputation (Start & End)
 Imputation rules vary by study; always follow the SAP. A common pattern:

### Start date (`AESTDTC`)

- `YYYY-MM-DD` → use as is  
- `YYYY-MM` → impute **day = 01** (first of month)  
- `YYYY` → impute **month = 01**, **day = 01** (`01-Jan`)

### End date (`AEENDTC`)

- `YYYY-MM-DD` → use as is  
- `YYYY-MM` → impute **last day of that month**  
- `YYYY` → impute **`31-Dec`**

### Censoring by death / last alive date

Additionally, bound `AEENDT` by death / last alive date (when present):

- `CENSOR = coalesce(DTHDT, LASTALVDT)`  
- `AEENDT = min(imputed_end, CENSOR)`

### Imputation flags

Add flags:

- `AESTDTF`, `AEENDTF`:
  - `"D"`   → day imputed  
  - `"M,D"` → month & day imputed  
- Add `"C"` if we had to **censor the end date** at `DTHDT` / `LASTALVDT`.


```{r}
library(dplyr, warn.conflicts = FALSE)
library(stringr, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)

df <- data.frame(
  USUBJID    = sprintf("01-001-%03d", 1:6),
  AESTDTC    = c("2015-10-19", "2015-10", "2015", "2019-02", "2020-02", "2018-12-31"),
  AEENDTC    = c("2018-12-19", "2018-10", "2017-05-29", "2019-02", "2020-02", "2020"),
  RANDDT     = ymd(c("2015-10-01", "2015-10-01", "2015-01-10", "2019-01-15", "2020-02-10", "2018-12-01")),
  DTHDT      = ymd(c(NA, NA, NA, NA, "2020-08-03", NA)),
  LASTALVDT  = ymd(c(NA, NA, NA, NA, NA, "2020-12-15"))
)

df_imp <- df |> 
  mutate(
    # Length of partial date strings
    AESTDTC_LEN = str_length(AESTDTC),
    AEENDTC_LEN = str_length(AEENDTC),
    ## --- AE start date imputation (minimum: first possible date) ---
    AESTDT = case_when(
      is.na(AESTDTC) ~ as.Date(NA),
      AESTDTC_LEN == 10 ~ ymd(AESTDTC, quiet = TRUE),
      AESTDTC_LEN == 7  ~ ymd(str_c(AESTDTC, "-01"), quiet = TRUE),
      AESTDTC_LEN == 4  ~ ymd(str_c(AESTDTC, "-01-01"), quiet = TRUE),
      TRUE ~ as.Date(NA)
    ),
    AESTDTF = case_when(
      is.na(AESTDTC) ~ NA_character_,
      AESTDTC_LEN == 10 ~ "",
      AESTDTC_LEN == 7  ~ "D",
      AESTDTC_LEN == 4  ~ "M,D",
      TRUE ~ "UNK"
    ),
    ## --- AE end date imputation (maximum: latest possible date) ---
    AEENDT_RAW = case_when(
      is.na(AEENDTC) ~ as.Date(NA),
      AEENDTC_LEN == 10 ~ ymd(AEENDTC, quiet = TRUE),
      AEENDTC_LEN == 7  ~ ymd(str_c(AEENDTC, "-01"), quiet = TRUE),
      AEENDTC_LEN == 4  ~ ymd(str_c(AEENDTC, "-12-31"), quiet = TRUE),
      TRUE ~ as.Date(NA)
    ),
    AEENDTF_RAW = case_when(
      is.na(AEENDTC) ~ NA_character_,
      AEENDTC_LEN == 10 ~ "",
      AEENDTC_LEN == 7  ~ "D",
      AEENDTC_LEN == 4  ~ "M,D",
      TRUE ~ "UNK"
    )
  ) %>%
  mutate(
    AEENDT_RAW = if_else(
      AEENDTC_LEN == 7 & !is.na(AEENDT_RAW),
      ceiling_date(AEENDT_RAW, "month") - days(1),
      AEENDT_RAW
    ),
    CENSOR = coalesce(DTHDT, LASTALVDT),
    AEENDT = case_when(
      !is.na(CENSOR) & !is.na(AEENDT_RAW) & AEENDT_RAW > CENSOR ~ CENSOR,
      TRUE ~ AEENDT_RAW
    ),
    AEENDTF = case_when(
      is.na(AEENDT) ~ AEENDTF_RAW,
      !is.na(CENSOR) & !is.na(AEENDT_RAW) & AEENDT_RAW > CENSOR & AEENDTF_RAW == ""  ~ "C",
      !is.na(CENSOR) & !is.na(AEENDT_RAW) & AEENDT_RAW > CENSOR & AEENDTF_RAW != "" ~ str_c(AEENDTF_RAW, ",C"),
      TRUE ~ AEENDTF_RAW
    )
  )


```

## 4. Durations & Analysis Variables

### 4.1 AE duration (AEDUR) in days

# Using difftime (base) or as.numeric on Date differences
`AEDUR <- as.integer(out$AEENDT - out$ASTDT)`

### 6.2 Handy lubridate parsers (subset)

-   Dates: `ymd()`, `mdy()`, `dmy()`, `yq()`\
-   Datetimes: `ymd_hms()`, `ymd_hm()`, `ymd_h()`\
-   Times: `hms()`, `hm()`, `ms()`\
-   Accessors: `year()`, `month()`, `mday()`, `wday()`, `hour()`, `minute()`, `second()`, `date()`\
-   Rounding: `floor_date()`, `ceiling_date()`, `round_date()`, `rollback()`

------------------------------------------------------------------------
