---
title: "Reading SAS Datasets "
---
# Reading SAS Datasets and Working with XPT Files

## Introduction

When working with SAS datasets in R, you'll often need to read various SAS file formats, write XPT (transport) files for regulatory submissions, and manage dataset and variable attributes. This chapter covers the essential tools and techniques for these tasks.

## Required Packages

```{r}
#| label: setup
#| message: false
#| warning: false

library(haven)      # For reading/writing SAS files
library(dplyr)      # For data manipulation
library(labelled)   # For working with labels
library(sjlabelled) # Additional label functions
library(foreign)    # Alternative for XPT files

```

## Reading SAS Datasets

### Reading SAS7BDAT Files

The `haven` package is the most reliable way to read modern SAS datasets:

```{r}
#| label: read-sas7bdat
#| eval: false

# Read a SAS dataset
data <- read_sas("path/to/dataset.sas7bdat")

# Read with specific encoding
data <- read_sas("path/to/dataset.sas7bdat", encoding = "latin1")

# Read and specify column types
data <- read_sas("path/to/dataset.sas7bdat", 
                 col_types = cols(
                   subjid = col_character(),
                   age = col_double(),
                   visit = col_factor()
                 ))
```

### Reading XPT (Transport) Files

XPT files are commonly used in pharmaceutical submissions:

```{r}
#| label: read-xpt
#| eval: false

# Using haven (recommended)
data <- read_xpt("path/to/dataset.xpt")

# Using foreign package
data <- read.xport("path/to/dataset.xpt")

# Using SASxport package for more control
data <- read.xport("path/to/dataset.xpt", verbose = TRUE)
```

### Handling Different SAS Formats

```{r}
#| label: read-formats
#| eval: false

# Read dataset with format catalog
data <- read_sas("dataset.sas7bdat", 
                 catalog = "formats.sas7bcat")

# Read older SAS transport files
data <- read_xpt("old_dataset.xpt")

# Handle missing values properly
data <- read_sas("dataset.sas7bdat", 
                 user_na = c(".", "", " "))
```

## Working with Labels and Attributes

### Understanding SAS Labels in R

When reading SAS datasets, labels are preserved as attributes:

```{r}
#| label: examine-labels
#| eval: false

# Check if data has labels
has_labels(data)

# View all variable labels
var_label(data)

# View label for specific variable
var_label(data$age)

# View dataset label
attr(data, "label")
```

### Adding and Modifying Labels

```{r}
#| label: add-labels
#| eval: false

# Add variable labels
data <- data %>%
  set_variable_labels(
    subjid = "Subject Identifier",
    age = "Age at Baseline (years)",
    sex = "Gender",
    visit = "Study Visit",
    weight = "Body Weight (kg)",
    height = "Height (cm)"
  )

# Add dataset label
attr(data, "label") <- "Demographics Dataset"

# Alternative using labelled package
data$age <- labelled(data$age, label = "Age at Baseline (years)")

# Add value labels (for categorical variables)
data$sex <- labelled(data$sex,
                     labels = c("Male" = "M", "Female" = "F"),
                     label = "Gender")
```

### Working with Formats

```{r}
#| label: formats
#| eval: false

# Apply format to a variable
data$visit <- factor(data$visit, 
                     levels = c(1, 2, 3, 4),
                     labels = c("Screening", "Baseline", "Week 12", "Follow-up"))

# Create custom format function
format_visit <- function(x) {
  case_when(
    x == 1 ~ "Screening",
    x == 2 ~ "Baseline", 
    x == 3 ~ "Week 12",
    x == 4 ~ "Follow-up",
    TRUE ~ as.character(x)
  )
}

data$visit_formatted <- format_visit(data$visit)
```

## Writing XPT Files

### Basic XPT File Creation

```{r}
#| label: write-xpt-basic
#| eval: false

# Write XPT file using haven
write_xpt(data, "output_dataset.xpt")

# Write with specific options
write_xpt(data, "output_dataset.xpt", 
          version = 8,  # SAS version
          name = "DEMOG") # Dataset name in XPT

# Write multiple datasets to separate XPT files
datasets <- list(
  demog = demographics,
  vital = vital_signs,
  lab = laboratory
)

iwalk(datasets, ~write_xpt(.x, paste0(.y, ".xpt")))
```

### Advanced XPT Writing with SASxport

```{r}
#| label: write-xpt-advanced
#| eval: false

# Prepare data for XPT with proper formatting
data_for_xpt <- data %>%
  # Ensure character variables are properly sized
  mutate(
    subjid = str_pad(subjid, width = 10, side = "right"),
    sex = str_pad(sex, width = 1, side = "right")
  ) %>%
  # Convert dates to SAS date format
  mutate(
    randdt = as.numeric(as.Date(randdt) - as.Date("1960-01-01"))
  )

# Write with SASxport package for more control
write.xport(list(DEMOG = data_for_xpt), 
            file = "demog.xpt",
            verbose = TRUE,
            sasVer = "7.00",
            osType = "WIN_PRO")
```

### Handling Variable Names and Lengths

```{r}
#| label: xpt-constraints
#| eval: false

# Ensure variable names comply with SAS naming rules
prepare_for_xpt <- function(data) {
  data %>%
    # Rename variables to meet SAS constraints
    rename_with(~str_to_upper(.)) %>%  # Uppercase
    rename_with(~str_sub(., 1, 8)) %>%  # Max 8 characters
    rename_with(~str_replace_all(., "[^A-Z0-9_]", "")) %>%  # Valid characters only
    # Ensure character variables are not too long
    mutate(across(where(is.character), ~str_trunc(., width = 200)))
}

data_xpt_ready <- prepare_for_xpt(data)
write_xpt(data_xpt_ready, "dataset.xpt")
```

## Adding Dataset Attributes

### Dataset-Level Attributes

```{r}
#| label: dataset-attributes
#| eval: false

# Add comprehensive dataset attributes
add_dataset_attributes <- function(data, 
                                   label = NULL, 
                                   creation_date = Sys.Date(),
                                   creator = Sys.info()["user"],
                                   description = NULL) {
  
  # Add standard attributes
  attr(data, "label") <- label
  attr(data, "creation.date") <- as.character(creation_date)
  attr(data, "created.by") <- creator
  attr(data, "description") <- description
  
  # Add CDISC-style attributes for regulatory compliance
  attr(data, "SAS.dataset.label") <- label
  attr(data, "SAS.dataset.name") <- deparse(substitute(data))
  
  return(data)
}

# Apply dataset attributes
demographics <- add_dataset_attributes(
  demographics,
  label = "Subject Demographics",
  description = "Baseline demographic and background characteristics",
  creator = "Clinical Data Management"
)
```

### Variable-Level Attributes

```{r}
#| label: variable-attributes
#| eval: false

# Function to add comprehensive variable attributes
add_variable_attributes <- function(data, var_specs) {
  
  for (var_name in names(var_specs)) {
    if (var_name %in% names(data)) {
      
      spec <- var_specs[[var_name]]
      
      # Add label
      if (!is.null(spec$label)) {
        attr(data[[var_name]], "label") <- spec$label
      }
      
      # Add format
      if (!is.null(spec$format)) {
        attr(data[[var_name]], "format.sas") <- spec$format
      }
      
      # Add derivation comment
      if (!is.null(spec$derivation)) {
        attr(data[[var_name]], "derivation") <- spec$derivation
      }
      
      # Add value labels for categorical variables
      if (!is.null(spec$values)) {
        data[[var_name]] <- labelled(data[[var_name]], 
                                     labels = spec$values,
                                     label = spec$label)
      }
    }
  }
  
  return(data)
}

# Define variable specifications
demog_specs <- list(
  SUBJID = list(
    label = "Subject Identifier",
    format = "$10."
  ),
  AGE = list(
    label = "Age at Randomization",
    format = "3.",
    derivation = "Calculated from birth date and randomization date"
  ),
  SEX = list(
    label = "Gender", 
    format = "$1.",
    values = c("Male" = "M", "Female" = "F")
  ),
  RACE = list(
    label = "Race",
    format = "$20.",
    values = c(
      "White" = "WHITE",
      "Black or African American" = "BLACK", 
      "Asian" = "ASIAN",
      "Other" = "OTHER"
    )
  )
)

# Apply variable attributes
demographics <- add_variable_attributes(demographics, demog_specs)
```

## Quality Checks and Validation

### Validating XPT Files

```{r}
#| label: validate-xpt
#| eval: false

# Function to validate XPT file compliance
validate_xpt_compliance <- function(data) {
  
  issues <- list()
  
  # Check variable names
  var_names <- names(data)
  invalid_names <- var_names[!str_detect(var_names, "^[A-Z][A-Z0-9_]{0,7}$")]
  if (length(invalid_names) > 0) {
    issues$invalid_names <- invalid_names
  }
  
  # Check character variable lengths
  char_vars <- data %>% select(where(is.character))
  long_vars <- char_vars %>%
    summarise(across(everything(), ~max(nchar(.x), na.rm = TRUE))) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "max_length") %>%
    filter(max_length > 200)
  
  if (nrow(long_vars) > 0) {
    issues$long_character_vars <- long_vars
  }
  
  # Check for missing labels
  unlabeled_vars <- var_names[sapply(data, function(x) is.null(attr(x, "label")))]
  if (length(unlabeled_vars) > 0) {
    issues$unlabeled_vars <- unlabeled_vars
  }
  
  return(issues)
}

# Validate before writing
validation_issues <- validate_xpt_compliance(demographics)
if (length(validation_issues) == 0) {
  cat("Dataset passes XPT compliance checks\n")
  write_xpt(demographics, "demographics.xpt")
} else {
  cat("Validation issues found:\n")
  print(validation_issues)
}
```

### Creating a Data Dictionary

```{r}
#| label: data-dictionary
#| eval: false

# Function to create a data dictionary
create_data_dictionary <- function(data) {
  
  dict <- data.frame(
    Variable = names(data),
    Label = sapply(data, function(x) attr(x, "label") %||% ""),
    Type = sapply(data, class),
    Length = sapply(data, function(x) {
      if (is.character(x)) max(nchar(x), na.rm = TRUE) else NA
    }),
    Format = sapply(data, function(x) attr(x, "format.sas") %||% ""),
    stringsAsFactors = FALSE
  )
  
  # Add value labels for categorical variables
  dict$Values <- sapply(names(data), function(var) {
    x <- data[[var]]
    if (is.labelled(x)) {
      labels <- attr(x, "labels")
      if (!is.null(labels)) {
        paste0(names(labels), "=", labels, collapse = "; ")
      } else ""
    } else ""
  })
  
  return(dict)
}

# Create and export data dictionary
data_dict <- create_data_dictionary(demographics)
write.csv(data_dict, "demographics_dictionary.csv", row.names = FALSE)
```

## Best Practices

### Recommended Workflow

1. **Read SAS data** with appropriate encoding and column types
2. **Validate data integrity** after reading
3. **Add comprehensive labels** for variables and datasets
4. **Apply proper formatting** for categorical variables
5. **Validate XPT compliance** before writing
6. **Create documentation** (data dictionary)
7. **Write XPT files** with proper attributes

### Performance Tips

```{r}
#| label: performance-tips
#| eval: false

# For large datasets, read in chunks
read_large_sas <- function(file_path, chunk_size = 10000) {
  # Use data.table for faster processing
  library(data.table)
  
  # Read metadata first
  meta <- read_sas(file_path, n_max = 1)
  
  # Process in chunks
  results <- list()
  skip <- 0
  
  repeat {
    chunk <- read_sas(file_path, skip = skip, n_max = chunk_size)
    if (nrow(chunk) == 0) break
    
    # Process chunk here
    results[[length(results) + 1]] <- chunk
    skip <- skip + chunk_size
  }
  
  # Combine results
  do.call(rbind, results)
}

# Use fst format for intermediate storage (faster than SAS)
library(fst)
write_fst(data, "intermediate_data.fst")
data <- read_fst("intermediate_data.fst")
```

## Summary

This chapter covered:

- Reading various SAS file formats (SAS7BDAT, XPT)
- Working with variable and dataset labels
- Adding comprehensive attributes for regulatory compliance
- Writing XPT files with proper formatting
- Validation and quality checks
- Best practices for performance and documentation

These techniques form the foundation for working with SAS datasets in R, particularly in regulated environments where proper documentation and file formats are critical.