---
title: "Custom Functions & Validation"
---


# Function introduction

In R, a **function** is a reusable block of code that performs a specific task.  
Functions help you:

- Avoid repeating the same code again and again
- Organize code into logical pieces
- Make your code easier to read, test, and maintain

R has many **built-in functions** (e.g., `sum()`, `mean()`, `lm()`), and you can also define your **own functions**.

```{r}
# Example: using a built-in function
x <- c(1, 2, 3, 4, 5)
mean(x)
```

## Function components
A function has three parts: 

* The `formals()`, the list of arguments that control how you call the function.
  
* The `body()`, the code inside the function.

* The `environment()`, the data structure that determines how the function finds 
the values associated with the names.

While the formals and body are specified explicitly when you create a function, the environment is specified implicitly, based on _where_ you defined the function. The function environment always exists, but it is only printed when the function isn't defined in the global environment.

```{r}
f02 <- function(x, y) {
  # A comment
  x + y
}

formals(f02)

body(f02)

environment(f02)
```

## Signalling conditions


There are three conditions that you can signal in code: errors, warnings, and messages. 

* Errors are the most severe; they indicate that there is no way for a function 
  to continue and execution must stop. 

* Warnings fall somewhat in between errors and message, and typically indicate 
  that something has gone wrong but the function has been able to at least 
  partially recover. 
  
* Messages are the mildest; they are way of informing users that some action 
  has been performed on their behalf. 
  
There is a final condition that can only be generated interactively: an interrupt, which indicates that the user has interrupted execution by pressing Escape, Ctrl + Break, or Ctrl + C (depending on the platform).

Conditions are usually displayed prominently, in a bold font or coloured red, depending on the R interface. You can tell them apart because errors always start with "Error", warnings with "Warning" or "Warning message", and messages with nothing.

```{r, eval=FALSE}
stop("This is what an error looks like")

warning("This is what a warning looks like")

message("This is what a message looks like")
```

## Ignoring conditions
You can choose to ignore conditions using `try()`, tryCatch()` or `suppressMessages()`, `suppressWarnings()`.

```{r, eval=FALSE}
# Example: ignoring a warning
suppressWarnings(log(-1))

suppressWarnings({
  warning("Uhoh!")
  warning("Another warning")
  1
})


suppressMessages({
  message("Hello there")
  2
})


suppressWarnings({
  message("You can still see me")
  3
})

suppressWarnings(library(dplyr))
suppressWarnings(library(ggplot2))
```
## tryCatch function
The `tryCatch()` function allows you to handle conditions (errors, warnings, messages) in a controlled way. You can specify custom behavior for each type of condition.
```{r}
result <- tryCatch({
  # Code that may produce an error
  log(-1)
},
warning = function(w) {
  cat("Caught a warning:", conditionMessage(w), "\n")
  NA  # Return NA on warning
},
error = function(e) {
  cat("Caught an error:", conditionMessage(e), "\n")
  NA  # Return NA on error
},
finally = {
  cat("Execution completed.\n")
})
result
```

## function scope
Functions in R have their own scope, meaning that variables defined inside a function are not accessible outside of it, and vice versa. This helps prevent naming conflicts and keeps functions self-contained.
```{r}
my_function <- function() {
  x <- 10  # x is local to the function
  return(x)
}
my_function()

```
## Creating Custom Functions
```{r}
# Define a simple function to add two numbers
add_numbers <- function(a, b) {
  return(a + b)
}
# Call the function
add_numbers(3, 5)
```
## Function Arguments
```{r}
# Function with default argument
greet <- function(name = "Guest") {
  paste("Hello,", name)
}
greet()          # Uses default
greet("Alice")   # Overrides default
```


## User-defined Functions in R Programming Language

User-defined functions are the functions that are created by the user. User defines the working, parameters, default parameter, etc. of that user-defined function. They can be only used in that specific code.

    1.  Single Input Single Output : Now create a function in R that will take a single input and gives us a single output
    2.  Multiple Input Multiple Output : Now create a function in R Language that will take multiple inputs and gives us multiple outputs using a list.

```{r}
# Single Input Single Outpu
calculate_bmi <- function(data) {
  # BMI formula: weight (kg) / height (m)^2
  bmi <- data$weight / (data$height ^ 2)
  return(bmi)
}

# Example data
patient_data <- data.frame(
  weight = c(70, 80, 65),  # in kg
  height = c(1.75, 1.8, 1.6)  # in meters
)

# Calculate BMI
bmi_values <- calculate_bmi(patient_data)
print(bmi_values)  # Outputs BMI for each patient

# Multiple Input Multiple Output
calculate_summary_stats <- function(age, weight, height) {
  age_stats <- list(
    mean_age = mean(age, na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE)
  )
  
  weight_stats <- list(
    mean_weight = mean(weight, na.rm = TRUE),
    median_weight = median(weight, na.rm = TRUE),
    sd_weight = sd(weight, na.rm = TRUE)
  )
  
  height_stats <- list(
    mean_height = mean(height, na.rm = TRUE),
    median_height = median(height, na.rm = TRUE),
    sd_height = sd(height, na.rm = TRUE)
  )
  
  # Return a list containing all summary stats
  return(list(age = age_stats, weight = weight_stats, height = height_stats))
}

# Example data
age <- c(40, 45, 50, 55, 60)
weight <- c(70, 75, 65, 80, 68)
height <- c(1.75, 1.8, 1.6, 1.65, 1.7)

summary_stats <- calculate_summary_stats(age, weight, height)
print(summary_stats)  # Outputs summary stats for age, weight, and height

```

What is `{{}}` in R?

The `{{` syntax is known as curly-curly syntax, and it is part of the tidy evaluation framework in the {r, lang} package. It’s used to capture variables directly and allows for easy manipulation within functions.

```{r}
library(dplyr)
library(tidyr)


stat <- function(data, grp, var) {
  # Group by specified grouping variables and summarize the statistics
  summary_stats <- data |> group_by(across({{ grp }})) |> summarise(
      n = as.character(n_distinct(USUBJID)),
      mean = as.character(round(mean({{ var }}, na.rm = TRUE), 2)),
      median = as.character(median({{ var }}, na.rm = TRUE)),
      sd = as.character(round(sd({{ var }}, na.rm = TRUE), 3)),
      min = as.character(min({{ var }}, na.rm = TRUE)),
      max = as.character(max({{ var }}, na.rm = TRUE)),
      .groups = 'drop'  # Ungroup after summarising
    )
  
  # Pivot to wider format
transposed_data <- summary_stats |> 
  pivot_longer(
    cols =c(n, mean, median, sd, min, max),        # Columns to pivot
    names_to = "stat",        # Name for the new 'stat' column
    values_to = "value"       # Name for the new 'value' column
  ) |> pivot_wider(
    names_from = {{ grp }},     # Treatment groups to become column names
    values_from = value       # Values for each statistic
  )
  
  return(transposed_data)
}

adsl <- haven::read_sas("./data/adam/adsl.sas7bdat")
age_stat <- stat(data=adsl,grp=c(TRT01A,TRT01AN),var=AGE)
wgt_stat <- stat(adsl,TRT01A,WEIGHTBL)

```

# difference !,!!,!!!
In R, the symbols `!`, `!!`, and `!!!` are used in different contexts, primarily in logical operations and tidy evaluation.
- `!` (Logical NOT): This is a logical operator used to negate a logical value. It takes a single logical value (TRUE or FALSE) and returns the opposite value. For example, `!TRUE` returns `FALSE`, and `!FALSE` returns `TRUE`.
- `!!` (Unquote): In the context of tidy evaluation (used in packages like dplyr), `!!` is used to unquote a single variable or expression. It allows you to evaluate a variable or expression within a tidy evaluation context. For example, if you have a variable `var_name` that holds the name of a column, you can use `!!var_name` to refer to that column in a dplyr function.
- `!!!` (Unquote-splice): This is also used in tidy evaluation and is used to unquote and splice a list of variables or expressions. It allows you to pass multiple arguments to a function by splicing a list. For example, if you have a list of column names, you can use `!!!` to pass them as individual arguments to a dplyr function.


```{r}
stat_simple <- function(data, grp, var, id_var = "USUBJID") {
  # Convert strings to symbols
  grp_sym <- sym(grp)
  var_sym <- sym(var)
  id_var_sym <- sym(id_var)
  
  # Group by specified grouping variables and summarize the statistics
  summary_stats <- data |> 
    group_by(!!grp_sym) |> 
    summarise(
      n = as.character(n_distinct(!!id_var_sym)),
      mean = as.character(round(mean(!!var_sym, na.rm = TRUE), 2)),
      median = as.character(round(median(!!var_sym, na.rm = TRUE), 2)),
      sd = as.character(round(sd(!!var_sym, na.rm = TRUE), 3)),
      min = as.character(round(min(!!var_sym, na.rm = TRUE), 2)),
      max = as.character(round(max(!!var_sym, na.rm = TRUE), 2)),
      .groups = 'drop'
    )
  
  # Pivot to longer format then wider
  transposed_data <- summary_stats |> 
    pivot_longer(
      cols = c(n, mean, median, sd, min, max),
      names_to = "stat",
      values_to = "value"
    ) |> 
    pivot_wider(
      names_from = !!grp_sym,
      values_from = value
    ) |> 
    arrange(factor(stat, levels = c("n", "mean", "median", "sd", "min", "max")))
  
  return(transposed_data)
}

adsl <- haven::read_sas("./data/adam/adsl.sas7bdat")
age_stat <- stat_simple(data=adsl,grp="TRT01A",var="AGE")
```

```{r}
#where function
 adsl1 <- adsl |> 
   dplyr::select(dplyr::where(is.character))
 
adsl2 <- adsl |> 
   dplyr::select(dplyr::where(is.numeric))

# across functions structure and explanation

#across(.cols, .fns, ..., .names = NULL, .unpack = FALSE)

df <- data.frame(
  id = 1:6,
  age = c(21, 34, 28, NA, 45, 30),
  wt = c(70, 80, 65, 72, NA, 68),
  sex = c("M","F","F","M","F","M")
)

# 2) Summaries grouped by sex
df2 <- df |>
  group_by(sex) |> 
  summarise(across(.cols=dplyr::where(is.numeric), 
                   .fns=list( mean = ~mean(., na.rm=TRUE),
                              sd   = ~sd(., na.rm=TRUE)
                             ),
                   .names = "{.col}_{.fn}")
  )

df_qc <- df |> 
  group_by(sex) |> 
  summarize(mm=mean(age,na.rm=TRUE),
            mn1=mean(wt,na.rm=TRUE))

df2 <- df |>
  group_by(sex) |> 
  summarise(across(.cols=c(age, wt), 
                   .fns=list(nrecs  = ~dplyr::n(),
                             mean = ~mean(., na.rm=TRUE),
                             sd   = ~sd(., na.rm=TRUE)
                   ),
                   .names = "{.col}_{.fn}")
  )

df2 <- df |>
  group_by(sex) |> 
  summarise(across(.cols=c(age, wt), 
                   .fns= ~mean(., na.rm=TRUE))
  )

a <- c(1, 2, 3, 4, 5, 9909)
b <- data.frame(
  var1 = c(1, 2, 3, 4, 5, 9909),
  var2 = c(5, 6, 7, 8, 9, 9909)
)
c <- list(a,b)


# Numeric: replace sentinel 9909 with 0
numericdata <- tibble::tribble(
  ~var1, ~var2, ~var3, ~var4, ~var5,
  9909,   1,     2,     3,     4,
  5,    9909,    6,     7,     8,
  9,      3,   9909,    2,     7
)
dm <- haven::read_sas("./data/sdtm/dm.sas7bdat") |> 
  dplyr::mutate(dplyr::across(where(is.character),~ dplyr::na_if(.,"")))

numericdata01 <- numericdata |>
  dplyr::mutate(dplyr::across(.cols=where(is.numeric), .fns=~ dplyr::if_else(.x == 9909, 0, .x)))


# multiply all numeric variables by 2
adsl3 <- adsl |> 
  dplyr::mutate(
    dplyr::across(.cols = where(is.numeric),
                  .fns = ~ .x * 2)
  )

```


# Error Handling
```{r}
robust_divide <- function(a, b) {
  tryCatch(a / b,
           warning = function(w) NA_real_,
           error   = function(e) NA_real_)
}
robust_divide(10, 0)
```

# Unit Testing with testthat
Install once: `install.packages(c("testthat","devtools","usethis","roxygen2"))`

```{r eval=FALSE}
usethis::use_testthat()
usethis::use_test("safe_mean")
```

Create `tests/testthat/test-safe_mean.R`:
```{r eval=FALSE}
testthat::test_that("safe_mean works", {
  x <- c(1,2,NA)
  testthat::expect_equal(safe_mean(x), 1.5)
  testthat::expect_error(safe_mean("oops"))
})
```

# Document with roxygen2
```{r}
#' Compute a safe mean
#' @param x Numeric vector
#' @param na.rm Logical; remove NAs
#' @return Numeric scalar
#' @examples
#' safe_mean(c(1,2,NA))
#' @export
safe_mean <- function(x, na.rm = TRUE) {
  stopifnot(is.numeric(x))
  mean(x, na.rm = na.rm)
}
```
Run:
```{r eval=FALSE}
devtools::document()
```

# Exercises
1. Write `winsorize(x, probs=c(0.05,0.95))` and test it.
2. Create `validate_columns(df, required=c("USUBJID","AGE"))` and add tests.
3. Add roxygen docs and build help pages.
