---
title: "Custom Functions & Validation"
---

# Introduction to Functions in R

In R, a **function** is a reusable block of code that performs a specific task.  Functions are fundamental building blocks that help you: 

- Avoid repeating the same code multiple times (DRY principle)
- Organize code into logical, manageable pieces
- Make your code easier to read, test, and maintain
- Create modular, scalable programs
- Enable code sharing and collaboration

R has thousands of **built-in functions** (e.g., `sum()`, `mean()`, `lm()`, `ggplot()`), and you can also define your **own functions** to solve specific problems.

```{r}
# Example: using built-in functions
numbers <- c(1, 2, 3, 4, 5, NA)
sum(numbers, na.rm = TRUE)
mean(numbers, na.rm = TRUE)
length(numbers)
```

# Function Fundamentals

## Anatomy of a Function

Every function in R has three essential components:

1. **The `formals()`**: The list of arguments that control how you call the function
2. **The `body()`**: The code inside the function that does the actual work
3. **The `environment()`**: The data structure that determines how the function finds values

```{r}
# Create a simple function
my_function <- function(x, y = 2) {
  # This is the body
  result <- x * y + 1
  return(result)
}

# Examine the components
formals(my_function)    # Shows the arguments
body(my_function)       # Shows the code
environment(my_function) # Shows the environment
```

## Basic Function Syntax

The general syntax for creating a function is:

```{r, eval=FALSE}
function_name <- function(arg1, arg2 = default_value, ...) {
  # Function body
  # Your code here
  return(result)  # Optional explicit return
}
```

### Simple Examples

```{r}
# 1. Function with no arguments
say_hello <- function() {
  return("Hello, World!")
}

say_hello()

# 2. Function with one argument
square <- function(x) {
  return(x^2)
}

square(4)
square(c(1, 2, 3, 4))

# 3. Function with multiple arguments
add_numbers <- function(a, b) {
  sum_result <- a + b
  return(sum_result)
}

add_numbers(3, 5)
add_numbers(c(1,2), c(3,4))
```

## Function Arguments and Parameters

### Default Arguments

```{r}
# Function with default arguments
greet <- function(name = "Guest", greeting = "Hello") {
  message <- paste(greeting, name, "!")
  return(message)
}

greet()                           # Uses all defaults
greet("Alice")                    # Uses default greeting
greet("Bob", "Hi")               # Overrides both defaults
greet(greeting = "Hey", name = "Charlie")  # Named arguments
```

### The `...` (Dots) Parameter

The `...` allows functions to accept any number of arguments:

```{r}
# Function using dots to pass arguments to other functions
my_summary <- function(x, ...) {
  cat("Length:", length(x), "\n")
  cat("Mean:", mean(x, ...), "\n")
  cat("SD:", sd(x, ...), "\n")
}

data_with_na <- c(1, 2, 3, NA, 5)
my_summary(data_with_na, na.rm = TRUE)

# Function that combines multiple vectors
combine_all <- function(... ) {
  all_args <- list(...)
  combined <- unlist(all_args)
  return(combined)
}

combine_all(1:3, 4:6, c(7, 8))
```

## Return Values

### Explicit vs Implicit Returns

```{r}
# Explicit return
explicit_return <- function(x) {
  result <- x * 2
  return(result)  # Explicitly return the value
}

# Implicit return (last expression is returned)
implicit_return <- function(x) {
  result <- x * 2
  result  # Last expression is automatically returned
}

explicit_return(5)
implicit_return(5)

# Multiple possible returns
check_number <- function(x) {
  if (x > 0) {
    return("Positive")
  } else if (x < 0) {
    return("Negative")
  } else {
    return("Zero")
  }
}

check_number(5)
check_number(-3)
check_number(0)
```

### Returning Multiple Values

```{r}
# Return multiple values using a list
calculate_stats <- function(x) {
  stats_list <- list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    n = length(x[!is.na(x)])
  )
  return(stats_list)
}

sample_data <- c(1, 2, 3, 4, 5, NA, 7, 8, 9, 10)
results <- calculate_stats(sample_data)
print(results)

# Access individual elements
results$mean
results$sd
```

# Intermediate Function Concepts

## Function Scope and Environments

Functions create their own local environment, separate from the global environment:

```{r}
# Global variable
global_var <- 100

scope_demo <- function() {
  local_var <- 50        # Local to this function
  global_var <- 200      # This creates a NEW local variable
  
  cat("Inside function - local_var:", local_var, "\n")
  cat("Inside function - global_var:", global_var, "\n")
  
  return(local_var + global_var)
}

result <- scope_demo()
cat("Outside function - global_var:", global_var, "\n")
# cat("Outside function - local_var:", local_var, "\n")  # This would cause an error
```

### Lexical Scoping

R uses lexical scoping - functions look for variables in the environment where they were defined:

```{r}
# Demonstrate lexical scoping
x <- 10

make_function <- function(multiplier) {
  # This function "captures" the multiplier value
  multiply_by <- function(y) {
    return(y * multiplier * x)  # Uses multiplier from parent and x from global
  }
  return(multiply_by)
}

multiply_by_5 <- make_function(5)
multiply_by_3 <- make_function(3)

multiply_by_5(4)  # 4 * 5 * 10 = 200
multiply_by_3(4)  # 4 * 3 * 10 = 120
```

## Input Validation and Error Handling

### Basic Input Validation

```{r}
# Simple validation with stopifnot
safe_sqrt <- function(x) {
  stopifnot("Input must be numeric" = is.numeric(x))
  stopifnot("Input must be non-negative" = all(x >= 0, na.rm = TRUE))
  
  return(sqrt(x))
}

safe_sqrt(c(4, 9, 16))
# safe_sqrt(-4)  # Would throw an error

# More detailed validation
validate_and_process <- function(data, column_name) {
  # Check if data is a data frame
  if (!is.data.frame(data)) {
    stop("Input 'data' must be a data frame")
  }
  
  # Check if column exists
  if (! column_name %in% names(data)) {
    stop(paste("Column", column_name, "not found in data"))
  }
  
  # Check if column is numeric
  if (!is.numeric(data[[column_name]])) {
    stop(paste("Column", column_name, "must be numeric"))
  }
  
  # Process the data
  result <- mean(data[[column_name]], na.rm = TRUE)
  return(result)
}

# Test data
test_data <- data.frame(
  age = c(25, 30, 35, NA, 45),
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve")
)

validate_and_process(test_data, "age")
# validate_and_process(test_data, "name")     # Would throw an error
# validate_and_process(test_data, "height")  # Would throw an error
```

### Advanced Error Handling with tryCatch

```{r}
# Robust function with error handling
robust_operation <- function(x, operation = "mean") {
  result <- tryCatch({
    switch(operation,
           "mean" = mean(x, na.rm = TRUE),
           "median" = median(x, na.rm = TRUE),
           "sd" = sd(x, na.rm = TRUE),
           stop("Unsupported operation")
    )
  },
  warning = function(w) {
    cat("Warning occurred:", conditionMessage(w), "\n")
    return(NA)
  },
  error = function(e) {
    cat("Error occurred:", conditionMessage(e), "\n")
    return(NA)
  },
  finally = {
    cat("Operation attempted for", operation, "\n")
  })
  
  return(result)
}

# Test the function
test_data <- c(1, 2, 3, NA, 5)
robust_operation(test_data, "mean")
robust_operation(test_data, "median")
robust_operation(test_data, "invalid_op")  # Handles error gracefully
```

# Advanced Function Concepts

## Higher-Order Functions

Functions that take other functions as arguments or return functions:

```{r}
# Function that takes another function as argument
apply_operation <- function(x, operation_func) {
  return(operation_func(x))
}

# Test with different functions
numbers <- c(1, 2, 3, 4, 5)
apply_operation(numbers, sum)
apply_operation(numbers, mean)
apply_operation(numbers, function(x) max(x) - min(x))

# Function that returns a function
make_multiplier <- function(n) {
  function(x) {
    return(x * n)
  }
}

double <- make_multiplier(2)
triple <- make_multiplier(3)

double(10)  # 20
triple(10)  # 30
```

## Anonymous Functions and Lambda Functions

```{r}
# Traditional function definition
square_traditional <- function(x) x^2

# Anonymous function (lambda)
numbers <- 1:5

# Using anonymous function with lapply
squared_numbers <- lapply(numbers, function(x) x^2)
print(unlist(squared_numbers))

# Using anonymous function with sapply
cubed_numbers <- sapply(numbers, function(x) x^3)
print(cubed_numbers)

# More complex anonymous function
complex_calc <- function(data, func) {
  sapply(data, func)
}

complex_calc(1:5, function(x) (x^2 + 2*x + 1))
```

## Working with Data Frames

### Single Input, Single Output Functions

```{r}
# Function to calculate BMI
calculate_bmi <- function(weight_kg, height_m) {
  # Input validation
  stopifnot(is.numeric(weight_kg), is.numeric(height_m))
  stopifnot(all(weight_kg > 0, na.rm = TRUE))
  stopifnot(all(height_m > 0, na.rm = TRUE))
  
  bmi <- weight_kg / (height_m^2)
  return(round(bmi, 2))
}

# Test the function
weights <- c(70, 80, 65, 75)
heights <- c(1.75, 1.80, 1.65, 1.70)
bmis <- calculate_bmi(weights, heights)
print(bmis)

# Create a data frame function
process_patient_data <- function(patient_df) {
  # Validate input
  required_cols <- c("weight", "height")
  if (!all(required_cols %in% names(patient_df))) {
    stop("Data frame must contain 'weight' and 'height' columns")
  }
  
  # Calculate BMI and add to data frame
  patient_df$bmi <- calculate_bmi(patient_df$weight, patient_df$height)
  
  # Add BMI category
  patient_df$bmi_category <- cut(patient_df$bmi,
                                breaks = c(0, 18.5, 25, 30, Inf),
                                labels = c("Underweight", "Normal", "Overweight", "Obese"),
                                include.lowest = TRUE)
  
  return(patient_df)
}

# Test data
patient_data <- data.frame(
  id = 1:4,
  weight = c(70, 80, 65, 75),
  height = c(1.75, 1.80, 1.65, 1.70),
  age = c(25, 30, 35, 40)
)

processed_data <- process_patient_data(patient_data)
print(processed_data)
```

### Multiple Input, Multiple Output Functions

```{r}
# Comprehensive statistical analysis function
comprehensive_analysis <- function(data, numeric_vars, group_var = NULL) {
  # Input validation
  if (!is.data.frame(data)) stop("Data must be a data frame")
  if (!all(numeric_vars %in% names(data))) stop("Some numeric variables not found")
  
  results <- list()
  
  # Basic descriptive statistics
  descriptive_stats <- data.frame(
    variable = numeric_vars,
    n = sapply(numeric_vars, function(var) sum(!is.na(data[[var]]))),
    mean = sapply(numeric_vars, function(var) round(mean(data[[var]], na.rm = TRUE), 3)),
    median = sapply(numeric_vars, function(var) round(median(data[[var]], na.rm = TRUE), 3)),
    sd = sapply(numeric_vars, function(var) round(sd(data[[var]], na.rm = TRUE), 3)),
    min = sapply(numeric_vars, function(var) round(min(data[[var]], na.rm = TRUE), 3)),
    max = sapply(numeric_vars, function(var) round(max(data[[var]], na.rm = TRUE), 3)),
    stringsAsFactors = FALSE
  )
  
  results$descriptive <- descriptive_stats
  
  # Correlation matrix
  if (length(numeric_vars) > 1) {
    cor_data <- data[, numeric_vars, drop = FALSE]
    cor_data <- cor_data[complete.cases(cor_data), ]
    results$correlation <- round(cor(cor_data), 3)
  }
  
  # Group analysis if group variable provided
  if (!is.null(group_var) && group_var %in% names(data)) {
    group_stats <- list()
    groups <- unique(data[[group_var]])
    groups <- groups[!is.na(groups)]
    
    for (group in groups) {
      group_data <- data[data[[group_var]] == group & !is.na(data[[group_var]]), ]
      
      group_descriptive <- data.frame(
        variable = numeric_vars,
        group = rep(group, length(numeric_vars)),
        n = sapply(numeric_vars, function(var) sum(!is.na(group_data[[var]]))),
        mean = sapply(numeric_vars, function(var) round(mean(group_data[[var]], na.rm = TRUE), 3)),
        median = sapply(numeric_vars, function(var) round(median(group_data[[var]], na.rm = TRUE), 3)),
        sd = sapply(numeric_vars, function(var) round(sd(group_data[[var]], na.rm = TRUE), 3)),
        stringsAsFactors = FALSE
      )
      
      group_stats[[as.character(group)]] <- group_descriptive
    }
    
    results$group_analysis <- do.call(rbind, group_stats)
  }
  
  # Data quality report
  quality_report <- data.frame(
    variable = names(data),
    type = sapply(data, class),
    missing_count = sapply(data, function(x) sum(is.na(x))),
    missing_percent = sapply(data, function(x) round(100 * sum(is.na(x)) / length(x), 2)),
    stringsAsFactors = FALSE
  )
  
  results$data_quality <- quality_report
  
  return(results)
}

# Test with sample data
sample_data <- data.frame(
  id = 1:100,
  age = rnorm(100, 40, 10),
  weight = rnorm(100, 70, 15),
  height = rnorm(100, 1.7, 0.1),
  group = sample(c("A", "B", "C"), 100, replace = TRUE),
  stringsAsFactors = FALSE
)

# Add some missing values
sample_data$age[c(5, 15, 25)] <- NA
sample_data$weight[c(10, 20)] <- NA

analysis_results <- comprehensive_analysis(
  data = sample_data,
  numeric_vars = c("age", "weight", "height"),
  group_var = "group"
)

# View results
print("Descriptive Statistics:")
print(analysis_results$descriptive)

print("\nCorrelation Matrix:")
print(analysis_results$correlation)

print("\nGroup Analysis (first few rows):")
print(head(analysis_results$group_analysis))

print("\nData Quality Report:")
print(analysis_results$data_quality)
```

# Tidy Evaluation and Advanced dplyr Usage

## Understanding `{{}}`, `!!`, and `!!!`

These operators are part of tidy evaluation, used extensively in the tidyverse:

```{r}
library(dplyr)
library(rlang)

# Using {{}} for direct variable capture
summarize_variable <- function(data, group_var, summary_var) {
  data %>%
    group_by({{ group_var }}) %>%
    summarise(
      n = n(),
      mean = mean({{ summary_var }}, na.rm = TRUE),
      median = median({{ summary_var }}, na.rm = TRUE),
      sd = sd({{ summary_var }}, na.rm = TRUE),
      .groups = 'drop'
    )
}

# Using !! for unquoting single variables
summarize_variable_unquote <- function(data, group_var, summary_var) {
  group_sym <- sym(group_var)
  summary_sym <- sym(summary_var)
  
  data %>%
    group_by(!! group_sym) %>%
    summarise(
      n = n(),
      mean = mean(!! summary_sym, na.rm = TRUE),
      median = median(!!summary_sym, na.rm = TRUE),
      sd = sd(!!summary_sym, na.rm = TRUE),
      .groups = 'drop'
    )
}

# Using ! !! for splicing multiple variables
select_multiple_vars <- function(data, ...) {
  vars <- list(...)
  data %>% select(!!! vars)
}

# Test data
test_df <- data.frame(
  group = rep(c("A", "B"), each = 50),
  value1 = rnorm(100, 10, 2),
  value2 = rnorm(100, 20, 3),
  id = 1:100
)

# Test the functions
result1 <- summarize_variable(test_df, group, value1)
print(result1)

result2 <- summarize_variable_unquote(test_df, "group", "value1")
print(result2)

selected_data <- select_multiple_vars(test_df, "id", "group", "value1")
print(head(selected_data))
```

## Advanced Data Processing Function

```{r}
library(tidyr)
library(rlang)
library(dplyr)
# Comprehensive data processing function using tidy evaluation
process_clinical_data <- function(data, group_vars, analysis_vars, 
                                 id_var = "USUBJID", stats = c("n", "mean", "median", "sd")) {
  
  # Input validation
  stopifnot(is.data.frame(data))
  stopifnot(all(c(group_vars, analysis_vars, id_var) %in% names(data)))
  
  # Convert character vectors to symbols
  group_syms <- syms(group_vars)
  analysis_syms <- syms(analysis_vars)
  id_sym <- sym(id_var)
  
  results <- list()
  
  # Process each analysis variable
  for (var in analysis_vars) {
    var_sym <- sym(var)
    
    # Create summary statistics
    summary_stats <- data %>%
      group_by(!!! group_syms) %>%
      summarise(
        n = n_distinct(!!id_sym),
        mean = round(mean(!!var_sym, na.rm = TRUE), 2),
        median = round(median(!!var_sym, na.rm = TRUE), 2),
        sd = round(sd(!! var_sym, na.rm = TRUE), 3),
        min = round(min(!!var_sym, na.rm = TRUE), 2),
        max = round(max(!!var_sym, na.rm = TRUE), 2),
        q25 = round(quantile(!! var_sym, 0.25, na.rm = TRUE), 2),
        q75 = round(quantile(!!var_sym, 0.75, na.rm = TRUE), 2),
        .groups = 'drop'
      )
    
    # Select only requested statistics
    summary_stats <- summary_stats %>%
      select(all_of(c(group_vars, stats)))
    
    # Reshape to long format then wide format for presentation
    if (length(group_vars) == 1) {
      transposed_data <- summary_stats %>%
        pivot_longer(
          cols = all_of(stats),
          names_to = "statistic",
          values_to = "value"
        ) %>%
        mutate(value = as.character(value)) %>%
        pivot_wider(
          names_from = all_of(group_vars[1]),
          values_from = value
        ) %>%
        mutate(variable = var) %>%
        select(variable, statistic, everything())
      
      results[[var]] <- transposed_data
    } else {
      results[[var]] <- summary_stats
    }
  }
  
  return(results)
}

# Create sample clinical data
set.seed(123)
clinical_data <- data.frame(
  USUBJID = paste0("SUBJ-", sprintf("%03d", 1:150)),
  TRT01A = sample(c("Placebo", "Treatment A", "Treatment B"), 150, replace = TRUE),
  AGE = round(rnorm(150, 65, 12)),
  WEIGHTBL = round(rnorm(150, 75, 15), 1),
  HEIGHTBL = round(rnorm(150, 170, 10), 1),
  SEX = sample(c("M", "F"), 150, replace = TRUE),
  stringsAsFactors = FALSE
)

# Process the data
clinical_results <- process_clinical_data(
  data = clinical_data,
  group_vars = "TRT01A",
  analysis_vars = c("AGE", "WEIGHTBL"),
  id_var = "USUBJID",
  stats = c("n", "mean", "median", "sd")
)

print("Age Analysis:")
print(clinical_results$AGE)

print("\nWeight Analysis:")
print(clinical_results$WEIGHTBL)
```

# Advanced Topics

## Function Factories and Closures

```{r}
# Function factory that creates specialized functions
create_validator <- function(min_val = -Inf, max_val = Inf, allow_na = TRUE) {
  function(x, var_name = "variable") {
    # Check if NA values are allowed
    if (! allow_na && any(is.na(x))) {
      stop(paste(var_name, "contains NA values, which are not allowed"))
    }
    
    # Check range
    x_clean <- x[!is.na(x)]
    if (any(x_clean < min_val | x_clean > max_val)) {
      stop(paste(var_name, "contains values outside the allowed range [", 
                 min_val, ",", max_val, "]"))
    }
    
    return(TRUE)
  }
}

# Create specialized validators
age_validator <- create_validator(min_val = 0, max_val = 120, allow_na = FALSE)
weight_validator <- create_validator(min_val = 0, max_val = 500, allow_na = TRUE)
percentage_validator <- create_validator(min_val = 0, max_val = 100, allow_na = FALSE)

# Test the validators
test_ages <- c(25, 35, 45, 55, 65)
age_validator(test_ages, "Age")

test_weights <- c(50, 75, 90, NA, 120)
weight_validator(test_weights, "Weight")

# This would throw an error: 
# bad_ages <- c(25, 35, -5, 200)
# age_validator(bad_ages, "Age")
```

## Memoization for Performance

```{r}
# Function with memoization for expensive calculations
create_memoized_function <- function(func) {
  cache <- new.env(parent = emptyenv())
  
  function(... ) {
    # Create a key from the arguments
    key <- paste(list(...), collapse = "_")
    
    # Check if result is in cache
    if (exists(key, cache)) {
      cat("Cache hit!\n")
      return(get(key, cache))
    }
    
    # Calculate result and store in cache
    cat("Computing result...\n")
    result <- func(...)
    assign(key, result, cache)
    return(result)
  }
}

# Expensive calculation function
expensive_calculation <- function(n) {
  Sys.sleep(1)  # Simulate expensive operation
  return(sum(1:n))
}

# Create memoized version
fast_calculation <- create_memoized_function(expensive_calculation)

# Test memoization
system.time(result1 <- fast_calculation(1000))  # First call - slow
system.time(result2 <- fast_calculation(1000))  # Second call - fast (cached)
system.time(result3 <- fast_calculation(2000))  # New argument - slow again
```

## Method Dispatch and S3 Classes

```{r}
# Create a custom class and methods
create_summary_stats <- function(x, ...) {
  UseMethod("create_summary_stats")
}

# Method for numeric vectors
create_summary_stats.numeric <- function(x, ...) {
  result <- list(
    data = x,
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    n = length(x),
    missing = sum(is.na(x))
  )
  class(result) <- "numeric_summary"
  return(result)
}

# Method for data frames
create_summary_stats.data.frame <- function(x, ...) {
  numeric_vars <- sapply(x, is.numeric)
  result <- list(
    data = x,
    numeric_summaries = lapply(x[numeric_vars], create_summary_stats.numeric),
    dimensions = dim(x),
    variables = names(x)
  )
  class(result) <- "dataframe_summary"
  return(result)
}

# Print methods
print.numeric_summary <- function(x, ...) {
  cat("Numeric Summary\n")
  cat("===============\n")
  cat("n:", x$n, "\n")
  cat("Missing:", x$missing, "\n")
  cat("Mean:", round(x$mean, 3), "\n")
  cat("Median:", round(x$median, 3), "\n")
  cat("SD:", round(x$sd, 3), "\n")
}

print.dataframe_summary <- function(x, ...) {
  cat("Data Frame Summary\n")
  cat("==================\n")
  cat("Dimensions:", x$dimensions[1], "rows x", x$dimensions[2], "columns\n")
  cat("Variables:", paste(x$variables, collapse = ", "), "\n")
  cat("Numeric variables:", length(x$numeric_summaries), "\n")
}

# Test the methods
numeric_data <- rnorm(100, 50, 10)
df_data <- data.frame(
  x = rnorm(50),
  y = rnorm(50, 10, 2),
  group = sample(letters[1:3], 50, replace = TRUE)
)

summary1 <- create_summary_stats(numeric_data)
summary2 <- create_summary_stats(df_data)

print(summary1)
print(summary2)
```

# Testing and Documentation

## Unit Testing with testthat

```{r, eval=FALSE}
# Install required packages (run once)
if (!require(testthat)) install.packages("testthat")
if (!require(devtools)) install.packages("devtools")

# Set up testing structure
usethis::use_testthat()
```

Example test file (`tests/testthat/test-statistical-functions.R`):

```{r, eval=FALSE}
# Test file content
library(testthat)

test_that("calculate_bmi works correctly", {
  # Test normal case
  expect_equal(calculate_bmi(70, 1.75), 22.86)
  
  # Test vector input
  weights <- c(70, 80)
  heights <- c(1.75, 1.8)
  expected <- c(22.86, 24.69)
  expect_equal(calculate_bmi(weights, heights), expected, tolerance = 0.01)
  
  # Test error conditions
  expect_error(calculate_bmi(-70, 1.75))  # Negative weight
  expect_error(calculate_bmi(70, -1.75))  # Negative height
  expect_error(calculate_bmi("70", 1.75)) # Non-numeric input
})

test_that("comprehensive_analysis handles edge cases", {
  # Test with minimal data
  minimal_data <- data.frame(x = c(1, 2, 3), y = c(4, 5, 6))
  result <- comprehensive_analysis(minimal_data, c("x", "y"))
  
  expect_true(is.list(result))
  expect_true("descriptive" %in% names(result))
  expect_true("correlation" %in% names(result))
  expect_equal(nrow(result$descriptive), 2)
  
  # Test with missing data
  na_data <- data.frame(x = c(1, NA, 3), y = c(4, 5, NA))
  result_na <- comprehensive_analysis(na_data, c("x", "y"))
  
  expect_true(any(result_na$descriptive$n < 3))  # Should handle missing values
})
```

## Documentation with roxygen2

```{r}
#' Calculate Body Mass Index (BMI)
#'
#' This function calculates BMI using the standard formula:  weight (kg) / height (m)^2
#'
#' @param weight_kg Numeric vector of weights in kilograms.  Must be positive.
#' @param height_m Numeric vector of heights in meters. Must be positive.
#'
#' @return Numeric vector of BMI values, rounded to 2 decimal places
#'
#' @details
#' BMI categories:
#' - Underweight: < 18.5
#' - Normal weight: 18.5-24.9
#' - Overweight: 25-29.9
#' - Obese: ≥ 30
#'
#' @examples
#' # Single values
#' calculate_bmi(70, 1.75)
#'
#' # Multiple values
#' weights <- c(60, 70, 80, 90)
#' heights <- c(1.6, 1.7, 1.8, 1.9)
#' calculate_bmi(weights, heights)
#'
#' @seealso
#' \url{https://www.cdc.gov/healthyweight/assessing/bmi/} for BMI information
#'
#' @export
calculate_bmi <- function(weight_kg, height_m) {
  # Input validation
  stopifnot("Weight must be numeric" = is.numeric(weight_kg))
  stopifnot("Height must be numeric" = is.numeric(height_m))
  stopifnot("Weight must be positive" = all(weight_kg > 0, na.rm = TRUE))
  stopifnot("Height must be positive" = all(height_m > 0, na.rm = TRUE))
  
  # Calculate BMI
  bmi <- weight_kg / (height_m^2)
  return(round(bmi, 2))
}

#' Comprehensive Statistical Analysis
#'
#' Performs comprehensive statistical analysis on numeric variables in a dataset,
#' including descriptive statistics, correlations, and optional group comparisons.
#'
#' @param data A data frame containing the variables to analyze
#' @param numeric_vars Character vector of numeric variable names to analyze
#' @param group_var Optional character string specifying a grouping variable for
#'   stratified analysis
#'
#' @return A list containing:
#'   \item{descriptive}{Data frame with descriptive statistics for each variable}
#'   \item{correlation}{Correlation matrix (if multiple numeric variables)}
#'   \item{group_analysis}{Stratified analysis by group (if group_var specified)}
#'   \item{data_quality}{Data quality report with missing value information}
#'
#' @examples
#' # Basic analysis
#' data(mtcars)
#' results <- comprehensive_analysis(mtcars, c("mpg", "hp", "wt"))
#'
#' # Analysis with grouping
#' results_grouped <- comprehensive_analysis(mtcars, c("mpg", "hp"), "cyl")
#'
#' @export
comprehensive_analysis <- function(data, numeric_vars, group_var = NULL) {
  # [Function body as defined earlier]
  # Input validation
  if (!is.data.frame(data)) stop("Data must be a data frame")
  if (!all(numeric_vars %in% names(data))) stop("Some numeric variables not found")
  
  results <- list()
  
  # Basic descriptive statistics
  descriptive_stats <- data.frame(
    variable = numeric_vars,
    n = sapply(numeric_vars, function(var) sum(!is.na(data[[var]]))),
    mean = sapply(numeric_vars, function(var) round(mean(data[[var]], na.rm = TRUE), 3)),
    median = sapply(numeric_vars, function(var) round(median(data[[var]], na.rm = TRUE), 3)),
    sd = sapply(numeric_vars, function(var) round(sd(data[[var]], na.rm = TRUE), 3)),
    min = sapply(numeric_vars, function(var) round(min(data[[var]], na.rm = TRUE), 3)),
    max = sapply(numeric_vars, function(var) round(max(data[[var]], na.rm = TRUE), 3)),
    stringsAsFactors = FALSE
  )
  
  results$descriptive <- descriptive_stats
  
  # Correlation matrix
  if (length(numeric_vars) > 1) {
    cor_data <- data[, numeric_vars, drop = FALSE]
    cor_data <- cor_data[complete.cases(cor_data), ]
    results$correlation <- round(cor(cor_data), 3)
  }
  
  # Group analysis if group variable provided
  if (!is.null(group_var) && group_var %in% names(data)) {
    group_stats <- list()
    groups <- unique(data[[group_var]])
    groups <- groups[!is.na(groups)]
    
    for (group in groups) {
      group_data <- data[data[[group_var]] == group & !is.na(data[[group_var]]), ]
      
      group_descriptive <- data.frame(
        variable = numeric_vars,
        group = rep(group, length(numeric_vars)),
        n = sapply(numeric_vars, function(var) sum(!is.na(group_data[[var]]))),
        mean = sapply(numeric_vars, function(var) round(mean(group_data[[var]], na.rm = TRUE), 3)),
        median = sapply(numeric_vars, function(var) round(median(group_data[[var]], na.rm = TRUE), 3)),
        sd = sapply(numeric_vars, function(var) round(sd(group_data[[var]], na.rm = TRUE), 3)),
        stringsAsFactors = FALSE
      )
      
      group_stats[[as.character(group)]] <- group_descriptive
    }
    
    results$group_analysis <- do.call(rbind, group_stats)
  }
  
  # Data quality report
  quality_report <- data.frame(
    variable = names(data),
    type = sapply(data, class),
    missing_count = sapply(data, function(x) sum(is.na(x))),
    missing_percent = sapply(data, function(x) round(100 * sum(is.na(x)) / length(x), 2)),
    stringsAsFactors = FALSE
  )
  
  results$data_quality <- quality_report
  
  return(results)
}
```

# Performance and Optimization

## Vectorization vs Loops

```{r}
# Demonstrate vectorization benefits
set.seed(123)
large_vector <- rnorm(1000000)

# Slow approach - using a loop
loop_square <- function(x) {
  result <- numeric(length(x))
  for (i in seq_along(x)) {
    result[i] <- x[i]^2
  }
  return(result)
}

# Fast approach - vectorized
vectorized_square <- function(x) {
  return(x^2)
}

# Benchmark the functions
library(microbenchmark)

benchmark_results <- microbenchmark(
  loop = loop_square(large_vector[1:1000]),
  vectorized = vectorized_square(large_vector[1:1000]),
  times = 100
)

print(benchmark_results)
```

## Memory Management

```{r}
# Efficient data processing function
efficient_data_processor <- function(data, chunk_size = 1000) {
  n_rows <- nrow(data)
  n_chunks <- ceiling(n_rows / chunk_size)
  
  results <- vector("list", n_chunks)
  
  for (i in seq_len(n_chunks)) {
    start_row <- (i - 1) * chunk_size + 1
    end_row <- min(i * chunk_size, n_rows)
    
    chunk <- data[start_row:end_row, , drop = FALSE]
    
    # Process chunk (example: calculate row means for numeric columns)
    numeric_cols <- sapply(chunk, is.numeric)
    if (any(numeric_cols)) {
      chunk_means <- rowMeans(chunk[, numeric_cols, drop = FALSE], na.rm = TRUE)
      results[[i]] <- data.frame(
        chunk = i,
        row_start = start_row,
        row_end = end_row,
        mean_value = mean(chunk_means, na.rm = TRUE)
      )
    }
  }
  
  return(do.call(rbind, results))
}

# Test with large dataset
large_data <- data.frame(
  id = 1:5000,
  x1 = rnorm(5000),
  x2 = rnorm(5000, 10, 2),
  x3 = rnorm(5000, -5, 3),
  group = sample(letters[1:5], 5000, replace = TRUE)
)

chunk_results <- efficient_data_processor(large_data, chunk_size = 1000)
print(chunk_results)
```

# Best Practices and Common Patterns

## Function Design Principles

```{r}
# Good function design example
#' Process Survey Responses
#'
#' A well-designed function that follows best practices:
#' - Single responsibility
#' - Clear naming
#' - Input validation
#' - Meaningful return values
#' - Error handling
process_survey_responses <- function(responses, 
                                   scale_min = 1, 
                                   scale_max = 5,
                                   remove_incomplete = TRUE,
                                   return_summary = FALSE) {
  
  # Input validation with clear error messages
  if (!is.data.frame(responses)) {
    stop("Argument 'responses' must be a data. frame")
  }
  
  if (!is.numeric(scale_min) || !is.numeric(scale_max) || scale_min >= scale_max) {
    stop("scale_min and scale_max must be numeric with scale_min < scale_max")
  }
  
  # Required columns check
  required_cols <- c("participant_id", "response_value")
  missing_cols <- setdiff(required_cols, names(responses))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }
  
  # Data processing
  processed_data <- responses
  
  # Remove out-of-range responses
  valid_range <- processed_data$response_value >= scale_min & 
                 processed_data$response_value <= scale_max
  processed_data <- processed_data[valid_range & !is.na(valid_range), ]
  
  # Remove incomplete responses if requested
  if (remove_incomplete) {
    complete_cases <- complete.cases(processed_data)
    processed_data <- processed_data[complete_cases, ]
  }
  
  # Add standardized score (0-1 scale)
  processed_data$standardized_score <- 
    (processed_data$response_value - scale_min) / (scale_max - scale_min)
  
  # Return summary or full data based on parameter
  if (return_summary) {
    summary_stats <- list(
      n_responses = nrow(processed_data),
      mean_response = mean(processed_data$response_value),
      mean_standardized = mean(processed_data$standardized_score),
      response_distribution = table(processed_data$response_value)
    )
    return(summary_stats)
  } else {
    return(processed_data)
  }
}

# Example usage
sample_responses <- data.frame(
  participant_id = paste0("P", 1:100),
  response_value = sample(1:5, 100, replace = TRUE, prob = c(0.1, 0.2, 0.4, 0.2, 0.1)),
  age_group = sample(c("18-30", "31-50", "51+"), 100, replace = TRUE)
)

processed_responses <- process_survey_responses(sample_responses)
summary_only <- process_survey_responses(sample_responses, return_summary = TRUE)

print("Processed data (first 10 rows):")
print(head(processed_responses, 10))

print("\nSummary statistics:")
print(summary_only)
```

## Error Handling Patterns

```{r}
# Comprehensive error handling example
robust_file_processor <- function(file_path, expected_columns = NULL) {
  
  # Create a custom error class for better error handling
  create_processing_error <- function(message, type = "general") {
    err <- structure(
      list(message = message, type = type, call = sys.call(-1)),
      class = c("processing_error", "error", "condition")
    )
    return(err)
  }
  
  tryCatch({
    # Check if file exists
    if (!file.exists(file_path)) {
      stop(create_processing_error(
        paste("File does not exist:", file_path), 
        "file_not_found"
      ))
    }
    
    # Try to read the file
    file_ext <- tools::file_ext(file_path)
    data <- switch(tolower(file_ext),
                   "csv" = read.csv(file_path, stringsAsFactors = FALSE),
                   "txt" = read.table(file_path, header = TRUE, stringsAsFactors = FALSE),
                   stop(create_processing_error(
                     paste("Unsupported file extension:", file_ext),
                     "unsupported_format"
                   ))
    )
    
    # Validate columns if specified
    if (!is.null(expected_columns)) {
      missing_cols <- setdiff(expected_columns, names(data))
      if (length(missing_cols) > 0) {
        stop(create_processing_error(
          paste("Missing expected columns:", paste(missing_cols, collapse = ", ")),
          "missing_columns"
        ))
      }
    }
    
    # Return successful result
    return(list(
      success = TRUE,
      data = data,
      message = paste("Successfully processed", nrow(data), "rows from", basename(file_path))
    ))
    
  }, processing_error = function(e) {
    return(list(
      success = FALSE,
      data = NULL,
      error_type = e$type,
      message = e$message
    ))
  }, error = function(e) {
    return(list(
      success = FALSE,
      data = NULL,
      error_type = "unexpected",
      message = paste("Unexpected error:", e$message)
    ))
  })
}

# Test the error handling (create a temporary file for demonstration)
temp_file <- tempfile(fileext = ".csv")
write.csv(data.frame(x = 1:5, y = letters[1:5]), temp_file, row.names = FALSE)

# Successful case
result_success <- robust_file_processor(temp_file, c("x", "y"))
print("Success case:")
print(result_success$message)

# Error case - missing columns
result_error <- robust_file_processor(temp_file, c("x", "y", "z"))
print("\nError case:")
print(paste("Error type:", result_error$error_type))
print(paste("Message:", result_error$message))

# Clean up
unlink(temp_file)
```

# Practical Exercises

## Exercise 1: Data Validation Function

Create a comprehensive data validation function: 

```{r}
#' Comprehensive Data Validation
#'
#' Validates a data frame against specified rules and returns a detailed report
validate_dataset <- function(data, rules = NULL) {
  validation_results <- list(
    is_valid = TRUE,
    errors = character(0),
    warnings = character(0),
    summary = list()
  )
  
  # Basic validation
  if (!is.data.frame(data)) {
    validation_results$is_valid <- FALSE
    validation_results$errors <- c(validation_results$errors, "Input is not a data frame")
    return(validation_results)
  }
  
  validation_results$summary$n_rows <- nrow(data)
  validation_results$summary$n_cols <- ncol(data)
  validation_results$summary$column_names <- names(data)
  
  # Check for empty dataset
  if (nrow(data) == 0) {
    validation_results$warnings <- c(validation_results$warnings, "Dataset is empty")
  }
  
  # Rule-based validation
  if (!is.null(rules)) {
    for (rule_name in names(rules)) {
      rule <- rules[[rule_name]]
      
      if (rule$type == "required_columns") {
        missing_cols <- setdiff(rule$columns, names(data))
        if (length(missing_cols) > 0) {
          validation_results$is_valid <- FALSE
          validation_results$errors <- c(
            validation_results$errors,
            paste("Missing required columns:", paste(missing_cols, collapse = ", "))
          )
        }
      }
      
      if (rule$type == "data_types") {
        for (col in names(rule$expected_types)) {
          if (col %in% names(data)) {
            expected_type <- rule$expected_types[[col]]
            actual_type <- class(data[[col]])[1]
            
            if (actual_type != expected_type) {
              validation_results$warnings <- c(
                validation_results$warnings,
                paste("Column", col, "expected type", expected_type, 
                     "but found", actual_type)
              )
            }
          }
        }
      }
      
      if (rule$type == "value_ranges") {
        for (col in names(rule$ranges)) {
          if (col %in% names(data) && is.numeric(data[[col]])) {
            range_rule <- rule$ranges[[col]]
            out_of_range <- data[[col]] < range_rule$min | data[[col]] > range_rule$max
            out_of_range <- out_of_range[!is.na(out_of_range)]
            
            if (any(out_of_range)) {
              validation_results$warnings <- c(
                validation_results$warnings,
                paste("Column", col, "has", sum(out_of_range), 
                     "values outside range [", range_rule$min, ",", range_rule$max, "]")
              )
            }
          }
        }
      }
    }
  }
  
  # Check for missing values
  missing_summary <- sapply(data, function(x) sum(is.na(x)))
  validation_results$summary$missing_values <- missing_summary[missing_summary > 0]
  
  return(validation_results)
}

# Define validation rules
validation_rules <- list(
  required_cols = list(
    type = "required_columns",
    columns = c("id", "age", "treatment")
  ),
  data_types = list(
    type = "data_types",
    expected_types = list(
      id = "character",
      age = "numeric",
      treatment = "character"
    )
  ),
  ranges = list(
    type = "value_ranges",
    ranges = list(
      age = list(min = 18, max = 100)
    )
  )
)

# Test data
test_data_valid <- data.frame(
  id = paste0("P", 1:10),
  age = sample(25:75, 10),
  treatment = sample(c("A", "B"), 10, replace = TRUE),
  stringsAsFactors = FALSE
)

test_data_invalid <- data.frame(
  id = paste0("P", 1:10),
  age = c(sample(25:75, 8), 15, 105),  # Some ages out of range
  weight = rnorm(10, 70, 10),          # Missing treatment column
  stringsAsFactors = FALSE
)

# Run validations
validation_valid <- validate_dataset(test_data_valid, validation_rules)
validation_invalid <- validate_dataset(test_data_invalid, validation_rules)

print("Valid dataset validation:")
print(paste("Is valid:", validation_valid$is_valid))
print("Warnings:")
print(validation_valid$warnings)

print("\nInvalid dataset validation:")
print(paste("Is valid:", validation_invalid$is_valid))
print("Errors:")
print(validation_invalid$errors)
print("Warnings:")
print(validation_invalid$warnings)
```

## Exercise 2: Advanced Statistical Function

```{r}
#' Advanced Statistical Analysis with Bootstrap Confidence Intervals
advanced_statistics <- function(data, variable, group = NULL, conf_level = 0.95, n_bootstrap = 1000) {
  
  # Input validation
  stopifnot(is.data.frame(data))
  stopifnot(variable %in% names(data))
  stopifnot(is.numeric(data[[variable]]))
  stopifnot(conf_level > 0 & conf_level < 1)
  
  # Bootstrap function for confidence intervals
  bootstrap_mean <- function(x, n_boot = n_bootstrap) {
    boot_means <- replicate(n_boot, {
      sample_data <- sample(x, length(x), replace = TRUE)
      mean(sample_data, na.rm = TRUE)
    })
    return(boot_means)
  }
  
  # Calculate confidence interval
  calc_ci <- function(x, conf_level) {
    alpha <- 1 - conf_level
    boot_samples <- bootstrap_mean(x[!is.na(x)])
    ci_lower <- quantile(boot_samples, alpha/2)
    ci_upper <- quantile(boot_samples, 1 - alpha/2)
    return(c(lower = ci_lower, upper = ci_upper))
  }
  
  # Analysis function
  analyze_group <- function(data_subset, group_name = "Overall") {
    values <- data_subset[[variable]]
    clean_values <- values[!is.na(values)]
    
    if (length(clean_values) < 2) {
      return(data.frame(
        group = group_name,
        n = length(clean_values),
        mean = NA,
        median = NA,
        sd = NA,
        se = NA,
        ci_lower = NA,
        ci_upper = NA,
        skewness = NA,
        kurtosis = NA
      ))
    }
    
    # Basic statistics
    mean_val <- mean(clean_values)
    median_val <- median(clean_values)
    sd_val <- sd(clean_values)
    se_val <- sd_val / sqrt(length(clean_values))
    
    # Confidence interval
    ci <- calc_ci(clean_values, conf_level)
    
    # Higher moments
    skewness <- function(x) {
      x_centered <- x - mean(x)
      n <- length(x)
      skew <- (sum(x_centered^3) / n) / (sum(x_centered^2) / n)^(3/2)
      return(skew)
    }
    
    kurtosis <- function(x) {
      x_centered <- x - mean(x)
      n <- length(x)
      kurt <- (sum(x_centered^4) / n) / (sum(x_centered^2) / n)^2 - 3
      return(kurt)
    }
    
    return(data.frame(
      group = group_name,
      n = length(clean_values),
      mean = round(mean_val, 3),
      median = round(median_val, 3),
      sd = round(sd_val, 3),
      se = round(se_val, 3),
      ci_lower = round(ci[1], 3),
      ci_upper = round(ci[2], 3),
      skewness = round(skewness(clean_values), 3),
      kurtosis = round(kurtosis(clean_values), 3)
    ))
  }
  
  # Perform analysis
  if (is.null(group)) {
    results <- analyze_group(data, "Overall")
  } else {
    if (! group %in% names(data)) {
      stop(paste("Group variable", group, "not found in data"))
    }
    
    groups <- unique(data[[group]])
    groups <- groups[!is.na(groups)]
    
    group_results <- lapply(groups, function(g) {
      subset_data <- data[data[[group]] == g & !is.na(data[[group]]), ]
      analyze_group(subset_data, as.character(g))
    })
    
    results <- do.call(rbind, group_results)
  }
  
  return(results)
}

# Test the advanced statistics function
set.seed(123)
advanced_test_data <- data.frame(
  id = 1:200,
  value = c(rnorm(100, 50, 10), rnorm(100, 55, 12)),
  group = rep(c("Control", "Treatment"), each = 100),
  stringsAsFactors = FALSE
)

# Add some missing values
advanced_test_data$value[sample(1:200, 10)] <- NA

# Run analysis
overall_analysis <- advanced_statistics(advanced_test_data, "value")
grouped_analysis <- advanced_statistics(advanced_test_data, "value", "group")

print("Overall Analysis:")
print(overall_analysis)

print("\nGrouped Analysis:")
print(grouped_analysis)
```

# Summary and Best Practices

## Key Takeaways

1. **Start Simple**: Begin with basic functions and gradually add complexity
2. **Validate Inputs**: Always check that inputs are what you expect
3. **Handle Errors Gracefully**: Use tryCatch and informative error messages
4. **Document Your Functions**: Use roxygen2 comments for clear documentation
5. **Test Your Functions**: Write unit tests to ensure reliability
6. **Follow Naming Conventions**: Use clear, descriptive function and variable names
7. **Keep Functions Focused**: Each function should do one thing well
8. **Consider Performance**:  Use vectorization and efficient algorithms
9. **Plan for Reusability**: Write functions that can be easily used in different contexts
10. **Version Control**: Track changes to your functions over time

## Function Writing Checklist

Before finalizing a function, ask yourself: 

- [ ] Does the function have a clear, single purpose?
- [ ] Are the inputs validated appropriately?
- [ ] Are error messages helpful and informative?
- [ ] Is the function documented with examples?
- [ ] Have edge cases been considered and tested?
- [ ] Is the function efficient for its intended use?
- [ ] Can the function be easily tested? 
- [ ] Is the code readable and well-commented? 
- [ ] Does the function follow naming conventions? 
- [ ] Are there any dependencies that should be explicit?

By following these principles and practices, you'll be able to create robust, reusable, and maintainable functions that will serve you well in your R programming journey. 