---
title: "R Package Development"
---

> A practical, end-to-end cookbook for building robust **R packages** with **devtools**, **roxygen2**, and **testthat**. Written for day‑to‑day use, from scaffolding to checks and tests.

## 1. Prerequisites

-   R (≥ 4.1 recommended so you have the base pipe `|>`)
-   RTools (Windows) / Xcode command line tools (macOS) for compilation
-   Recommended packages:
    -   `install.packages(c("devtools", "roxygen2", "usethis", "testthat", "pkgdown"))`

```{r, warning=FALSE,error=FALSE}
# Load helpers
library(devtools)
library(usethis)
```

## 2. Create the Package Skeleton

### 2.1 Create a new package directory

```{r, eval=FALSE warning=FALSE,error=FALSE}
# Creates simutils with DESCRIPTION, NAMESPACE placeholder, R/ etc.
#usethis::create_package("simutils")
# Or from an existing project folder
# usethis::create_package(getwd())
```

### 2.2 Initialize Git and a README

```{r, eval=FALSE}
usethis::use_git()
usethis::use_readme_rmd()  # README.Rmd + badge placeholders
```

### 2.3 The minimal structure (what you get)

```{r, eval=FALSE,warning=FALSE,error=FALSE}        
simutils/
  DESCRIPTION
  NAMESPACE
  R/
  man/
```

> **Tip:** Keep all your R code inside `R/`. Avoid subfolders there. Group related functions in sensible files (e.g., `na.R`, `sample.R`).

## 3. DESCRIPTION: Package Metadata

Run helpers and then edit as needed.

```{r, eval=FALSE}
usethis::use_description(fields = list(
  Title = "Simulation Utilities",
  Description = "Convenience functions for sampling and NA counting.",
  `Authors@R` = 'person(given = "Your", family = "Name", role = c("aut","cre"), email = "you@example.com")',
  License = "MIT + file LICENSE"
))

usethis::use_mit_license("Your Name")
```

Key fields you usually set:

-   `Package`, `Title`, `Version`, `Description`
-   `Authors@R` (with `aut`, `cre` roles)
-   `License` (e.g., MIT)
-   `Depends`, `Imports`, `Suggests` (use helpers below)

## 4. NAMESPACE: Exports and Imports (auto‑generated)

> You generally **do not edit NAMESPACE by hand**. It’s generated from your roxygen tags in code.

-   `@export` marks a function for users.
-   `@import`, `@importFrom` declare your dependencies at function level.

You’ll generate it with `devtools::document()`.

## 5. Write Functions + roxygen2 Headers

Create your first R file:

```{r, eval=FALSE}
# File: R/na.R
#' Count NAs in a vector
#'
#' Count number of missing values (NA) in a numeric/logical/character vector.
#'
#' @param x A vector.
#' @return Integer count of NAs in `x`.
#' @examples
#' sum_na(c(1, NA, 3))
#' sum_na(airquality$Ozone)
#' @export
sum_na <- function(x) {
  sum(is.na(x))
}
```

Add a second exported function that works on data frames/matrices:

```{r, eval=FALSE}
# File: R/na_counter.R
#' Column-wise NA counts
#'
#' Returns the number of NAs per column for a data frame or matrix.
#'
#' @param x A data frame or matrix.
#' @return A named integer vector of NA counts per column.
#' @examples
#' na_counter(airquality)
#' @export
na_counter <- function(x) {
  if (!is.data.frame(x) && !is.matrix(x)) stop("x must be a data.frame or matrix")
  colSums(is.na(x))
}
```

A simple sampler utility with a documented argument:

```{r, eval=FALSE}
# File: R/sample_from_data.R
#' Sample rows from a data frame
#'
#' Draw a row-wise sample with optional replacement.
#'
#' @param data A data frame.
#' @param size Number of rows to sample.
#' @param replace Logical; sample with replacement?
#' @return A data frame containing the sampled rows.
#' @examples
#' sample_from_data(airquality, 10)
#' sample_from_data(airquality, 10, replace = TRUE)
#' @export
sample_from_data <- function(data, size, replace = FALSE) {
  stopifnot(is.data.frame(data))
  idx <- sample.int(nrow(data), size, replace = replace)
  data[idx, , drop = FALSE]
}
```

## 6. Generate Documentation

```{r, eval=FALSE}
# Parses roxygen headers -> creates/updates NAMESPACE and man/*.Rd
devtools::document()
```

This will populate `man/` with help files and refresh `NAMESPACE`.

## 7. Add Data, Vignettes, Tests (Optional but Recommended)

### 7.1 Package data

```{r, eval=FALSE}
# Create example data
sim_dat <- data.frame(
  ID = 1:10,
  Value = sample(1:11, 10),
  Apples = sample(c(TRUE, FALSE), 10, replace = TRUE)
)

usethis::use_data(sim_dat, overwrite = TRUE)
```

### 7.2 Vignettes

```{r, eval=FALSE}
usethis::use_vignette("getting-started")
# Edit vignettes/getting-started.Rmd, then:
devtools::build_vignettes()
```

### 7.3 Tests with testthat

```{r, eval=FALSE}
usethis::use_testthat()
# Creates tests/testthat.R and tests/testthat/ directory

usethis::use_test("na_counter")
# -> tests/testthat/test-na_counter.R (edit and add tests)
```

Example tests:

```{r, eval=FALSE}
# File: tests/testthat/test-na_counter.R

library(testthat)

test_that("na_counter counts NAs correctly for data.frame and matrix", {
  test_matrix <- matrix(c(NA, 1, 4, NA, 5, 6), nrow = 2)
  air_expected <- c(Ozone = 37, Solar.R = 7, Wind = 0, Temp = 0, Month = 0, Day = 0)
  mat_expected <- c(V1 = 1, V2 = 1, V3 = 0)

  expect_equal(na_counter(airquality), air_expected)
  expect_equal(na_counter(test_matrix), mat_expected)
})

# Non-exported functions can be called with ::: if truly needed
# expect_equal(simutils:::sum_na(airquality$Ozone), 37)
```

Common expectations:

-   `expect_identical()`, `expect_equal(tolerance = ...)`, `expect_equivalent()`
-   `expect_error()`, `expect_warning()`
-   `expect_output()` / `expect_output_file()` for console text

Run the test suite:

```{r eval=FALSE}
devtools::test()
```

## 8. Manage Dependencies

Use helpers to declare dependencies in `DESCRIPTION` (rather than editing by hand):

```{r, eval=FALSE}
usethis::use_package("dplyr")                # add to Imports
usethis::use_package("ggplot2", type = "Suggests")
```

Guidelines:

-   **Imports**: packages needed at runtime for your functions.
-   **Suggests**: optional packages used in examples/tests/vignettes.
-   **Depends**: rarely needed; if used, it attaches packages for users (discouraged).

## 9. Check Your Package Thoroughly

``` {r, eval=FALSE}
# Full check like CRAN does; read 00check.log for details
rc <- devtools::check()
rc
```

Typical findings & quick fixes:

-   **Missing fields** (e.g., `License`) → add via `use_mit_license()` or set in DESCRIPTION.
-   **Undocumented args** → ensure every parameter appears in `@param` and usage matches.
-   **Broken examples** → run your examples; fix typos (e.g., `airquality`, not `airuality`).
-   **Missing system tools** for building PDF manual/vignettes → install LaTeX if you need PDFs.
-   **Failing tests** → run `devtools::test()`; update your code or expectations.

## 10. Build & Install Locally

```{r, eval=FALSE}
# Build source tarball
devtools::build()

# Install the package from source
devtools::install()

# Try it out
#library(simutils)
na_counter(airquality)
```

## 11. Continuous Integration (optional)

You can wire up CI so checks run automatically on each commit. (Any provider works; pick what your org supports.)

```{r, eval=FALSE}
# Example: set up a CI service config (adjust per provider)
# usethis::use_github_action_check_standard()
# or legacy: usethis::use_travis()
```

## 12. Package Website (optional)

```{r, eval=FALSE}
usethis::use_pkgdown()
pkgdown::build_site()
```

## 13. Release Checklist

-   [ ] `R CMD check` passes locally with **0 errors, 0 warnings, 0 notes**
-   [ ] All exported functions documented with examples
-   [ ] Vignettes render without errors
-   [ ] Tests pass on your platforms (and CI)
-   [ ] `NEWS.md` updated; version bumped in `DESCRIPTION`
-   [ ] License added and included

------------------------------------------------------------------------

### Appendix A — Common Files (at a glance)

-   `DESCRIPTION`: metadata & dependencies
-   `NAMESPACE`: auto‑generated exports/imports
-   `R/`: function source files
-   `man/`: help files (`.Rd`) generated by roxygen2
-   `tests/`: unit tests (testthat)
-   `vignettes/`: long‑form docs
-   `data/` & `R/sysdata.rda`: package data

### Appendix B — Troubleshooting Snippets

```{r, eval=FALSE}
# 1) Missing License in check
usethis::use_mit_license("Your Name")

# 2) Undocumented argument in check
# -> Make sure @param includes it and re-run document()

devtools::document(); devtools::check()

# 3) Examples fail in check (typo)
# -> Fix the example code, rerun check

# 4) pdflatex not found during check
# -> Install a LaTeX distribution or run check without manual building

# 5) Dependency not available
# -> Ensure it’s in Imports/Suggests and installed in your environment
```

> Turn this file into your team’s template. Duplicate it, search/replace `simutils`, and you’re ready to scaffold a new package in minutes.
