---
title: "Git in RStudio (Setup & Auth)"
---

## One-Time Setup
- Install Git and ensure `git --version` works.
- In R:
```{r eval=FALSE}
usethis::use_git_config(user.name = "Your Name", user.email = "you@example.com")
```

## Initialize Git for the Current Project
```{r eval=FALSE}
usethis::use_git()
```

## Connect to GitHub
- Create a GitHub account.
- In R:
```{r eval=FALSE}
usethis::create_github_token()
gitcreds::gitcreds_set()  # paste token when prompted
usethis::use_github(protocol = "https")
```
Or set up SSH keys via RStudio (`Tools > Global Options > Git/SVN`).

## Typical Workflow
1. Stage changes (Git pane in RStudio).
2. Commit with a clear message.
3. Push to origin (GitHub).

## Remove Git from a Project (macOS/RStudio)
- In Finder/Terminal, delete the hidden `.git` folder in the project root (careful!).
- Or from Terminal at project root:
```
rm -rf .git
```
- Reopen project in RStudio; Git pane will disappear.

**Exercises**
- Create a new repo for this Quarto course and push it.
- Branch, make a change, open a Pull Request on GitHub.
