---
title: "TLFs: Table, Figure, Listing"
---

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(gt)
library(ggplot2)
library(survival)
library(table1)
library(r2rtf)
library(haven)
library(tidyr)
library(stringr)
library(tools)
```

We reuse `adsl` from the previous chapter (or synthesize if missing).

```{r, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE
)

options(dplyr.summarise.inform = FALSE)

rtf2pdf <- function(input) {
  input <- normalizePath(input)
  x <- "export LD_LIBRARY_PATH=:/usr/lib/libreoffice/program:/usr/lib/x86_64-linux-gnu/"
  y <- paste0("libreoffice --invisible --headless --nologo --convert-to pdf --outdir tlf/ ", input)
  z <- paste(x, y, sep = " && ")
  if (Sys.getenv("GITHUB_ACTIONS") != "") system(z) else invisible(NULL)
}

# Customize data frame and tibble printing methods.
# See <https://github.com/elong0527/r4csr/issues/115> for details.
more_rows <- function(x, n) if (nrow(x) <= n) NULL else paste0("# ℹ ", nrow(x) - n, " more rows")
knit_print.data.frame <- function(x, ...) {
  paste(
    c(
      capture.output(base::print.data.frame(head(x, 4))),
      more_rows(x, 4)
    ),
    collapse = "\n"
  )
}

registerS3method(
  "knit_print", "data.frame", knit_print.data.frame,
  envir = asNamespace("knitr")
)

options(pillar.advice = FALSE)

knit_print.tbl_df <- function(x, ...) {
  paste(capture.output(print(x, n = 4)), collapse = "\n")
}

registerS3method(
  "knit_print", "tbl_df", knit_print.tbl_df,
  envir = asNamespace("knitr")
)


```

```{r}
adsl <- haven::read_sas("data/adam/adsl.sas7bdat")  # adjust path as needed
if (!exists("adsl")) {
  set.seed(123)
  adsl <- tibble::tibble(
    USUBJID = sprintf("XYZ-%03d", 1:60),
    TRT01P = sample(c("Placebo","Active"), 60, replace=TRUE),
    AGE = round(rnorm(60, 60, 8)),
    SEX = sample(c("M","F"), 60, replace=TRUE)
  )
}
```

## Table 1: Baseline Characteristics by Treatment
```{r}
tbl1 <- adsl |>
  dplyr::group_by(TRT01P) |>
  dplyr::summarise(
    N = dplyr::n(),
    mean_age = mean(AGE, na.rm=TRUE),
    sd_age = sd(AGE, na.rm=TRUE),
    pct_female = mean(SEX == "F")*100
  )

gt(tbl1) |>
  tab_header(title = "Table 1. Baseline Characteristics by Treatment") |>
  fmt_number(columns = c(mean_age, sd_age, pct_female), decimals = 1)
```

```{r}
ana <- adsl %>%
  mutate(
    SEX = factor(SEX, c("F", "M"), c("Female", "Male")),
    RACE = toTitleCase(tolower(RACE))
  )

tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = ana)
tbl

```

## Listing: Subject-Level Listing
```{r}
lst <- adsl |>
  arrange(USUBJID) |>
  select(USUBJID, TRT01P, AGE, SEX) |>
  head(20)

gt(lst) |>
  tab_header(title = "Listing: First 20 Subjects")
```

The code below transfer the output into a dataframe
that only contains ASCII characters
recommended by regulatory agencies.
`tbl_base` is used as input for r2rtf to create the final report.

```{r}
tbl_base <- tbl %>%
  as.data.frame() %>%
  as_tibble() %>%
  mutate(across(
    everything(),
    ~ str_replace_all(.x, intToUtf8(160), " ")
  ))


names(tbl_base) <- str_replace_all(names(tbl_base), intToUtf8(160), " ")
tbl_base
```
We define the format of the output. We highlight items that
are not discussed in previous discussion.

`text_indent_first` and `text_indent_left` are used to control the indent
space of text. They are helpful when you need to control the white space
of a long phrase, "AMERICAN INDIAN OR ALASKA NATIVE"
in the table provides an example.

```{r}
colheader1 <- paste(names(tbl_base), collapse = "|")
colheader2 <- paste(tbl_base[1, ], collapse = "|")
rel_width <- c(2.5, rep(1, 4))

tbl_base[-1, ] %>%
  rtf_title(
    "Baseline Characteristics of Participants",
    "(All Participants Randomized)"
  ) %>%
  rtf_colheader(colheader1,
    col_rel_width = rel_width
  ) %>%
  rtf_colheader(colheader2,
    border_top = "",
    col_rel_width = rel_width
  ) %>%
  rtf_body(
    col_rel_width = rel_width,
    text_justification = c("l", rep("c", 4)),
    text_indent_first = -240,
    text_indent_left = 180
  ) %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_base.rtf")
```

```{r, include=FALSE}
#rtf to pdf convert

rtf2pdf("tlf/tlf_base.rtf")
```

```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
if (exists("tlf/tlf_base.pdf")) {
knitr::include_graphics("tlf/tlf_base.pdf")
}
```



In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows:

- Step 1: Read the data set.
- Step 2: Use `table1::table1()` to get the baseline characteristics table.
- Step 3: Transfer the output from Step 2 into a data frame that only contains ASCII characters.
- Step 4: Define the format of the RTF table by using the R package r2rtf.


**Exercises**
1. Format Table 1 to `N (mean ± SD)` for age.
2. Add risk table to the KM plot (use an extension like survminer outside of this minimal example).
3. Create a listing that includes population flags once you derive them.
