---
title: "ggplot2 in R with CDISC ADaM Examples"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(ggplot2)
library(dplyr)
library(tidyr)
library(survival)   # for ADTTE time-to-event example
library(survminer)  # convenient KM plotting (ggplot2-based)
set.seed(123)
```

# Introduction

This tutorial is a **step-by-step guide to ggplot2** in R, designed to take you from **beginner to advanced**, using
**CDISC ADaM-style datasets** throughout.

- `ADSL`: Subject-level data
- `ADLB`: Laboratory data
- `ADVS`: Vital signs (optional example)
- `ADTTE`: Time-to-event data (e.g. PFS)

> These are **simulated examples** with ADaM-like structures and variable names. In a real project you would replace
> them with your actual ADSL/ADLB/ADTTE datasets.

We will cover:

- Basic **Grammar of Graphics** concepts
- Building plots layer by layer with `ggplot()`
- Common geoms: points, lines, bars, histograms, boxplots
- Aesthetics: color, shape, size, linetype
- Faceting, scales, themes, coordinates
- Combining `dplyr` + `ggplot2` in a tidy ADaM workflow
- Advanced: annotations, secondary axes, simple KM plot from ADTTE
- Saving plots for reports/presentations

---

# Creating Simulated ADaM-Style Data

In this section we create small ADaM-like datasets entirely in R so the tutorial is self-contained.

## ADSL: Subject-Level Data

```{r create-adsl}
ADSL <- haven::read_sas("./data/adam/adsl.sas7bdat")

ADSL <- ADSL |> 
       dplyr::mutate(TRTA=ARM)

```

## ADLB: Laboratory Data (e.g. ALT)

We simulate longitudinal ALT values (`PARAMCD = "ALT"`) at multiple visits:

```{r create-adlb}
ADLB <- haven::read_sas("./data/adam/adlbhy.sas7bdat") |> 
              dplyr::filter(PARAMCD == "ALT")
```

## ADTTE: Time-to-Event Data (e.g. PFS)

We simulate a simple ADTTE:

- `PARAMCD = "PFS"`
- `AVAL`: time (months)
- `CNSR`: censor flag (0 = event, 1 = censored)

```{r create-adtte}
ADTTE <- ADSL %>%
  select(STUDYID, USUBJID, ARM,TRTA, SEX, AGE) %>%
  mutate(
    PARAMCD = "PFS",
    PARAM   = "Progression-Free Survival (Months)",
    # Arm-specific exponential rates
    rate = case_when(
      ARM == "Placebo" ~ 0.12,
      ARM == "Dose 1"  ~ 0.09,
      TRUE             ~ 0.07
    ),
    AVAL = rexp(n(), rate = rate),
    AVAL = pmin(AVAL, 36),  # administrative censoring at 36 months
    CNSR = if_else(AVAL >= 36, 1, 0)
  ) %>%
  select(-rate)

ADTTE %>% head()
```

With these datasets we can now explore ggplot2 concepts **in an ADaM context**.

---

# Grammar of Graphics with ADaM

The basic idea of ggplot2 (Grammar of Graphics):

- **Data**: ADSL / ADLB / ADTTE
- **Aesthetics** (`aes`): how variables map to x, y, color, etc.
- **Geoms**: geometric objects (points, lines, bars, etc.)
- **Scales, facets, coordinates, theme**: control look and structure

The general pattern:

```{r, eval=FALSE}
ADSL <- haven::read_sas("./data/adam/adsl.sas7bdat")
ggplot(data = ADSL, aes(x = AGE, y = BMIBL)) +
  geom_point()
```

We will start with subject-level plots from **ADSL**, then move to longitudinal labs from **ADLB**, and finally to **ADTTE**.

---

# Basic Plots from ADSL (Beginner Level)

## Scatterplot: AGE vs BMIBL

```{r adsl-scatter-basic}
ggplot(ADSL, aes(x = AGE, y = BMIBL)) +
  geom_point() +
  labs(
    title = "Scatterplot of Baseline BMI vs Age",
    x = "Age (Years)",
    y = "Baseline BMI (kg/m^2)"
  )
```

### Mapping vs Setting Aesthetics

- **Mapping** (inside `aes()`): color/shape/size depend on data (e.g. `ARM`).
- **Setting** (outside `aes()`): fixed values.

Map `ARM` to color:

```{r adsl-scatter-map-color}
ggplot(ADSL, aes(x = AGE, y = BMIBL, color = TRT01A)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "Baseline BMI vs Age by Treatment Arm",
    color = "Treatment Arm",
    x = "Age (Years)",
    y = "Baseline BMI (kg/m^2)"
  )
```

Set color manually (not based on data):

```{r adsl-scatter-set-color}
ggplot(ADSL, aes(x = AGE, y = BMIBL)) +
  geom_point(color = "steelblue", size = 2.5, alpha = 0.8) +
  labs(
    title = "Baseline BMI vs Age (All Arms)",
    x = "Age (Years)",
    y = "Baseline BMI (kg/m^2)"
  )
```

---

## Bar Plot: Subject Counts per Arm

Using `ADSL`, we can show how many subjects per arm:

```{r adsl-bar-arm}
ggplot(ADSL, aes(x = TRT01A)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Number of Subjects per Treatment Arm",
    x = "Treatment Arm",
    y = "Count of Subjects"
  )
```

If you pre-compute counts (e.g. for TLG), use `geom_col()`:

```{r adsl-bar-arm-summarised}
adsl_counts <- ADSL %>%
  count(TRT01A, name = "n")

adsl_counts
```

```{r adsl-bar-arm-col}
ggplot(adsl_counts, aes(x = TRT01A, y = n)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Number of Subjects per Treatment Arm",
    x = "Treatment Arm",
    y = "Count of Subjects"
  )
```

---

## Histogram and Density: Age Distribution

Histogram of `AGE`:

```{r adsl-age-hist}
ggplot(ADSL, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "lightgreen", color = "white") +
  labs(
    title = "Age Distribution (ADSL)",
    x = "Age (Years)",
    y = "Frequency"
  )
```

Density plot:

```{r adsl-age-density}
ggplot(ADSL, aes(x = AGE)) +
  geom_density(fill = "orange", alpha = 0.6) +
  labs(
    title = "Age Density (ADSL)",
    x = "Age (Years)",
    y = "Density"
  )
```

---

## Boxplot: BMIBL by ARM

Boxplots are useful in CSR or listings to compare distributions across arms.

```{r adsl-bmi-box}
ggplot(ADSL, aes(x = TRT01A, y = BMIBL)) +
  geom_boxplot(fill = "lightgray") +
  labs(
    title = "Baseline BMI by Treatment Arm",
    x = "Treatment Arm",
    y = "Baseline BMI (kg/m^2)"
  )
```

You can combine boxplot with jittered points:

```{r adsl-bmi-box-jitter}
ggplot(ADSL, aes(x = TRT01A, y = BMIBL)) +
  geom_boxplot(fill = "lightgray", outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, color = "blue") +
  labs(
    title = "Baseline BMI by Treatment Arm (with Individual Subjects)",
    x = "Treatment Arm",
    y = "Baseline BMI (kg/m^2)"
  )
```

---

# Longitudinal Plots from ADLB (Intermediate)

Now we use **ADLB** to illustrate time course plots, which are common in clinical reports.

## Line Plot: Mean ALT Over Time by ARM

First summarise ADLB:

```{r adlb-summary}

adlb_alt_summary <- ADLB %>%
  group_by(TRTA, AVISITN,AVISIT) %>%
  summarise(
    mean_ALT = mean(AVAL,na.rm=TRUE),
    sd_ALT   = sd(AVAL,na.rm=TRUE),
    n        = n(),
    se_ALT   = sd_ALT / sqrt(n),
    .groups = "drop"
  )

adlb_alt_summary$AVISIT <- factor(adlb_alt_summary$AVISIT, levels=unique(adlb_alt_summary$AVISIT[order(adlb_alt_summary$AVISITN)]))  
adlb_alt_summary
```

Line plot with error bars:

```{r adlb-line-arm}
ggplot(adlb_alt_summary,
       aes(x = AVISIT, y = mean_ALT, group = TRTA, color = TRTA)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(ymin = mean_ALT - se_ALT, ymax = mean_ALT + se_ALT),
    width = 0.15
  ) +
  labs(
    title = "Mean ALT Over Time by Treatment Arm",
    x = "Visit",
    y = "Mean ALT (U/L)",
    color = "Treatment Arm"
  ) +
  theme_minimal()
```

---

## Individual Profiles (Spaghetti Plot)

To see variability, we can plot individual patients (subset to avoid overcrowding):

```{r adlb-spaghetti}

ADLB_sub <- ADLB |> 
  head(20)
ADLB_sub$AVISIT <- factor(ADLB_sub$AVISIT, levels=unique(ADLB_sub$AVISIT[order(ADLB_sub$AVISITN)]))  
ggplot(ADLB_sub,
       aes(x = AVISIT, y = AVAL, group = USUBJID, color = TRTA)) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.6, size = 1.5) +
  labs(
    title = "Individual ALT Trajectories ",
    x = "Visit",
    y = "ALT (U/L)",
    color = "Arm"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# Faceting with ADaM Data

Faceting is useful when splitting results by subgroups, lab parameters, or regions.

Here we only have `PARAMCD = "ALT"`, but we can pretend we have multiple parameters.

## Example: Facet ALT by Sex

```{r adlb-facet-sex}
ggplot(adlb_alt_summary,
       aes(x = AVISIT, y = mean_ALT, group = TRTA, color = TRTA)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ TRTA) +
  labs(
    title = "Mean ALT Over Time – Faceted by Treatment TRTA",
    x = "Visit",
    y = "Mean ALT (U/L)",
    color = "TRTA"
  ) +
  theme_minimal()
```

You can similarly facet by `SEX`, `RACE`, or any other ADSL variable once merged into ADLB.

---

# Scales, Labels, and Themes in ADaM Context

## Changing Axis Labels and Titles

```{r adsl-scatter-labels}
ggplot(ADSL, aes(x = AGE, y = BMIBL, color = TRT01A)) +
  geom_point(alpha = 0.8) +
  labs(
    title = "Baseline BMI vs Age by Treatment TRTA",
    subtitle = "Simulated ADaM ADSL Data",
    x = "Age (Years)",
    y = "Baseline BMI (kg/m^2)",
    color = "Treatment TRTA"
  )
```

## Controlling Color Scales

For discrete `TRTA`:

```{r adsl-color-brewer}
ggplot(ADSL, aes(x = AGE, y = BMIBL, color = TRT01A)) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Custom Color Scale for TRTA",
    color = "TRTA"
  ) +
  theme_minimal()
```

For continuous values (e.g. ALT):

```{r adlb-color-gradient}
ggplot(ADLB, aes(x = AGE, y = AVAL, color = AVAL)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "lightyellow", high = "darkred") +
  labs(
    title = "ALT vs Age (Colored by ALT)",
    x = "Age (Years)",
    y = "ALT (U/L)",
    color = "ALT"
  ) +
  theme_minimal()
```

## Transforming Scales (e.g., log10 ALT)

```{r adlb-log-scale}
ggplot(ADLB, aes(x = AGE, y = AVAL, color = TRTA)) +
  geom_point(alpha = 0.7) +
  scale_y_log10() +
  labs(
    title = "ALT vs Age (Log10 ALT)",
    x = "Age (Years)",
    y = "ALT (U/L) – log10 scale",
    color = "TRTA"
  ) +
  theme_minimal()
```

---

# Themes and Publication-Style Graphics

## Using Built-In Themes

```{r theme-examples}
p_base <- ggplot(ADSL, aes(x = AGE, y = BMIBL, color = TRT01A)) +
  geom_point(size = 2.2, alpha = 0.8) +
  labs(
    title = "Baseline BMI vs Age by TRTA",
    x = "Age (Years)",
    y = "Baseline BMI"
  )

p_base + theme_bw()
```

```{r theme-minimal}
p_base + theme_minimal()
```

```{r theme-classic}
p_base + theme_classic()
```

## Custom Theme for Clinical Reports

```{r theme-adam-pub}
theme_adam_pub <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 13),
      axis.title = element_text(face = "bold"),
      panel.grid.major = element_line(color = "grey85"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )
}

ADLB$AVISIT <- factor(ADLB$AVISIT, levels=unique(ADLB$AVISIT[order(ADLB$AVISITN)])) 
ggplot(ADLB, aes(x = AVISIT, y = AVAL, color = TRTA)) +
  geom_boxplot() +
  labs(
    title = "ALT Distribution by Visit and TRTA",
    x = "Visit",
    y = "ALT (U/L)",
    color = "TRTA"
  ) +
  theme_adam_pub()
```

---

# Coordinates and Positions

## Flipping Coordinates: Boxplot of BMI by Sex

```{r coord-flip}
ggplot(ADSL, aes(x = SEX, y = BMIBL)) +
  geom_boxplot(fill = "lightblue") +
  coord_flip() +
  labs(
    title = "Baseline BMI by Sex",
    x = "Sex",
    y = "Baseline BMI"
  )
```

## Stacked vs Dodged Bar Plots: TRTA by SEX

```{r bar-stacked}
ggplot(ADSL, aes(x = TRTA, fill = SEX)) +
  geom_bar(position = "stack") +
  labs(
    title = "Subjects by Treatment and Sex (Stacked)",
    x = "TRTA",
    y = "Count",
    fill = "Sex"
  )
```

```{r bar-dodged}
ggplot(ADSL, aes(x = TRTA, fill = SEX)) +
  geom_bar(position = position_dodge(width = 0.8)) +
  labs(
    title = "Subjects by Treatment and Sex (Dodged)",
    x = "TRTA",
    y = "Count",
    fill = "Sex"
  )
```

---

# Adding Statistical Layers

## Linear Regression: BMIBL vs AGE

```{r stat-smooth-lm}
ggplot(ADSL, aes(x = AGE, y = BMIBL)) +
  geom_point(alpha = 0.7) +
  stat_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Regression of Baseline BMI on Age",
    x = "Age (Years)",
    y = "Baseline BMI"
  ) +
  theme_minimal()
```

## Summaries: Mean ALT ± SD per Visit and TRTA

We already computed `adlb_alt_summary`. Plot mean ± SD:

```{r adlb-mean-sd}
ggplot(adlb_alt_summary,
       aes(x = AVISIT, y = mean_ALT, color = TRTA, group = TRTA)) +
  geom_point(size = 2.5) +
  geom_line() +
  geom_errorbar(
    aes(ymin = mean_ALT - sd_ALT, ymax = mean_ALT + sd_ALT),
    width = 0.2, alpha = 0.7
  ) +
  labs(
    title = "Mean ALT ± SD by Visit and TRTA",
    x = "Visit",
    y = "ALT (U/L)",
    color = "TRTA"
  ) +
  theme_minimal()
```

---

# Tidy Workflow: dplyr + ggplot2 with ADaM

A very common pattern in clinical reporting:

1. Start from ADaM dataset
2. Filter / summarise with `dplyr`
3. Plot with `ggplot2`

Example: Mean BMI by TRTA and SEX:

```{r tidy-workflow}
ADSL %>%
  group_by(TRTA, SEX) %>%
  summarise(
    mean_BMIBL = mean(BMIBL),
    sd_BMIBL   = sd(BMIBL),
    n          = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = TRTA, y = mean_BMIBL, fill = SEX)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_errorbar(
    aes(ymin = mean_BMIBL - sd_BMIBL, ymax = mean_BMIBL + sd_BMIBL),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  labs(
    title = "Mean Baseline BMI by Treatment TRTA and Sex",
    x = "Treatment TRTA",
    y = "Mean Baseline BMI",
    fill = "Sex"
  ) +
  theme_adam_pub()
```

---

# Advanced: Simple Kaplan–Meier Plot from ADTTE

In ADaM, **ADTTE** contains time-to-event information (e.g. PFS, OS).

Here we show a simple KM plot using `survminer::ggsurvplot()` which uses ggplot2 under the hood.

```{r adtte-km}
adtte_pfs <- ADTTE %>%
  filter(PARAMCD == "PFS")

fit_pfs <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_pfs)

km_plot <- ggsurvplot(
  fit_pfs,
  data = adtte_pfs,
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.25,
  ggtheme = theme_minimal(),
  legend.title = "TRTA",
  legend.labs = levels(factor(adtte_pfs$TRTA)),
  title = "Kaplan–Meier Curve for PFS by Treatment TRTA",
  xlab = "Time (Months)",
  ylab = "Progression-Free Survival Probability"
)

km_plot
```

For a **pure ggplot2** approach, you could extract the `survfit` summary into a data frame and use `geom_step()`
manually, but `ggsurvplot()` is convenient and still fully compatible with ggplot theming.

---

# Advanced: Annotations and Secondary Axes

## Annotations on an ADaM Plot

Example: highlight older patients with high BMI:

```{r annotation-example}
ggplot(ADSL, aes(x = AGE, y = BMIBL)) +
  geom_point(alpha = 0.7) +
  annotate(
    "rect", xmin = 70, xmax = 80,
    ymin = 30, ymax = 40,
    alpha = 0.1, fill = "red"
  ) +
  annotate(
    "text", x = 75, y = 41,
    label = "Older, higher BMI
(Example region)",
    size = 3, color = "red"
  ) +
  labs(
    title = "Annotations on ADSL Scatterplot",
    x = "Age (Years)",
    y = "Baseline BMI"
  ) +
  theme_minimal()
```

## Secondary Axis (Use with Care)

ggplot2 allows secondary axes only if they are a simple transformation of the primary axis.

Example (purely illustrative) converting ALT to approximate multiple:

```{r secondary-axis}
ggplot(adlb_alt_summary, aes(x = AVISIT, y = mean_ALT, group = TRTA, color = TRTA)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(
    name = "ALT (U/L)",
    sec.axis = sec_axis(~ . / 40, name = "ALT (x Upper Normal Limit)")
  ) +
  labs(
    title = "ALT Mean Over Time with Secondary Axis",
    x = "Visit",
    color = "TRTA"
  ) +
  theme_minimal()
```

> In production, ensure the secondary axis has a clinically correct and meaningful transformation.

---

# Writing Reusable Plot Functions for ADaM

You can encapsulate standard plots (e.g., lab time course by TRTA) into functions.

```{r adam-plot-function}
plot_lab_timecourse <- function(adlb_df, paramcd = "ALT") {
  df <- adlb_df %>%
    filter(PARAMCD == paramcd) %>%
    group_by(TRTA, AVISIT) %>%
    summarise(
      mean_aval = mean(AVAL),
      se_aval   = sd(AVAL) / sqrt(n()),
      .groups = "drop"
    )

  ggplot(df, aes(x = AVISIT, y = mean_aval, color = TRTA, group = TRTA)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    geom_errorbar(
      aes(ymin = mean_aval - se_aval, ymax = mean_aval + se_aval),
      width = 0.15
    ) +
    labs(
      title = paste("Mean", paramcd, "Over Time by TRTA"),
      x = "Visit",
      y = paste("Mean", paramcd),
      color = "TRTA"
    ) +
    theme_adam_pub()
}

plot_lab_timecourse(ADLB, paramcd = "ALT")
```

This approach is very powerful for building **standardized plotting utilities** for your ADaM pipeline.

---

# Saving Plots for Reports

Use `ggsave()` to save plots for use in RTF, Word, or PowerPoint.

```{r save-plot, eval=FALSE}
p <- ggplot(ADSL, aes(x = AGE, y = BMIBL, color = TRTA)) +
  geom_point() +
  labs(
    title = "Baseline BMI vs Age by TRTA",
    x = "Age (Years)",
    y = "Baseline BMI"
  ) +
  theme_adam_pub()

# Save as PNG
ggsave("bmi_vs_age_by_TRTA.png", plot = p, width = 6, height = 4, dpi = 300)

# Save as PDF
ggsave("bmi_vs_age_by_TRTA.pdf", plot = p, width = 6, height = 4)
```


```{r, include=FALSE}
# Core packages
library(ggplot2)
library(dplyr)
library(tidyr)

# For KM and forest plots
library(survival)
library(survminer)
library(grid)
library(forestploter)
```

# Introduction

This document walks through **six common oncology graphs** step by step, using the exact code you provided:

1. Bar graph of **sum of lesion diameters** plus individual lesions  
2. **Waterfall plot** of best percentage change from baseline  
3. **Swimmer plot** for duration of response  
4. **Spider plot** (tumor trajectories over time)  
5. **Kaplan–Meier** survival curve  
6. **Forest plot** for subgroup hazard ratios  

For each figure, we will:

- Build a simple **example dataset** in R (similar structure to what you’d get from ADaM like ADTR/ADTTE).
- Explain the main **ggplot2 layers** and options.
- Show the **final code** to generate the plot.

In a real clinical project, you would replace the toy data here with your ADaM datasets, but the plotting logic stays the same.

---

# Figure 1 – Bar Graph with Individual Lesion Lines

This plot shows, for a single subject:

- Bars: **Sum of Lesion Diameters (SUMLD)** at each visit  
- Lines/points: each **individual lesion** (tumor) over time  

Clinically, this helps you see how the total tumor burden and individual lesions evolve across visits.

## 1.1 Create bar-graph data

We first build a small dataset with:

- `visit`: visit label
- `SUMLD`: sum of lesion diameters
- `trstresn_tumor1/2/3`: individual lesion diameters

```{r fig1-data}
bar_data <- data.frame(
  visit = c("Baseline", "Week 4", "Week 8", "Week 12", "Week 16"),
  SUMLD = c(100, 85, 70, 65, 60),  # Sum of Lesion Diameters
  trstresn_tumor1 = c(50, 45, 35, 30, 28),
  trstresn_tumor2 = c(30, 25, 20, 20, 18),
  trstresn_tumor3 = c(20, 15, 15, 15, 14)
)

bar_data
```

In real ADaM:

- `visit` could map to `AVISIT`,
- `SUMLD` would be a derived variable (sum of target lesions),
- `trstresn_tumorX` are separate lesions (often separate rows in ADTR; here we keep them as columns for simplicity).

## 1.2 Reshape to long format for lines

To draw multiple lesions as separate lines, we reshape the wide columns (`trstresn_tumor1/2/3`) into a long format using `pivot_longer()`:

```{r fig1-long}
bar_data_long <- bar_data %>%
  pivot_longer(
    cols = starts_with("trstresn"), 
    names_to = "tuloc", 
    values_to = "trstresn"
  )

bar_data_long
```

- `tuloc` acts like **tumor location / lesion identifier**.  
- `trstresn` holds the lesion diameter for each visit–lesion combination.

## 1.3 Build the bar + line plot

Now we combine:

- `geom_col()` for SUMLD (bars)
- `geom_line()` and `geom_point()` for individual lesions
- `scale_y_continuous()` with a **secondary axis** label (note: both axes use same numeric scale, only labels differ)

```{r fig1-plot, fig.width=7, fig.height=5}
p1 <- ggplot() +
  # Bars = Sum of lesion diameters per visit
  geom_col(
    data = bar_data, 
    aes(x = visit, y = SUMLD), 
    fill = "#4472C4", 
    width = 0.6,
    alpha = 0.7
  ) +
  # Lines = individual lesions
  geom_line(
    data = bar_data_long, 
    aes(x = visit, y = trstresn, 
        group = tuloc, color = tuloc),
    size = 1.2
  ) +
  # Points = individual lesion measurements
  geom_point(
    data = bar_data_long, 
    aes(x = visit, y = trstresn, 
        color = tuloc),
    size = 3
  ) +
  # Primary and secondary y-axes (same numeric scale)
  scale_y_continuous(
    name = "Lesion Diameter (mm)",
    sec.axis = sec_axis(~., name = "Sum of Lesion Diameters (mm)")
  ) +
  labs(
    title = "Figure 1
Tumor Response in Individual Patients with Measurable Disease",
    subtitle = "(ITT Population)
Subject: 1001",
    x = "Visit",
    color = "Tumor Location"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.title = element_text(face = "bold", size = 11),
    legend.position = "bottom"
  )

p1
```

Key points:

- We used an **empty `ggplot()`** and passed data separately to each layer.  
- This is useful when we have different data frames for bars (`bar_data`) and lines (`bar_data_long`).

---

# Figure 2 – Waterfall Plot of Best Percentage Change from Baseline

Waterfall plots show each patient’s **best % change from baseline** in tumor size.  
They’re very common in oncology to visualize response categories like CR/PR/SD/PD.

## 2.1 Create waterfall data

We simulate:

- `patientid`: patient identifier
- `pchg`: best % change from baseline in sum of diameters
- `response`: categorical RECIST-like response (CR/PR/SD/PD)

```{r fig2-data}
waterfall_data <- data.frame(
  patientid = paste0("PT", sprintf("%03d", 1:30)),
  pchg = c(-65, -55, -48, -42, -38, -35, -30, -28, -25, -22,
           -18, -15, -12, -10, -8, -5, -2, 5, 8, 12,
           15, 20, 25, 30, 35, 40, 45, 50, 55, 60),
  response = c(rep("CR", 2), rep("PR", 13), rep("SD", 10), rep("PD", 5))
)

waterfall_data
```

## 2.2 Sort patients by change

To get the typical waterfall shape (bars sorted from best to worst response), we sort by `pchg` and then set factor levels:

```{r fig2-sort}
waterfall_data <- waterfall_data %>%
  arrange(pchg) %>%
  mutate(patientid = factor(patientid, levels = patientid))

response_colors <- c(
  "CR" = "#00B050",
  "PR" = "#92D050", 
  "SD" = "#FFC000",
  "PD" = "#FF0000"
)

waterfall_data
```

## 2.3 Build the waterfall plot

- `geom_col()` for bars  
- `geom_hline()` to draw RECIST cut-off lines (e.g. -30%, +20%)  
- `coord_cartesian()` to fix y-axis limits  

```{r fig2-plot, fig.width=7, fig.height=5}
p2 <- ggplot(waterfall_data, aes(x = patientid, y = pchg, fill = response)) +
  geom_col(width = 0.8) +
  # Baseline line at 0% change
  geom_hline(yintercept = 0, linetype = "solid", color = "black", size = 0.8) +
  # Typical RECIST thresholds
  geom_hline(yintercept = -30, linetype = "dashed", color = "blue", size = 0.6) +
  geom_hline(yintercept = 20, linetype = "dashed", color = "red", size = 0.6) +
  scale_fill_manual(values = response_colors) +
  labs(
    title = "Figure 2
Waterfall Plot of Best Percentage Change from Baseline",
    subtitle = "in Target Lesion per RECIST 1.1",
    x = "Patient ID",
    y = "Best % Change from Baseline in Sum of Diameters",
    fill = "Response"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.title = element_text(face = "bold", size = 11),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) +
  coord_cartesian(ylim = c(-80, 80))

p2
```

In real ADaM:

- You’d compute `pchg` from ADTR using baseline and best post-baseline sum of diameters.

---

# Figure 3 – Swimmer Plot: Duration of Response

Swimmer plots show **time on treatment / response** for each patient:

- Horizontal bar per patient: **treatment duration**  
- Color: response category  
- Symbols/markers: progression, death, etc.

## 3.1 Create swimmer data

We simulate:

- `start_time` / `end_time`: treatment/response duration (months)
- `response`: CR/PR/SD/PD
- `progression` / `death`: logical indicators

```{r fig3-data}
set.seed(123)  # for reproducibility

swimmer_data <- data.frame(
  patientid = paste0("PT", sprintf("%03d", 1:20)),
  start_time = rep(0, 20),
  end_time = c(12, 15, 8, 20, 18, 14, 10, 22, 16, 19,
               7, 25, 13, 17, 11, 9, 21, 24, 15, 18),
  response = sample(c("CR", "PR", "SD", "PD"), 20, replace = TRUE),
  progression = c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE,
                  TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE),
  death = c(FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE,
            FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE)
)

# Sort by duration so longest bars at top
swimmer_data <- swimmer_data %>%
  arrange(desc(end_time)) %>%
  mutate(patientid = factor(patientid, levels = patientid))

swimmer_data
```

## 3.2 Build the swimmer plot

- `geom_segment()` draws horizontal bars  
- Triangles (`shape = 17`) mark progression  
- Crosses (`shape = 4`) mark death  

```{r fig3-plot, fig.width=7, fig.height=5}
p3 <- ggplot(swimmer_data, aes(y = patientid)) +
  # Treatment/response duration
  geom_segment(
    aes(x = start_time, xend = end_time, 
        yend = patientid, color = response),
    size = 3
  ) +
  # ▲ progression (black triangle)
  geom_point(
    data = swimmer_data %>% filter(progression),
    aes(x = end_time), 
    shape = 17, size = 4, color = "black"
  ) +
  # ✕ death (red cross)
  geom_point(
    data = swimmer_data %>% filter(death),
    aes(x = end_time), 
    shape = 4, size = 4, color = "red", stroke = 2
  ) +
  scale_color_manual(values = c(
    "CR" = "#00B050", "PR" = "#92D050",
    "SD" = "#FFC000", "PD" = "#FF0000"
  )) +
  labs(
    title = "Figure 3
Swimmer Plot - Duration of Response",
    subtitle = "Individual Patient Treatment Duration and Response",
    x = "Time (Months)",
    y = "Patient ID",
    color = "Response"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.title = element_text(face = "bold", size = 11),
    legend.position = "bottom"
  ) +
  annotate(
    "text", x = 26, y = 21, 
    label = "▼ = Progression
✕ = Death", 
    hjust = 0, size = 3
  )

p3
```

In an ADaM workflow, these durations often come from ADTTE / custom response datasets (first response date, progression date, death date, etc.).

---

# Figure 4 – Spider Plot: Tumor Trajectories Over Time

Spider plots show multiple **patients’ tumor trajectories** on the same axes:

- x-axis: time (e.g. months from baseline scan)  
- y-axis: relative tumor size (ratio vs baseline)  

Each line represents one patient.

## 4.1 Create spider data

We create data for 15 patients, with tumor size values over time:

```{r fig4-data}
set.seed(123)

spider_data <- expand.grid(
  patientid = paste0("PT", sprintf("%03d", 1:15)),
  dur = seq(-10, 100, by = 10)
) %>%
  group_by(patientid) %>%
  mutate(
    # Simple random trend around baseline = 1
    size = 1 + rnorm(1, 0, 0.3) + (dur/100) * rnorm(1, -0.5, 0.4)
  ) %>%
  ungroup()

spider_data
```

- At `dur = 0` we are around `size = 1` (baseline).  
- Later timepoints can go up (> 1) or down (< 1).

## 4.2 Build the spider plot

```{r fig4-plot, fig.width=7, fig.height=5}
p4 <- ggplot(
  spider_data, 
  aes(
    x = dur, y = size, 
    group = patientid, 
    color = patientid
  )
) +
  geom_line(size = 1.2, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.8) +
  # Horizontal line at baseline (1.0)
  geom_hline(
    yintercept = 1.0, linetype = "solid", 
    color = "black", size = 0.8
  ) +
  scale_y_continuous(breaks = seq(0, 2, by = 0.20)) +
  scale_x_continuous(breaks = seq(-100, 100, by = 10)) +
  labs(
    title = "Figure 4
Relative Change in Tumor Size",
    x = "Months from Baseline Scan (Time 0)",
    y = "Tumor Size Relative to Baseline"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 11),
    legend.position = "none"
  ) +
  coord_cartesian(xlim = c(-10, 100), ylim = c(0, 2))

p4
```

In practice, this would also come from **ADTR** (longitudinal tumor measurements) with a derived `relative_size = AVAL / BASE`.

---

# Figure 5 – Kaplan–Meier Survival Curve

Kaplan–Meier (KM) plots visualize **time-to-event endpoints** like PFS or OS:

- Step function for survival probability over time  
- Separate curve for each treatment arm  
- Optionally: confidence intervals and number-at-risk table  

## 5.1 Create KM data

We simulate:

- `time`: event/censoring time  
- `status`: 1 = event, 0 = censored  
- `treatment`: Treatment A / Treatment B  

```{r fig5-data}
set.seed(456)

km_data <- data.frame(
  time = c(rexp(100, 0.05), rexp(100, 0.03)),
  status = c(rbinom(100, 1, 0.7), rbinom(100, 1, 0.6)),
  treatment = rep(c("Treatment A", "Treatment B"), each = 100)
)

head(km_data)
```

## 5.2 Fit the KM model and plot

We use `survival::survfit()` and `survminer::ggsurvplot()`:

```{r fig5-plot, fig.width=7, fig.height=6}
km_fit <- survfit(Surv(time, status) ~ treatment, data = km_data)

p5 <- ggsurvplot(
  km_fit,
  data = km_data,
  pval = TRUE,           # log-rank p-value
  conf.int = TRUE,       # confidence interval bands
  risk.table = TRUE,     # number at risk under the plot
  risk.table.height = 0.25,
  ggtheme = theme_minimal(),
  palette = c("#E7B800", "#2E9FDF"),
  title = "Figure 5
Kaplan-Meier Survival Curve",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  legend.title = "Treatment Group",
  legend.labs = c("Treatment A", "Treatment B"),
  font.main = c(14, "bold"),
  font.x = c(11, "bold"),
  font.y = c(11, "bold"),
  font.legend = c(10)
)

p5
```

In ADaM:

- This would typically be based on **ADTTE**, using variables like `AVAL` (time), `CNSR` (censor flag), and `TRT01A` or similar.

---

# Figure 6 – Forest Plot for Subgroup Hazard Ratios

Forest plots summarize **effect estimates (e.g. hazard ratios)** across subgroups:

- Each row: one subgroup  
- Point: HR  
- Horizontal line: 95% CI  
- Vertical reference line at HR = 1  

Here we use the `forestploter` package and its example data.

## 6.1 Load example data and prepare

```{r fig6-data}
dt <- read.csv(system.file("extdata", "example_data.csv", package = "forestploter"))

head(dt)
```

We:

- Indent subgroup labels when there is a number in `Placebo` column (for visual hierarchy).  
- Replace `NA` with empty strings for display columns.  
- Compute standard error (`se`) from CI.  
- Add a blank column (`" "`) for the plot.  
- Create a text column for **HR (95% CI)**.

```{r fig6-prep}
# Indent subgroup if there is a number in the placebo column
dt$Subgroup <- ifelse(
  is.na(dt$Placebo), 
  dt$Subgroup,
  paste0("   ", dt$Subgroup)
)

# NA to blank for display columns
dt$Treatment <- ifelse(is.na(dt$Treatment), "", dt$Treatment)
dt$Placebo   <- ifelse(is.na(dt$Placebo), "", dt$Placebo)

# Standard error derived from CI on log scale
dt$se <- (log(dt$hi) - log(dt$est)) / 1.96

# Blank column used by forestploter for CI area
dt$` ` <- paste(rep(" ", 20), collapse = " ")

# Display column for HR (95% CI)
dt$`HR (95% CI)` <- ifelse(
  is.na(dt$se), "",
  sprintf("%.2f (%.2f to %.2f)", dt$est, dt$low, dt$hi)
)

head(dt)
```

## 6.2 Define forest theme and build the plot

We define a theme (fonts, refline color, arrow labels) and then call `forest()`:

```{r fig6-plot, fig.width=7, fig.height=6}
tm <- forest_theme(
  base_size = 10,
  refline_col = "red",
  arrow_type = "closed",
  footnote_gp = gpar(col = "blue", cex = 0.6)
)

p <- forest(
  dt[, c(1:3, 20:21)],
  est     = dt$est,
  lower   = dt$low, 
  upper   = dt$hi,
  sizes   = dt$se,
  ci_column = 4,
  ref_line  = 1,
  arrow_lab = c("Placebo Better", "Treatment Better"),
  xlim      = c(0, 4),
  ticks_at  = c(0.5, 1, 2, 3),
  footnote  = "This is the demo data. Please feel free to change
anything you want.",
  theme     = tm
)

plot(p)
```

Interpretation:

- Values **< 1** favor **treatment**, values **> 1** favor **placebo** (depending on how you define HR).  
- The arrow labels at the bottom make this direction explicit.

In real analyses, the columns `est`, `low`, `hi` would be generated from Cox models or other regressions per subgroup.

---

# Summary and Next Steps

In this tutorial, you:

- Learned the **ggplot2 grammar of graphics** using **ADSL, ADLB, and ADTTE** style data
- Built common clinical plots:
  - Subject counts by arm
  - Histograms/boxplots of demographics
  - Longitudinal lab plots with summary statistics
  - Simple KM curves from ADTTE
- Customized scales, themes, coordinates, and annotations
- Combined `dplyr` + `ggplot2` for tidy ADaM workflows
- Wrote a small **reusable plotting function** tailored to ADaM

From here, you can:

- Swap the simulated data with your real **ADaM datasets**
- Extend functions for specific TLFs (e.g., safety labs, efficacy endpoints)
- Integrate these plots into **R Markdown/Quarto CSRs** or **Shiny dashboards** for interactive review.

Happy plotting with ggplot2 and ADaM!
