---
title: "loops"
---

# Introduction to Loops in  R Programming

Loops are fundamental programming constructs that allow us to execute code repeatedly.  In pharmaceutical data science and research using R, loops are essential for processing clinical trial data, analyzing drug interactions, calculating pharmacokinetic parameters, and automating repetitive analytical tasks. 

## Types of Loops in R

### 1. For Loops

For loops iterate over a sequence of items for a predetermined number of times. 

#### Basic Syntax (R)
```r
for (variable in sequence) {
  # code to execute
}
```

#### Example 1: Processing Patient Data
```{r}
# Patient IDs in a clinical trial
patient_ids <- c("PT001", "PT002", "PT003", "PT004", "PT005")

# Processing each patient
for (patient_id in patient_ids) {
  cat("Processing data for patient:", patient_id, "\n")
  # Simulate data processing
  cat("  - Vital signs checked\n")
  cat("  - Adverse events recorded\n")
  cat("  - Drug concentration measured\n")
  cat("\n")
}
```

#### Example 2: Drug Dosage Calculations
```{r}
# Different patient weights (kg)
patient_weights <- c(65, 72, 58, 80, 95, 68)

# Drug dosage:  5 mg/kg
dose_per_kg <- 5

cat("Patient Dosage Calculations:\n")
cat(strrep("-", 40), "\n")

for (i in seq_along(patient_weights)) {
  weight <- patient_weights[i]
  dosage <- weight * dose_per_kg
  cat(sprintf("Patient %d: %dkg → %dmg\n", i, weight, dosage))
}
```

#### Example 3: Pharmacokinetic Parameter Calculation
```{r}
# Drug concentration over time (mg/L)
concentrations <- c(100, 87, 75, 65, 57, 49, 43)
time_points <- c(0, 1, 2, 3, 4, 5, 6)  # hours

cat("Pharmacokinetic Analysis:\n")
cat(strrep("-", 30), "\n")

# Calculate half-life using multiple time points
initial_conc <- concentrations[1]
target_conc <- initial_conc / 2

for (i in 2:length(concentrations)) {
  if (concentrations[i] <= target_conc) {
    # Linear interpolation for half-life
    t_half <- time_points[i-1] + 
              (time_points[i] - time_points[i-1]) * 
              (target_conc - concentrations[i-1]) / 
              (concentrations[i] - concentrations[i-1])
    
    cat(sprintf("Estimated half-life: %. 2f hours\n", t_half))
    break
  }
}
```

#### Example 4: Creating Data Frame with Loop
```{r}
# Create a comprehensive patient dataset
patients <- data.frame(
  patient_id = character(0),
  age = numeric(0),
  weight = numeric(0),
  dose = numeric(0),
  bmi = numeric(0),
  stringsAsFactors = FALSE
)

# Patient data
patient_data <- list(
  list(id = "PT001", age = 45, weight = 70, height = 1.75),
  list(id = "PT002", age = 32, weight = 65, height = 1.68),
  list(id = "PT003", age = 58, weight = 80, height = 1.82),
  list(id = "PT004", age = 29, weight = 55, height = 1.60),
  list(id = "PT005", age = 67, weight = 90, height = 1.78)
)

# Process each patient and add to data frame
for (i in seq_along(patient_data)) {
  patient <- patient_data[[i]]
  
  # Calculate BMI
  bmi <- patient$weight / (patient$height^2)
  
  # Calculate dose (5mg/kg, max 400mg)
  dose <- min(patient$weight * 5, 400)
  
  # Add row to data frame
  patients[i, ] <- list(
    patient_id = patient$id,
    age = patient$age,
    weight = patient$weight,
    dose = dose,
    bmi = bmi
  )
}

print(patients)
```

### 2. While Loops

While loops continue executing as long as a condition remains true. 

#### Basic Syntax (R)
```r
while (condition) {
  # code to execute
}
```

#### Example 1: Drug Titration Protocol
```{r}
# Initial dose and target concentration
current_dose <- 10  # mg
target_concentration <- 25  # mg/L
current_concentration <- 8  # mg/L
bioavailability <- 0.8
volume_distribution <- 5  # L

day <- 1

cat("Drug Titration Protocol:\n")
cat(strrep("-", 35), "\n")

while (current_concentration < target_concentration && day <= 14) {
  cat(sprintf("Day %d:\n", day))
  cat(sprintf("  Current dose: %. 1fmg\n", current_dose))
  cat(sprintf("  Current concentration: %.1fmg/L\n", current_concentration))
  
  # Increase dose by 20% if below target
  if (current_concentration < target_concentration) {
    current_dose <- current_dose * 1.2
    # Simulate concentration response
    current_concentration <- (current_dose * bioavailability) / volume_distribution
  }
  
  day <- day + 1
}

cat(sprintf("\nFinal dose achieved: %.1fmg\n", current_dose))
cat(sprintf("Final concentration: %.1fmg/L\n", current_concentration))
```

#### Example 2: Quality Control Testing
```{r}
# Set seed for reproducible results
set.seed(123)

# Quality control for drug batch
batch_size <- 1000
tested_units <- 0
defective_units <- 0
max_defects_allowed <- 5  # 0. 5% defect rate

cat("Quality Control Testing:\n")
cat(strrep("-", 30), "\n")

while (tested_units < batch_size && defective_units <= max_defects_allowed) {
  tested_units <- tested_units + 1
  
  # Simulate testing (5% chance of defect)
  is_defective <- runif(1) < 0.05
  
  if (is_defective) {
    defective_units <- defective_units + 1
    cat(sprintf("Unit %d:  DEFECTIVE (Total defects: %d)\n", 
               tested_units, defective_units))
  }
  
  # Report every 100 tests
  if (tested_units %% 100 == 0) {
    cat(sprintf("Tested %d units, %d defects found\n", 
               tested_units, defective_units))
  }
}

if (defective_units > max_defects_allowed) {
  cat(sprintf("\nBATCH REJECTED: Too many defects (%d)\n", defective_units))
} else {
  cat(sprintf("\nBATCH APPROVED: %d defects in %d units\n", 
             defective_units, tested_units))
}
```

### 3. Nested Loops

Nested loops are loops within loops, useful for processing multi-dimensional data.

#### Example 1: Clinical Trial Multi-Site Analysis
```{r}
# Clinical trial data:  sites and their patients
trial_sites <- list(
  Site_A = c("PT001", "PT002", "PT003"),
  Site_B = c("PT004", "PT005", "PT006", "PT007"),
  Site_C = c("PT008", "PT009")
)

# Adverse events per patient (simulated)
adverse_events <- list(
  PT001 = c("Headache", "Nausea"),
  PT002 = c("Fatigue"),
  PT003 = character(0),
  PT004 = c("Dizziness", "Nausea"),
  PT005 = c("Headache"),
  PT006 = c("Fatigue", "Insomnia"),
  PT007 = character(0),
  PT008 = c("Nausea"),
  PT009 = c("Headache", "Dizziness")
)

cat("Clinical Trial Adverse Events Summary:\n")
cat(strrep("=", 45), "\n")

total_patients <- 0
total_events <- 0

# Outer loop:  iterate through sites
for (site_name in names(trial_sites)) {
  patients <- trial_sites[[site_name]]
  cat(sprintf("\n%s:\n", site_name))
  cat(strrep("-", 20), "\n")
  site_events <- 0
  
  # Inner loop: iterate through patients at each site
  for (patient in patients) {
    patient_events <- adverse_events[[patient]]
    if (is.null(patient_events)) patient_events <- character(0)
    
    site_events <- site_events + length(patient_events)
    total_events <- total_events + length(patient_events)
    
    if (length(patient_events) > 0) {
      cat(sprintf("  %s: %s\n", patient, paste(patient_events, collapse = ", ")))
    } else {
      cat(sprintf("  %s: No adverse events\n", patient))
    }
  }
  
  total_patients <- total_patients + length(patients)
  cat(sprintf("  Site total: %d events in %d patients\n", 
             site_events, length(patients)))
}

cat(sprintf("\nTrial Summary:\n"))
cat(sprintf("Total patients: %d\n", total_patients))
cat(sprintf("Total adverse events: %d\n", total_events))
cat(sprintf("Average events per patient: %.2f\n", total_events/total_patients))
```

#### Example 2: Drug Interaction Matrix
```{r}
# Common drugs in cardiology
drugs <- c("Aspirin", "Metoprolol", "Lisinopril", "Atorvastatin", "Warfarin")

# Interaction severity (0=None, 1=Mild, 2=Moderate, 3=Severe)
interaction_matrix <- matrix(c(
  0, 1, 0, 1, 3,  # Aspirin
  1, 0, 2, 0, 1,  # Metoprolol
  0, 2, 0, 0, 2,  # Lisinopril
  1, 0, 0, 0, 1,  # Atorvastatin
  3, 1, 2, 1, 0   # Warfarin
), nrow = 5, byrow = TRUE)

severity_labels <- c("None", "Mild", "Moderate", "Severe")

cat("Drug Interaction Matrix:\n")
cat(strrep("=", 50), "\n")

# Print header
cat(sprintf("%-12s", "Drug"))
for (drug in drugs) {
  cat(sprintf("%-12s", drug))
}
cat("\n")

# Nested loops to print the matrix
for (i in seq_along(drugs)) {
  cat(sprintf("%-12s", drugs[i]))
  for (j in seq_along(drugs)) {
    severity <- interaction_matrix[i, j] + 1  # +1 for 1-based indexing
    label <- severity_labels[severity]
    cat(sprintf("%-12s", label))
  }
  cat("\n")
}

# Find severe interactions
cat("\nSevere Drug Interactions:\n")
cat(strrep("-", 30), "\n")
for (i in 1:length(drugs)) {
  for (j in (i+1):length(drugs)) {  # Avoid duplicates
    if (j <= length(drugs) && interaction_matrix[i, j] == 3) {
      cat(sprintf("⚠️  %s + %s:  SEVERE\n", drugs[i], drugs[j]))
    }
  }
}
```

### 4. Loop Control Statements

#### Break Statement
Exits the loop prematurely when a condition is met.

```{r}
# Finding the first patient with severe adverse event
patients_data <- data.frame(
  id = c("PT001", "PT002", "PT003", "PT004", "PT005"),
  severity = c("Mild", "Moderate", "Severe", "Mild", "Severe"),
  stringsAsFactors = FALSE
)

cat("Searching for first severe adverse event:\n")
for (i in 1:nrow(patients_data)) {
  patient <- patients_data[i, ]
  cat(sprintf("Checking patient %s:  %s\n", patient$id, patient$severity))
  
  if (patient$severity == "Severe") {
    cat(sprintf("⚠️  ALERT: First severe AE found in patient %s\n", patient$id))
    break  # Exit loop immediately
  }
}
```

#### Next Statement (R equivalent of continue)
Skips the current iteration and moves to the next one.

```{r}
# Processing only patients with complete data
patients_lab_values <- data.frame(
  id = c("PT001", "PT002", "PT003", "PT004", "PT005"),
  creatinine = c(1.2, NA, 0.9, 1.1, 1.3),
  alt = c(25, 30, NA, 28, 35),
  stringsAsFactors = FALSE
)

cat("Processing patients with complete lab data:\n")
cat(strrep("-", 45), "\n")

for (i in 1:nrow(patients_lab_values)) {
  patient <- patients_lab_values[i, ]
  
  # Skip patients with missing data
  if (is.na(patient$creatinine) || is.na(patient$alt)) {
    cat(sprintf("%s:  Skipping - incomplete data\n", patient$id))
    next  # Skip to next iteration
  }
  
  # Process patient with complete data
  creat <- patient$creatinine
  alt <- patient$alt
  
  # Calculate estimated GFR (simplified formula)
  egfr <- 175 * (creat^(-1.154))  # Simplified calculation
  
  cat(sprintf("%s:  Creatinine=%.1f, ALT=%.0f, eGFR=%.1f\n", 
             patient$id, creat, alt, egfr))
  
  # Check for abnormal values
  if (creat > 1.2) {
    cat("  ⚠️  Elevated creatinine\n")
  }
  if (alt > 40) {
    cat("  ⚠️  Elevated ALT\n")
  }
}
```

### 5. Apply Functions (R Alternative to Loops)

R's apply family functions often provide more efficient alternatives to loops.

#### Example 1: BMI Calculations using lapply
```{r}
# Patient data as list
patients_list <- list(
  list(weight = 70, height = 1.75),
  list(weight = 65, height = 1.68),
  list(weight = 80, height = 1.82),
  list(weight = 55, height = 1.60),
  list(weight = 90, height = 1.78)
)

# Function to calculate BMI
calculate_bmi <- function(patient) {
  bmi <- patient$weight / (patient$height^2)
  category <- if (bmi < 18.5) "Underweight" 
              else if (bmi < 25) "Normal"
              else if (bmi < 30) "Overweight" 
              else "Obese"
  
  return(list(bmi = bmi, category = category))
}

# Calculate BMI for all patients using lapply
bmi_results <- lapply(patients_list, calculate_bmi)

cat("Patient BMI Analysis:\n")
cat(strrep("-", 25), "\n")

for (i in seq_along(patients_list)) {
  patient <- patients_list[[i]]
  result <- bmi_results[[i]]
  
  cat(sprintf("Patient %d: %. 0fkg, %. 2fm → BMI: %.1f (%s)\n", 
             i, patient$weight, patient$height, 
             result$bmi, result$category))
}
```

#### Example 2: Dose Adjustments using sapply
```{r}
# Base doses for different age groups
patient_ages <- c(25, 45, 65, 75, 85)
base_dose <- 100  # mg

# Function for age-adjusted dosing
adjust_dose <- function(age, base_dose = 100) {
  if (age > 65) {
    # Reduce by 1% for every year over 65
    adjustment_factor <- 1 - (age - 65) * 0.01
    return(base_dose * adjustment_factor)
  } else {
    return(base_dose)
  }
}

# Calculate adjusted doses using sapply
adjusted_doses <- sapply(patient_ages, adjust_dose, base_dose = base_dose)

cat("Age-Adjusted Dosing:\n")
cat(strrep("-", 25), "\n")

for (i in seq_along(patient_ages)) {
  age <- patient_ages[i]
  base <- base_dose
  adjusted <- adjusted_doses[i]
  adjustment <- ((adjusted - base) / base) * 100
  
  cat(sprintf("Age %d: %dmg → %. 1fmg (%+.1f%%)\n", 
             age, base, adjusted, adjustment))
}
```

## Advanced Loop Patterns in  R Programming

### 1. Data Validation with Comprehensive Checks
```{r}
validate_patient_data <- function(patients_df) {
  required_fields <- c('id', 'age', 'weight', 'height')
  errors <- character(0)
  valid_rows <- logical(nrow(patients_df))
  
  for (i in 1:nrow(patients_df)) {
    patient <- patients_df[i, ]
    patient_errors <- character(0)
    
    # Check required fields
    for (field in required_fields) {
      if (is.na(patient[[field]]) || is.null(patient[[field]])) {
        patient_errors <- c(patient_errors, paste("Missing", field))
      }
    }
    
    # Validate data ranges
    if (! is.na(patient$age)) {
      if (patient$age < 18 || patient$age > 100) {
        patient_errors <- c(patient_errors, 
                           paste("Invalid age:", patient$age))
      }
    }
    
    if (!is.na(patient$weight)) {
      if (patient$weight < 30 || patient$weight > 300) {
        patient_errors <- c(patient_errors, 
                           paste("Invalid weight:", patient$weight, "kg"))
      }
    }
    
    if (!is.na(patient$height)) {
      if (patient$height < 1.0 || patient$height > 2.5) {
        patient_errors <- c(patient_errors, 
                           paste("Invalid height:", patient$height, "m"))
      }
    }
    
    if (length(patient_errors) > 0) {
      error_msg <- sprintf("Patient %d (%s): %s", 
                          i, 
                          ifelse(is.na(patient$id), "Unknown", patient$id),
                          paste(patient_errors, collapse = ", "))
      errors <- c(errors, error_msg)
      valid_rows[i] <- FALSE
    } else {
      valid_rows[i] <- TRUE
    }
  }
  
  return(list(
    valid_patients = patients_df[valid_rows, , drop = FALSE],
    errors = errors
  ))
}

# Example usage
test_patients <- data.frame(
  id = c('PT001', 'PT002', 'PT003', 'PT004'),
  age = c(45, 150, NA, 35),  # PT002 has invalid age, PT003 missing age
  weight = c(70, 65, 80, 55),
  height = c(1.75, 1.68, 1.82, NA),  # PT004 missing height
  stringsAsFactors = FALSE
)

validation_result <- validate_patient_data(test_patients)

cat("Data Validation Results:\n")
cat(strrep("=", 30), "\n")
cat(sprintf("Valid patients: %d\n", nrow(validation_result$valid_patients)))
cat(sprintf("Invalid patients: %d\n", length(validation_result$errors)))

if (length(validation_result$errors) > 0) {
  cat("\nValidation Errors:\n")
  for (error in validation_result$errors) {
    cat(sprintf("❌ %s\n", error))
  }
}

if (nrow(validation_result$valid_patients) > 0) {
  cat("\nValid Patients:\n")
  print(validation_result$valid_patients)
}
```

### 2. Pharmacokinetic Simulation with Loops
```{r}
simulate_drug_concentration <- function(dose, ka, ke, vd, time_points) {
  # Simulate drug concentration over time using one-compartment model
  # dose: dose in mg
  # ka: absorption rate constant (1/h)
  # ke: elimination rate constant (1/h)
  # vd: volume of distribution (L)
  
  concentrations <- numeric(length(time_points))
  
  for (i in seq_along(time_points)) {
    t <- time_points[i]
    
    if (ka != ke) {
      # One-compartment model with first-order absorption
      conc <- (dose/vd) * (ka/(ka-ke)) * (exp(-ke*t) - exp(-ka*t))
    } else {
      # Special case when ka = ke
      conc <- (dose/vd) * ka * t * exp(-ke*t)
    }
    
    concentrations[i] <- max(0, conc)  # Concentration can't be negative
  }
  
  return(concentrations)
}

# Simulation parameters
dose <- 500  # mg
ka <- 1.5    # absorption rate (1/h)
ke <- 0.2    # elimination rate (1/h)
vd <- 50     # volume of distribution (L)

time_points <- seq(0, 24, by = 0.5)  # 0 to 24 hours, every 0.5h

concentrations <- simulate_drug_concentration(dose, ka, ke, vd, time_points)

cat("Pharmacokinetic Simulation:\n")
cat(strrep("=", 35), "\n")
cat(sprintf("Dose: %dmg\n", dose))
cat(sprintf("Absorption rate constant: %. 1f/h\n", ka))
cat(sprintf("Elimination rate constant: %.1f/h\n", ke))
cat(sprintf("Volume of distribution: %dL\n", vd))
cat("\n")

# Find key pharmacokinetic parameters
max_conc <- max(concentrations)
max_time <- time_points[which.max(concentrations)]
half_life <- log(2) / ke

cat(sprintf("Maximum concentration (Cmax): %.2f mg/L\n", max_conc))
cat(sprintf("Time to maximum (Tmax): %.1f hours\n", max_time))
cat(sprintf("Elimination half-life:  %.2f hours\n", half_life))

# Print concentration at key time points
key_times <- c(0, 0.5, 1, 2, 4, 8, 12, 24)
cat("\nConcentration at key time points:\n")

for (t in key_times) {
  # Find closest time point
  closest_idx <- which.min(abs(time_points - t))
  actual_time <- time_points[closest_idx]
  conc <- concentrations[closest_idx]
  cat(sprintf("T = %4.1fh: %6.2f mg/L\n", actual_time, conc))
}
```

### 3. Clinical Trial Enrollment Simulation
```{r}
# Set seed for reproducibility
set.seed(456)

simulate_trial_enrollment <- function(target_enrollment, 
                                    sites, 
                                    max_weeks = 52) {
  # Initialize tracking variables
  total_enrolled <- 0
  week <- 0
  enrollment_log <- data.frame(
    week = integer(0),
    site = character(0),
    patients_enrolled = integer(0),
    cumulative_enrolled = integer(0),
    stringsAsFactors = FALSE
  )
  
  cat("Clinical Trial Enrollment Simulation\n")
  cat(strrep("=", 40), "\n")
  cat(sprintf("Target enrollment: %d patients\n", target_enrollment))
  cat(sprintf("Number of sites: %d\n", length(sites)))
  cat("\n")
  
  while (total_enrolled < target_enrollment && week < max_weeks) {
    week <- week + 1
    week_enrolled <- 0
    
    # Each site attempts to enroll patients
    for (site in names(sites)) {
      # Random enrollment based on site capacity
      site_capacity <- sites[[site]]
      enrolled_this_week <- rpois(1, lambda = site_capacity)
      
      if (enrolled_this_week > 0) {
        total_enrolled <- total_enrolled + enrolled_this_week
        week_enrolled <- week_enrolled + enrolled_this_week
        
        # Log the enrollment
        new_row <- data.frame(
          week = week,
          site = site,
          patients_enrolled = enrolled_this_week,
          cumulative_enrolled = total_enrolled,
          stringsAsFactors = FALSE
        )
        enrollment_log <- rbind(enrollment_log, new_row)
        
        # Stop if target reached
        if (total_enrolled >= target_enrollment) {
          break
        }
      }
    }
    
    # Report progress every 4 weeks
    if (week %% 4 == 0) {
      cat(sprintf("Week %d: %d patients enrolled (Total: %d/%.0f%%)\n", 
                 week, week_enrolled, total_enrolled,
                 (total_enrolled/target_enrollment)*100))
    }
  }
  
  cat("\n")
  if (total_enrolled >= target_enrollment) {
    cat(sprintf("✅ Target enrollment reached in week %d!\n", week))
  } else {
    cat(sprintf("⚠️  Enrollment incomplete after %d weeks (%d/%d patients)\n", 
               max_weeks, total_enrolled, target_enrollment))
  }
  
  return(enrollment_log)
}

# Define sites with their enrollment capacity (patients per week on average)
trial_sites <- list(
  "Site_A" = 2.5,  # Academic medical center
  "Site_B" = 1.8,  # Community hospital
  "Site_C" = 3.1,  # Large clinical research center
  "Site_D" = 1.2,  # Small clinic
  "Site_E" = 2.0   # Regional hospital
)

# Run simulation
enrollment_log <- simulate_trial_enrollment(
  target_enrollment = 100,
  sites = trial_sites,
  max_weeks = 30
)

# Summary by site
if (nrow(enrollment_log) > 0) {
  cat("\nEnrollment Summary by Site:\n")
  cat(strrep("-", 30), "\n")
  
  for (site in unique(enrollment_log$site)) {
    site_data <- enrollment_log[enrollment_log$site == site, ]
    total_by_site <- sum(site_data$patients_enrolled)
    weeks_active <- nrow(site_data)
    
    cat(sprintf("%s: %d patients (%d weeks active)\n", 
               site, total_by_site, weeks_active))
  }
}
```

## Best Practices for Loops in Pharmaceutical R Programming

### 1. Vectorization vs Loops
```{r}
# Example:  Calculating dose adjustments

# Sample data
patient_weights <- c(60, 75, 80, 65, 90, 55, 70, 85)
base_dose_mg_per_kg <- 5

cat("Comparison:  Loop vs Vectorized Operations\n")
cat(strrep("=", 45), "\n")

# Method 1: Using a loop (less efficient)
cat("Method 1: Using for loop\n")
doses_loop <- numeric(length(patient_weights))

start_time <- Sys.time()
for (i in seq_along(patient_weights)) {
  doses_loop[i] <- patient_weights[i] * base_dose_mg_per_kg
}
loop_time <- Sys.time() - start_time

cat(sprintf("Loop result: %s\n", paste(doses_loop, collapse = ", ")))
cat(sprintf("Time taken: %.6f seconds\n\n", as.numeric(loop_time)))

# Method 2: Using vectorization (more efficient)
cat("Method 2: Using vectorization\n")
start_time <- Sys.time()
doses_vectorized <- patient_weights * base_dose_mg_per_kg
vector_time <- Sys.time() - start_time

cat(sprintf("Vectorized result: %s\n", paste(doses_vectorized, collapse = ", ")))
cat(sprintf("Time taken: %.6f seconds\n", as.numeric(vector_time)))

# Verify results are identical
cat(sprintf("Results identical: %s\n", identical(doses_loop, doses_vectorized)))
```

### 2. Error Handling in Loops
```{r}
process_lab_results_safe <- function(results) {
  # Process lab results with proper error handling
  
  if (length(results) == 0) {
    cat("⚠️  No lab results to process\n")
    return(invisible(NULL))
  }
  
  processed_count <- 0
  errors_count <- 0
  
  for (i in seq_along(results)) {
    # Use tryCatch for error handling
    result <- tryCatch({
      current_result <- results[[i]]
      
      # Validate structure
      if (!is.list(current_result)) {
        stop("Result must be a list")
      }
      
      if (!all(c('value', 'reference_min', 'reference_max') %in% names(current_result))) {
        stop("Missing required fields")
      }
      
      # Extract values
      value <- current_result$value
      ref_min <- current_result$reference_min
      ref_max <- current_result$reference_max
      
      # Validate values are numeric
      if (!is.numeric(value) || !is.numeric(ref_min) || !is.numeric(ref_max)) {
        stop("Values must be numeric")
      }
      
      # Determine status
      status <- if (value < ref_min) "Low" 
               else if (value > ref_max) "High" 
               else "Normal"
      
      cat(sprintf("Result %d: %. 1f (Ref: %.1f-%.1f) - %s\n", 
                 i, value, ref_min, ref_max, status))
      
      processed_count <<- processed_count + 1
      "SUCCESS"
      
    }, error = function(e) {
      cat(sprintf("❌ Error processing result %d:  %s\n", i, e$message))
      errors_count <<- errors_count + 1
      "ERROR"
    })
  }
  
  cat(sprintf("\nProcessing Summary:\n"))
  cat(sprintf("Successfully processed:  %d\n", processed_count))
  cat(sprintf("Errors encountered: %d\n", errors_count))
}

# Example usage with mixed valid/invalid data
lab_data <- list(
  list(value = 4.5, reference_min = 3.5, reference_max = 5.5),
  list(value = 2.1, reference_min = 3.5, reference_max = 5.5),  # Low
  list(value = 6.8, reference_min = 3.5, reference_max = 5.5),  # High
  list(value = "invalid"),  # Invalid format
  list(value = 4.2, reference_min = 3.5, reference_max = 5.5),
  list(value = 3.8)  # Missing reference values
)

process_lab_results_safe(lab_data)
```

### 3. Progress Tracking for Long Operations
```{r}
analyze_large_dataset <- function(dataset, batch_size = 1000) {
  # Analyze large dataset with progress indication
  
  total_records <- length(dataset)
  processed <- 0
  
  cat(sprintf("Processing %d records in batches of %d.. .\n", 
             total_records, batch_size))
  cat(strrep("-", 50), "\n")
  
  # Process in batches
  for (start_idx in seq(1, total_records, batch_size)) {
    end_idx <- min(start_idx + batch_size - 1, total_records)
    batch <- dataset[start_idx:end_idx]
    
    # Simulate processing each record in batch
    for (i in seq_along(batch)) {
      # Simulate some processing time
      Sys.sleep(0.001)  # Very short delay for demonstration
      processed <- processed + 1
    }
    
    # Show progress
    progress <- (processed / total_records) * 100
    cat(sprintf("Progress: %d/%d (%.1f%%) - Batch %d complete\n", 
               processed, total_records, progress,
               ceiling(start_idx / batch_size)))
  }
  
  cat("✅ Analysis complete!\n")
  
  return(processed)
}

# Example with simulated dataset (smaller for demonstration)
set.seed(789)
sample_dataset <- replicate(5000, list(
  id = paste0("PT", sprintf("%04d", sample(1:9999, 1))),
  value = rnorm(1, 100, 15)
), simplify = FALSE)

# Process the dataset
# processed_count <- analyze_large_dataset(sample_dataset, batch_size = 1000)
```

## Summary

Loops are essential tools in pharmaceutical R programming for:

- **Data Processing**: Iterating through patient records, lab results, and clinical data
- **Calculations**: Computing dosages, pharmacokinetic parameters, and statistical measures  
- **Quality Control**: Validating data integrity and checking compliance
- **Simulations**: Modeling drug behavior and treatment outcomes
- **Analysis**: Processing large datasets and generating reports

### Key Takeaways for R: 

1. **Choose the right approach**: Consider vectorization before loops
2. **Use apply functions** when appropriate for better performance
3. **Always validate data** before processing  
4. **Include error handling** with `tryCatch()` for robust code
5. **Add progress indicators** for long-running operations
6. **Use meaningful variable names** for clarity
7. **Consider data. frame operations** instead of loops when possible

### R-Specific Best Practices:

- **Vectorization** is usually faster than loops in R
- **Pre-allocate vectors/lists** to improve performance
- **Use `lapply()`, `sapply()`, `mapply()`** for functional programming
- **Avoid growing objects** inside loops (use indexing instead)
- **Consider `data.table`** for very large datasets

Remember that efficient loop design and choosing the right approach (loops vs vectorization vs apply functions) can significantly impact the performance and reliability of pharmaceutical data analysis systems in R. 